\documentclass[handout]{mcs}

\begin{document}

\paper{team-survey2}{}{Team Survey II}

This is an optional survey.  Your TA will present a summary of answers
to your team for discussion, but your survey form and specific answers
will not be revealed to other students, and your name will not be
recorded.

\begin{itemize}

\item 
Circle your session \textbf{Time}: \hspace{0.4in} 9:30 \hspace{0.4in}
11 \hspace{0.4in} 2:30 \hspace{0.5in} Team/Table \textbf{number}\examrule[0.5in]

\iffalse

\item Would you like the staff (Meyer/TA/LA) to present solutions to
  selected problems in class after your team has finished working on
  them?
\begin{center}
YES \hspace{1in} NO
\end{center}
\fi

\item How helpful to you so far has the class teamwork been in
  mastering the class material?

\begin{center}
very\hspace{0.5in} somewhat \hspace{0.5in} neutral\hspace{0.5in} not
very \hspace{0.5in} waste of time
\end{center}

\item For you personally, the speed at which your team works through
  the class problems is:

\begin{center}
very slow\hspace{0.4in} somewhat slow\hspace{0.4in} about
right\hspace{0.4in} somewhat fast\hspace{0.4in} very fast
\end{center}

\item How helpful to you has your TA/LA been during team sessions in class?

\begin{center}
very\hspace{0.5in} somewhat \hspace{0.5in} not very \hspace{0.5in} not at all
\end{center}

\item How helpful to you has the instructor been during team sessions in class?

\begin{center}
very\hspace{0.5in} somewhat \hspace{0.5in} not very \hspace{0.5in} not at all
\end{center}


\item Name one person on your team who in the last few weeks has
  \begin{itemize}
    \item become more helpful to you personally \hfill\examrule[2in]
    \item seemed to be better prepared for class\hfill\examrule[2in]
    \item given better explanations to others\hfill\examrule[2in]
    \item improved at board writing\hfill\examrule[2in]
    \item improved at catching mistakes in board solutions\hfill\examrule[2in]
  \end{itemize}


\item How helpful to you is the online tutor?

\begin{center}
very\hspace{0.5in} somewhat \hspace{0.5in} neutral\hspace{0.5in} not
very \hspace{0.5in} waste of time
\end{center}

\item How helpful to you is the piazza forum?

\begin{center}
very\hspace{0.5in} somewhat \hspace{0.5in} neutral\hspace{0.5in} not
very \hspace{0.5in} waste of time
\end{center}


\item How well have you been able to understand the posted problem
  solutions (possibly with help)?
 
\begin{center}
very well\hspace{0.4in} mostly\hspace{0.4in} adequately\hspace{0.4in}
partially\hspace{0.4in} poorly
\end{center}

\item If the option to take a 30 minute quiz ---which would
  be graded as much on quality of exposition as correctness ---instead of
  attending a week's classes were available, for how many more weeks do you think
  you would do this?  (There are 6.5 weeks left in the term.)\hfill\examrule[0.7in]

\item How many times have you been to TA office hours?\hfill\examrule[0.7in]

\item How well do you feel your miniquiz performance reflects your
  understanding of the class material?

\begin{center}
very well\hspace{0.4in} adequately\hspace{0.4in} partially\hspace{0.4in} poorly
\end{center}

\end{itemize}

%\newpage %\examspace

\begin{itemize}

\item Should the instructor's presentations, be longer, \dots shorter,
  \dots use the white board instead of slides, \dots be more
  interactive, \dots be changed in some other way?

\vspace{1in}%\examspace[1in]

\item Feel free to suggest ways to improve the way the class works:

%\examspace[2in]

\end{itemize}


\end{document}
