\documentclass[handout]{mcs}

\begin{document}

\paper{team-survey1}{}{Team Survey}

This is an optional survey.  Your TA will present a summary of answers
to your team for discussion, but your survey form and specific answers
will not be revealed to other students, and your name will not be
recorded.

\begin{itemize}

\item 
Circle your session Time: \hspace{0.4in} 9:30 \hspace{0.4in} 11 \hspace{0.4in} 2:30

Team/Table number\brule{0.5in}

\item For you personally, the speed at which your team works through
  the class problems is (circle one):

very slow\hspace{0.4in} somewhat slow\hspace{0.4in} about
right\hspace{0.4in} somewhat fast\hspace{0.4in} very fast

\item Would you like the staff (Meyer/TA/LA) to present solutions to
  selected problems in class after your team has finished working on
  them?
\begin{center}
YES \hspace{1in} NO
\end{center}

\item How helpful to you so far has the class teamwork been in
  mastering the class material?

\begin{center}
very\hspace{0.5in} somewhat \hspace{0.5in} neutral\hspace{0.5in} not
very \hspace{0.5in} waste of time
\end{center}

\item Who on your team has
  \begin{itemize}
    \item been most helpful to you personally?\hfill\brule{2in}
    \item seemed to have best prepared for class?\hfill\brule{2in}
    \item given the best explanations?\hfill\brule{2in}
    \item done the best board writing?\hfill\brule{2in}
    \item been best at catching mistakes in board solutions?\hfill\brule{2in}
  \end{itemize}

\item How comfortable are you in the present class learning environment?

\begin{center}
Very \hspace{0.5in} somewhat comfortable\hspace{0.5in}
Somewhat \textbf{UN}comfortable \hspace{0.5in} Very \textbf{UN}
\end{center}


\item Feel free to suggest ways to improve the way the class works:
\vspace{1in}




\end{itemize}

\end{document}
