\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{Converted from sampling-conceptual.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 05:25:50 PM EDT}
\end{pcomments}

\begin{problem}

%% type: multi-part
%% title: Random Sampling

You work for the president and you want to estimate the fraction~$p$
of voters in the entire nation that will prefer him in the upcoming
elections.  You do this by random sampling.  Specifically, you select
$n$ voters independently and randomly, ask them who they are going to
vote for, and use the fraction $P$ of those that say they will vote
for the President as an estimate for $p$.

\bparts

\ppart
%% type: short-answer
%% title: Facts

Our theorems about sampling and distributions allow us to calculate
how confident we can be that the random variable $P$ takes a value
near the constant $p$.  This calculation uses some facts about voters
and the way they are chosen.  Which of the following facts are true?

\begin{enumerate}

\item
Given a particular voter, the probability of that voter preferring the
President is~$p$.

\item
Given a particular voter, the probability of that voter preferring the
President is 1 or~0.

\item
The probability that some voter is chosen more than once in the
sequence goes to zero as $n$ increases.

\item
All voters are equally likely to be selected as the third in our
sequence of $n$ choices of voters (assuming $n \ge 3$).

\item
The probability that the second voter chosen will favor the President,
given that the first voter chosen prefers the President, is greater
than~$p$.

\item
The probability that the second voter chosen will favor the President,
given that the second voter chosen is from the same state as the
first, may not equal~$p$.

\end{enumerate}

\begin{solution}

2,
4,
6

The preference of any particular voter is a constant: either ``the
President'' or ``not the President'', so (1) is false and (2) is
true. (3) is false; in fact, the Birthday ``paradox'' implies the
probability of some voter being chosen more than once rapidly
approaches 1 as $n$ grows beyond 100.  (4) holds by definition of
randomly choosing an item from a set.  (5) is false because successive
voters in the sequence are chosen independently.  (6) is true because,
for example, the fraction of voters who prefer the President in the
largest states may all be ${} < p$.
\end{solution}

\ppart
%% type: short-answer
%% title: What do you say?

Suppose that, according to your calculations the following is true
about your polling:

\begin{equation*}
    \pr{|P-p| \le 0.04} \ge 0.95
\end{equation*}

\ppart
You do the asking, you count how many said they will vote for the
President, you divide by $n$, and find 0.53.  You call the President,
and\dots what do you say?

\begin{enumerate}

\item Mr.\ President, $p=0.53$!

\item
Mr.\ President, with probability at least 95\%, $p$ is within 0.04 of
0.53.

\item
Mr.\ President, either $p$ is within 0.04 of 0.53 or something very
strange (5-in-100) has happened.

\item
Mr.\ President, we can be 95\% confident that $p$ is within 0.04 of 0.53.

\end{enumerate}

% Give the list of numbers that correspond to statements that you are
% justified to say.

\begin{solution}

3, 4

You cannot say (1): the only way to know the exact value of the
constant $p$ is to ask all 250,000,000 voters.

You cannot say (2) either: $p$ is a \emph{constant} which can either
be or not be within 0.04 of 0.53.  If it is, then the probability that
it is is~1, and thus at least 0.95, and therefore (2) will be true. If
it is not, then the probability that it is is~0, and thus smaller than
0.95, and therefore (2) will be false.

You can say (3): To see why, start with the statement
\begin{quote}
    either $|0.53 - p| \le  0.04$ or $|0.53 - p| > 0.04$
\end{quote}
which is obviously true.  Now read it as follows: \emph{Either} $p$ is
within 0.04 of 0.53 \emph{or} it is not and therefore my random
variable $P$ took a value from a set that is hit only 5 times in~100.
So, clearly, \emph{either} $p$ is within 0.04 of 0.53 \emph{or}
something strange has happened.

You can say (4): by rephrasing (2) as ``confidence'' rather than
probability, you are correctly indicating that you are talking about
the probable behavior of your methodology for sampling $p$, not the
actual value of~$p$
\end{solution}

\eparts

\end{problem}

\endinput
