\documentclass[12pt]{article}
\usepackage{light}

% add not exists = forall not

\hidesolutions
\showsolutions

\newcommand{\card}[1]{\left|#1\right|}

\newcommand{\C}{\textit{incabal}}

\begin{document}

\recitation{1}{September 5, 2014}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Team Problem:  Contrapositive}
Prove by truth table that an implication is equivalent to its contrapositive.
\solution[\vspace{0.3in}]{
\[
\begin{array}{ccccccc}
\text{x} & \text{y} & \text{x} \rightarrow \text{y} & \neg \text{y} & \neg \text{x} & \neg \text{y} \rightarrow \neg \text{x} & (\text{x} \rightarrow \text{y}) \leftrightarrow (\neg \text{y} \rightarrow \neg \text{x}) \\ \hline
T & T & T & F & F & T & T \\
T & F & F & T & F & F & T \\
F & T & T & F & T & T & T \\
F & F & T & T & T & T & T \\
   &    &    &    &    &    &   \\
   &    &    &    &    &    &   \\
   &    &    &    &    &    &
\end{array}
\]
In every row, $\text{x} \rightarrow \text{y}$ is T precisely when $\neg \text{y} \rightarrow \neg \text{x}$ is T.  Thus, we conclude that an implication is equivalent to its contrapositive.
}

\pagebreak

\section{Team Problem:  A Mystery}

A certain cabal within the 6.042 course staff is plotting to make the
final exam \textit{ridiculously hard}.  (``Problem 1.  Prove that the
axioms of mathematics are complete and consistent.  Express your
answer in Mayan hieroglyphics.'')  The only way to stop their evil plan
is to determine exactly who is in the cabal.  The course staff
consists of nine people:
%
\[
%\set{\text{Spyros}, \text{Ling}, \text{Chieu}, \text{Nick},
%\text{Bill}, \text{Jay}, \text{Brooke}, \text{Marten}, \text{Tom}}
\set{\text{Madalina}, \text{Catherine}, \text{Jeffrey}, \text{Brando}, \text{Nirvan}, \text{Ashley}, \text{Chennah}, \text{Ankur}, \text{Tom}}
\]
% Spyros -> Madalina
%Ling -> Chennah
% Chieu->Catherine
%Marten->Ankur
%Nick->Jeffrey
%Bill->Nirvan
%Brooke->Ashley
%Jay->Brando
%
The cabal is a subset of these nine.  A membership roster has been
found and appears below, but it is deviously encrypted in logic
notation.  The predicate $\C$ indicates who is in the cabal; that is,
$\C(x)$ is true if and only if $x$ is a member.  Translate each
statement below into English and deduce who is in the cabal.

\begin{enumerate}[\upshape (i)]

\item\label{eee} $\exists x \ \exists y \ \exists z \
    (x \neq y \wedge
     x \neq z \wedge
     y \neq z \wedge
     \C(x) \wedge \C(y) \wedge \C(z))$

\solution[\vspace{0.3in}]{A direct English paraphrase would be ``There
exist people we'll call $x,y$, and $z$, who are all different, such that
$x,y$ and $z$ are each in the cabal.''  A better version would use the
fact that there's no need in this case to give names to the people.
Namely, a better paraphrase is, ``There are 3 different people in the
cabal.''  Perhaps a simpler way to say this is, ``The cabal is of size at
least 3.''}

\item\label{nYW} $\neg (\C(\text{Chennah}) \wedge \C(\text{Nirvan}))$

\solution[\vspace{0.3in}]{Chennah and Nirvan are not both in the cabal.
Equivalently: at least one of Chennah and Nirvan is not in the cabal.}

\item\label{Fall} ($\C(\text{Ashley}) \vee \C(\text{Jeffrey})) \rightarrow \forall x \ \C(x)$

\solution[\vspace{0.3in}]{If either Ashley or Jeffrey is in the cabal, then everyone is.}

\item\label{YW} $\C(\text{Chennah}) \rightarrow \C(\text{Nirvan})$

\solution[\vspace{0.3in}]{If Chennah is in the cabal, then Nirvan is also.}

\item\label{CB} $\C(\text{Catherine}) \rightarrow \C(\text{Ashley})$

\solution[\vspace{0.3in}]{If Catherine is in the cabal, then Ashley is also.}

\item\label{SJnT}
$(\C(\text{Madalina}) \vee \C(\text{Brando})) \rightarrow \neg \C(\text{Tom})$

\solution[\vspace{0.3in}]{If either of Madalina or Brando is in the cabal,
then Tom is not.  Equivalently, if Tom \emph{is} in the cabal, then neither
Madalina nor Brando is.}

\item\label{SWnM}
$(\C(\text{Madalina}) \vee \C(\text{Nirvan})) \rightarrow \neg \C(\text{Ankur})$

\solution[\vspace{0.3in}]{If either of Madalina or Nirvan is in the cabal,
then Ankur is not.  Equivalently, if Ankur \emph{is} in the cabal, then
neither Madalina nor Nirvan is.  }
\end{enumerate}

\insolutions{So much for the translations.  We now argue that the only
cabal satisfying all seven propositions above is one whose members are
exactly Madalina, Nirvan, and Brando.

We first observe that by~\eqref{nYW}, there must be someone --- either
Chennah or Nirvan --- who is not in the cabal.  But if either Ashley
or Jeffrey were in the cabal, then by~\eqref{Fall}, everyone would be.
So we conclude by contradiction that

\begin{equation}\label{nF}
\text{Ashley and Jeffrey are not in the cabal.}
\end{equation}

Now consider that~\eqref{CB} implies its contrapositive: if Ashley is
not in the cabal, then neither is Catherine. Therefore, since Ashley is
not in the cabal,

\begin{equation}
\text{Catherine is not in the cabal.}
\end{equation}

Next observe that if Chennah were in the cabal, then by~\eqref{YW},
Nirvan would be too, contradicting~\eqref{nYW}.  So by again
contradiction, we conclude that
\begin{equation}\label{nC}
\text{Chennah is not in the cabal.}
\end{equation}

Now suppose Tom is in the cabal.  Then by~\eqref{SJnT}, Madalina and
Brando are not. We already know Ashley, Jeffrey, Catherine, and Chennah are not
in the cabal, leaving only three who could be --- Tom, Ankur, and
Nirvan.  But by~\eqref{eee} the cabal must have at least three
members, so it follows that the cabal must consist of exactly these
three.  This proves:
\begin{lemma}\label{TMW}
\text{If Tom is in the cabal, then Ankur and Nirvan are in the cabal.}
\end{lemma}

But by~\eqref{SWnM}, if Nirvan is the cabal, then Ankur is not.  That is, 
\begin{lemma}\label{WnM}
\text{Nirvan and Ankur cannot both be in the cabal.}
\end{lemma}
Now from Lemma~\ref{WnM} we conclude that the conclusion of
Lemma~\ref{TMW} is false.  So by contrapositive, the hypothesis of
Lemma~\ref{TMW} must also be false, namely,
\begin{equation}\label{nT}
\text{Tom is not in the cabal.}
\end{equation}

Finally, suppose Ankur is in the cabal.  Then by~\eqref{SWnM},
Madalina and Nirvan are not, and we already know Ashley, Jeffrey, Catherine,
Chennah, and Tom are not. So the cabal must consist of at most two
people (Ankur and Brando). This contradicts~\eqref{eee}, and we conclude
by contradiction that
\begin{equation}\label{nM}
\text{Ankur is not in the cabal.}
\end{equation}
So the only remaining people who could be in the cabal are Madalina,
Nirvan, and Brando.  Since the cabal must have at least three members,
we conclude that
\begin{lemma}
The only possible cabal consists of Madalina, Nirvan, and Brando.
\end{lemma}

But we're not done yet: we haven't shown that a cabal consisting of
Madalina, Nirvan, and Brando actually does satisfy all seven conditions.
So let $\mathcal{A} =\set{\text{Madalina}, \text{Nirvan},
  \text{Brando}}$, and let's quickly check that $\mathcal{A}$
satisfies~\eqref{eee}--\eqref{SWnM}:

\begin{itemize}

\item $\card{A} = 3$, so $A$ satisfies~\eqref{eee}.
\item Chennah is not in $A$, so $A$ satisfies~\eqref{nYW} and~\eqref{YW}.
\item Neither Ashley nor Jeffrey is in $A$, so the hypothesis
  of~\eqref{Fall} is false, which means that $A$
  satisfies~\eqref{Fall}.
\item Catherine is not in $A$, so $A$ satisfies~\eqref{CB}.
\item Finally, Tom and Ankur are not in $A$, so the conclusions of
both~\eqref{SJnT} and~\eqref{SWnM} are true, and so $A$
satisfies~\eqref{SJnT} and ~\eqref{SWnM}.

\end{itemize}

So now we have proved
\begin{proposition*}
$\set{\text{Madalina}, \text{Nirvan}, \text{Brando}}$ is the \emph{unique} cabal
satisfying conditions~\eqref{eee}--\eqref{SWnM}.
\end{proposition*}}

\end{document}
