\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{TP_Flipping_coins}
    \pcomment{Converted from prob2.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 05:11:14 PM EDT}
\end{pcomments}

\begin{problem}

%% type: short-answer
%% title: Flipping coins

Suppose you flip a fair coin 100 times.  The coin flips are all
mutually independent.

\bparts

\ppart
What is the expected number of heads?

\begin{solution}

50

Let $X$ denote the number of heads.  Let $X_{i}$ be the indicator
variable that is~1 when the $i$th coin flip is heads. Then
\begin{equation*}
    X = X_{1} + X_{2} + \dots + X_{100}.
\end{equation*}

Hence, by linearity of expectation, 
\begin{equation*}
\Ex[X]
    = \Ex[X_{1} + X_{2} + \dots + X_{100}]
    = \Ex[X_{1}] + \dots + \Ex[X_{100}]
\end{equation*}
and we need to find the expectation of each individual indicator. We
compute:
\begin{equation*}
\Ex[X_{i}]
    = 1 \cdot Pr{\text{flip head}} + 0 \cdot P{\text{flip tail}}
     = 1/2.  
\end{equation*}
Putting everything together, we conclude that $\Ex[X] = 100 \cdot 1/2
= 50$.
\end{solution}

\ppart
What is an upper bound on the probability that the number of heads is
at least~70 according to Markov's Theorem?

\begin{solution}

5/7

The expected number of heads is~50.  So the probability that the
number of heads is at least~70 is at most $50/70 = 5/7$.
\end{solution}

\ppart
What is the variance of the number of heads?

\begin{solution}

25

Let $X$ and $X_{i}$ be as in Question~1.  Then, by the independence of
the $X_{i}$, we know

\begin{equation*}
\Var[X] = 
\Var[X_{1} + \dots  + X_{100}] = 
\Var[X_{1}]+ \dots  + \Var[X_{100}]
\end{equation*}
and we need to find the variance of each indicator. We compute:
\begin{equation*}
\Var[X_{i}]
    = \Ex[X^{2}] - \Ex[X]^{2}
    = 1/2 - (1/2)^{2}
    = 1/4.
\end{equation*}
Putting everything together, we have $\Var[X] = 100 \cdot (1/4) = 25$.
\end{solution}

\ppart

What is an upper bound on probability that the number of heads is less
than~30 or greater than~70 according to Chebyshev's Theorem?

\begin{solution}

0.0625

The mean is 50.  So, the variable is less than~30 and more than~70 iff
it deviates from its mean by $20=|30-50|=|70-50|$ or more. How big is
this deviation in comparison to the \emph{standard deviation}?  Since
the variance of the number of heads is~25, its standard deviation is
5.  So, deviation by 20 is 4 times the standard one.

Now, Chebyshev's Theorem says that the probability of the number of
heads deviating from its mean by 4 times the standard deviation is
${}\le 1/(4^{2}) = 1/16$.
\end{solution}

\eparts

\end{problem}

\endinput
