\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{Converted from prob1.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 03:52:26 PM EDT}
\end{pcomments}

\begin{problem}

%% type: short-answer
%% title: Pigeonhole Principle

Below is a list of properties that a group of people might possess.
 
For each property, either give the minimum number of people that must
be in a group to ensure that the property holds, or else write
\textbf{nh} to indicate that the property need \textbf{n}ot
\textbf{h}old even for arbitrarily large groups of people.

(Assume that every year has exactly 365 days; ignore leap years.)

\bparts

\ppart 
At least 2 people were born on the same day of the year (ignore year
of birth).

\begin{solution}

366

We let the pigeons be people and the days of the year be the holes
(365 holes).  If we have $365+1$ pigeons, two must be in the same hole
(i.e. the two must be born on the same day).
\end{solution}

\ppart
At least 2 people were born on January 1. 

\begin{solution}
nhNHNh

No matter how many people you have, you cannot force one of them to be
born on a specific day.  For example, everyone might be born on
January~2.
\end{solution}

\ppart 
At least 3 people were born on the same day of the week.

\begin{solution}

15

We let people be pigeons and the days on the week be holes (7
holes). Using the generalized pigeonhole principle, we need $2*7+1$
people to force 3 of them to be in the same hole (i.e. born on the
same day of the week).
\end{solution}

\ppart
At least 4 people were born in the same month. 

\begin{solution}
37

We let the people be the pigeons and the months of the year be the
holes (12 holes).  Using the generalized pigeonhole principle, we need
$3*12+1$ people to force 4 of them to be born in the same month.
\end{solution}

\ppart
At least 2 people were born exactly one week apart. 

\begin{solution}
nhNHNh

Again, you cannot force this.  For example, everyone might be born on
the same day of the year.
\end{solution}

\eparts

\end{problem}

\endinput
