\documentclass[12pt,twoside]{article}
\usepackage{light}

\begin{problem}[10]
Let $G$ be a graph.  Prove that there is a path from every
vertex of odd degree to some other vertex of odd degree in $G$.

\solution{
\begin{proof}
Note that there is a path between two vertices iff there is a \emph{walk}
between them, because the shortest walk between them must be a path.  So
we need only show that from every vertex of odd degree there is a
\emph{walk} to some other odd-degree vertex.

Now let $v$ be a vertex of odd degree.  Consider a longest walk starting
at $v$ in which each edge occurs at most once.  Then every edge leaving
the final vertex on the walk must already be on the walk.  We will show
that this final vertex of the walk is another vertex of odd degree.

Since $v$ has odd degree, the walk contains at least one edge.
Furthermore, the final vertex of the walk cannot be $v$, since the
edges incident to $v$ would then consist of the first and last edge of
the walk plus two edges for each time that the walk crosses $v$--- an
even total.  Therefore, the edges incident to the final vertex of the
walk consist of the final edge of the walk plus two edges for each
time that the walk crosses the final vertex--- an odd total, as
claimed.
\end{proof}

\begin{proof}
An alternative proof begins with the observation from lecture that the sum
of the degrees of the vertices in any connected component is twice the
number of edges in the component.  So there must be an even number of
odd-degree vertices in any connected component.  In particular, if there
is a vertex of odd degree in the component, there must be an odd number of
additional odd-degree vertices connected to it.
\end{proof}
}
\end{problem}
