\documentclass{article}
\usepackage{../light}

\hidesolutions

\begin{document}

%%%% directed graph problem. (Oscar)
\begin{problem}{20}
Consider a strongly connected directed graph with $\mathrm{indegree}(v) = \mathrm{outdegree}(v)$ for all $v \in V$. We will prove such a graph has a (directed) Eulerean tour, by considering its longest path.

\ppart {10} Show the longest sequence of adjacent edges (walk or tour) where no edges are repeated is a tour.

\solution{Consider the longest walk or tour in the graph (whichever is longer). We want to show it is indeed a tour. By way of contradiction, suppose not. Then we have a walk where each edge is different. The starting node $v_s$ may have appeared in the walk on several occasions. We can count the number of edges used when occasions this happens. There is initial start edge.  From then on the rest of the edges are paired: one in for each one out of $v_s$. Hence we used up one more outgoing edge than incoming edges, so we can extend the path backwards by prefixing that missing incoming edge to $v_s$. This means our walk was not as long as possible.}

\ppart {10} Show no directed edge is left out of the longest possible walk or tour. 
\solution{
Suppose some edge $(u,w)$ is left out from the longest path. Since the graph is strongly connected, there is a shortest path from $u$  to some $v$ in the longest tour. We can then construct an even longer path: $u,w,\cdots,v,\cdots\v$ Where we prefix the path onto the existing cycle.
}

\end{problem}

\end{document}
