\documentclass[handout]{mcs}

\begin{document}

\paper{team-survey1}{}{Team Survey}

This is an optional, anonymous survey.  Your TA will present a summary
of answers to your team for discussion, but your survey form and
specific answers will not be revealed to other students.

We hope you will take this opportunity cite teammates \textbf{by name}
for contributions (solely \textbf{positive}) to the team activities.
This is an opportunity for you to express your appreciation to helpful
teammates and to provide useful feedback for improving teamwork.
These responses will not be recorded or graded.
\begin{center}
\textbf{Circle your}\quad   \teaminfo 
\end{center}

\begin{itemize}

\item For you personally, the speed at which your team works through
  the class problems is (circle one):

very slow\hspace{0.4in} somewhat slow\hspace{0.4in} about
right\hspace{0.4in} somewhat fast\hspace{0.4in} very fast

\item How helpful to you so far has the class teamwork been in
  mastering the class material?

very\hspace{0.5in} somewhat \hspace{0.5in} neutral\hspace{0.5in} not
very \hspace{0.5in} waste of time

\item How \textbf{comfortable} are you in the present class learning environment?

very comfortable \hspace{0.5in} somewhat\hspace{0.5in}
Somewhat \textbf{UN}comfortable \hspace{0.5in} Very \textbf{UN}

\item How helpful has the staff (LA/TA/Instructor) been in enabling
  your team to work together on the class problems?

very helpful \hspace{0.5in} moderately\hspace{0.5in}
somewhat \hspace{0.5in} not much \hspace{0.5in} not at all

\item Who on your team has
  \begin{itemize}
    \item been most helpful to you personally?\hfill\brule{2in}
    \item seemed to have best prepared for class?\hfill\brule{2in}
    \item given the best explanations?\hfill\brule{2in}
    \item done the best board writing?\hfill\brule{2in}
    \item been best improving board solutions?\hfill\brule{2in}
  \end{itemize}

\iffalse

\item Would you like the staff (Meyer/TA/LA) to present solutions to
  selected problems in class after your team has finished working on
  them?

\begin{center}
Never \hspace{0.7in} Rarely \hspace{0.7in} Sometimes \hspace{0.7in}
Regularly \hspace{0.7in} Always
\end{center}
\fi

\item Feel free to suggest ways to improve the way the class works
  (use back for more space):
\vspace{1in}

\end{itemize}

\end{document}
