\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_3_and_7_cent_stamps_by_induction}
  \pcomment{from: F09.mq3}
  \pcomment{analysis of postage below 12\mcents\ dropped by ARM, 2/23/12}
\end{pcomments}

\pkeywords{
  induction
  strong_induction
  postage
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
  Any amount of 12 or more cents postage can be made using only
  3\mcents\ and 7\mcents\ stamps.  Prove this \emph{by induction}
  using the induction hypothesis
  \[
  S(n) \eqdef n+12 \text{ cents postage can be made using
  only 3\mcents\ and 7\mcents\ stamps}.
  \]

\begin{staffnotes}
The proof we give below uses strong induction.  If need be, tell
students that strong induction counts as being a kind of ``proof by
induction.''

Maybe mention that essentially same proof works for ordinary induction using hypothesis
\[
  \[
  S(n) \eqdef \forall m \leq n. m+12 \text{ cents postage can be made using
  only 3\mcents\ and 7\mcents\ stamps}.
  \]
\]
\end{staffnotes}

\begin{solution}

  \begin{proof}
    \inductioncase{Base case} ($n=0$):  12\mcents\ postage can be made with four 3\mcents\ stamps.

    \inductioncase{Inductive step}:  We assume the strong hypothesis that
    $S(k)$ for $n \ge k \ge 0$.  Now we must prove $S(n+1)$.  The
    proof is by cases:

\textbf{case} $n=0$: $S(0+1)$ holds because 12+1=13 cents postage can
be made using one 7\mcents\ and two 3\mcents\ stamps.

\textbf{case} $n=1$: $S(1+1)$ holds because $(1+1)+12=
14$\mcents\ postage can be made using two 7\mcents\ stamps.

\textbf{case} $n\ge 2$: Since $n > n-2 \ge 0$, we know by strong
induction that $S(n-2)$ holds.  But including an extra 3\mcents\ stamp
in the collection of 3\mcents\ and 7\mcents\ stamps that made
$(n-2)+12$ cents gives a collection that makes $(n-2)+12+3 = (n+1)+12$
cents, which proves $S(n+1)$.

Since $S(n+1)$ holds in any case, the inductive step has been proved.

It follows by strong induction $S(n)$ holds for all $n \in \naturals$.
That is, every amount of postage of $12$ cents or more can be made
with 3\mcents\ and 7\mcents\ stamps.
\end {proof}

\end{solution}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
