\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_3_and_5_cent_stamps_by_induction.tex}
  \pcomment{from: F09.mq3}
  \pcomment{similar to CP_10_and_15_cent_stamps_by_WOP but uses induction}
\end{pcomments}

\pkeywords{
  induction
  strong_induction
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

  Find all possible (nonzero) amounts of postage that can be paid
  exactly using 3\textcent\ and 5\textcent\ stamps.  Use induction to
  prove that your answer is correct.

  \hint Let $S(n)$ mean that exactly $n$\textcent\ postage can be paid
  using only 3\textcent\ and 5\textcent\ stamps.  Prove that the
  following proposition is true as part of your solution.
  \[
    \forall n.\ ( n \geq 8 ) \QIMPLIES\ S(n).
  \]

% \ppart
%
%   Let $S(n)$ mean that exactly $n$ cents of postage can be paid using
%   only 5 and 3 cent stamps.
%
%   Find the smallest $k$ such that the following
%   proposition is true.
%   \[
%   \forall n.\ ( n \geq k ) \QIMPLIES\ S(n).
%   \]
%
% \vspace{1in}
  % Use induction to prove that your answer to the previous problem
  % part is correct.

\begin{solution}

  \begin{proof}
    The following proof is by strong induction on $n$.

    We can begin by observing that the following postage amounts can
    be made by 3 and 5 cents stamps:

    \begin{align*}
    3 & = 3 \\
    5 & = 5 \\
    8 & = 3 + 5 \\
    9 & = 3 + 3 + 3 \\
    10 & = 5 + 5
    \end{align*}

    The 3 consecutive postage values $8, 9, 10$ will be our base cases
    for the induction proof.

    \textbf{Base cases:} $S(8)$, $S(9)$ and $S(10)$ are shown to hold
    by explicit calculations.

    \textbf{Inductive step:} For all $n \geq 10$, we assume that
    $P(8),\dots, P(n)$ are true in order to prove that $P(n+1)$ is
    true.

    By the assumption that $P(n-2)$ is true, we know that the postage value
    $n - 2$ can be paid with 3 and 5 cent stamps. By adding one 3 cent
    stamp to that postage, we will be able to pay for a postage of
    $n - 2 + 3 = n + 1$ cents, showing that $P(n+1)$ is true. It follows
    by strong induction that $P(n)$ holds for all $n \geq 10$.

    We have therefore shown that all postage values $ \geq 8 $
    can be paid by 3 and 5 cent stamps.
  \end {proof}

  Therefore, we have shown that the postage amounts 3, 5, and any
  $k \geq 8$ can be paid by 3 and 5 cent stamps.

\begin{editingnotes}
  Note that we have actually seen a proof for this using WOP in the
  \href{http://courses.csail.mit.edu/6.042/fall09/slides2m.pdf}{Well Ordering Principle}
  lecture.
\end{editingnotes}

\end{solution}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
