\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_AMM_root_2_proof}
  \pcomment{from: S09.cp1r}
\end{pcomments}

\pkeywords{
  root_2
  irrational
  well_ordering
  WOP
  contradiction
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Here is a different proof that $\sqrt{2}$ is irrational, taken from the
American Mathematical Monthly, v.116, \#1, Jan. 2009, p.69:

\begin{proof}
  Suppose for the sake of contradiction that $\sqrt{2}$ is rational, and
  choose the least integer, $q>0$, such that $\paren{\sqrt{2}-1}q$ is a
  nonnegative integer.  Let $q' \eqdef \paren{\sqrt{2}-1}q$.  Clearly $0<
  q' < q$.  But an easy computation shows that $\paren{\sqrt{2}-1}q'$ is a
  nonnegative integer, contradicting the minimality of $q$.
\end{proof}

\bparts

\ppart This proof was written for an audience of college teachers, and
at this point it is a little more concise than desirable.  Write out a
more complete version which includes an explanation of each step.

\begin{solution}
The points that need justification are:

\begin{enumerate}
\item Why is there a positive integer, $q$, such that $\paren{\sqrt{2}-1}q$
  is a nonnegative integer?  \emph{Answer:} Since $\sqrt{2}$ is rational,
  so is $\sqrt{2}-1$.  So $\sqrt{2}-1$ can be expressed as an integer
  quotient with positive denominator; now just let $q$ be that
  denominator.

\item Why is there such a \emph{least} positive integer, $q$?
  \emph{Answer:} As long as there is one such positive integer, there has
  to be a least one.  This obvious fact is known as the
  \emph{Well Ordering Principle}.

\item Why is $0< q' < q$?  \emph{Answer:} We know that $1 < \sqrt{2} < 2$,
  so $0 < \sqrt{2}-1 < 1$.  Therefore, $0 < \paren{\sqrt{2}-1}r < r$ for any
  real number $r>0$.

\item Why is $(\sqrt{2}-1)q'$ a nonnegative integer?  \emph{Answer:} It's
  actually positive, because it is a product of positive numbers.  It's
  integer because
  \[
  \paren{\sqrt{2}-1}q' = \paren{\sqrt{2}-1}^2q = 2q - 2q\sqrt{2} + q =
  q-2\cdot\brac{\paren{\sqrt{2}-1}q}
  \]
  and the last term is of the form
  $\ang{\text{integer}-2 \cdot \brac{\text{integer}}}$.

\end{enumerate}
\end{solution}

\ppart Now that you have justified the steps in this proof, do you have a
preference for one of these proofs over the other?  Why?  Discuss these
questions with your teammates for a few minutes and summarize your team's
answers on your whiteboard.

\begin{solution}
Both proofs seem about equally easy to understand.  The previous
  problems shows that the first proof generalizes pretty directly from
  square roots to $k$th roots, which doesn't seems as clear for the this
  second proof.  On the other hand, the first proof requires appeal to
  Unique Prime Factorization, while the second just uses simple algebra.
\end{solution}

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
