\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_build_MSTs}
  \pcomment{ARM 10/27/11}
  \pcomment{F11.cp9m}
  \pcomment{Figure should be inserted (3 copies); then delete last part}
\end{pcomments}

\pkeywords{
spanning_tree
weighted_tree
minimum_weight
MST
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Let $G$ be a $4 \times 4$ grid with vertical and horizontal
edges between neighboring vertices.  Formally,
\[
\vertices{G} = [0,3]^2 \eqdef \set{(k,j) \suchthat 0 \leq k,j \leq 3}.
\]
Letting $h_{i,j}$ be the horizontal edge $\edge{(i,j)}{(i+1,j)}$ and
$v_{j,i}$ be the vertical edge $\edge{(j,i)}{(j,i+1)}$ for $i\in[0,2],
j \in [0,3]$.   The weights of these edges are
\begin{align*}
w(h_{i,j}) & \eqdef  \frac{4i+j}{100},\\
w(v_{j,i}) & \eqdef 1+\frac{i+4j}{100}.
\end{align*}

(A picture of $G$ would help; you might like to draw one.)

\bparts

\ppart Construct a minimum weight spanning tree (MST) for $G$ by
initially selecting the minimum weight edge, and then successively
selecting the minimum weight edge that does not create a cycle with
the previously selected edges.  Stop when the selected edges form a
spanning tree of $G$.  (This is Kruskal's MST algorithm.)
\inhandout{Explain how the ``gray edge'' Lemma~\bref{lem:edgeextends}
  justifies this algorithm.}

\begin{solution}

The edges are in the order that they are constructed by the given algorithm.

Answer: $h_{0,0}h_{0,1}h_{0,2}h_{0,3}h_{1,0}h_{1,1}h_{1,2}h_{1,3}h_{2,0}h_{2,1}h_{2,2}h_{2,3}v_{0,0}v_{0,1}v_{0,2}$

\inhandout{From the text: An edge does not create a cycle iff it connects
different components.  The edge chosen by Kruskal's algorithm will be
the minimum weight gray edge when the components it connects are
assigned different colors.}

\end{solution}

\ppart Grow an MST for $G$ starting with the tree consisting of the
single vertex $(1,2)$ and successively adding the minimum weight edge
with exactly one endpoint in the tree.  Stop when the tree spans $G$.
(This is Prim's MST algorithm.)  \inhandout{Explain how the ``gray edge''
Lemma~\bref{lem:edgeextends} justifies this algorithm.}

\begin{solution}

Answer:
$h_{0,2}h_{1,2}h_{2,2}v_{0,1}h_{0,1}h_{1,1}h_{2,1}v_{0,0}h_{0,0}h_{1,0}h_{2,0}v_{0,2}h_{0,3}h_{1,3}h_{2,3}$

\inhandout{From the text: This is the algorithm that comes from coloring the
growing tree white and all the vertices not in the tree black.  Then
the gray edges are the ones with exactly one endpoint in the tree.
}
\end{solution}


\ppart Grow an MST for $G$ by treating the vertices $(0,0), (0,3),
(2,3)$ as 1-vertex trees and then successively adding, for each tree
in parallel, the minimum weight edge among the edges with one endpoint
in the tree.  Continue as long as there is no edge between two trees,
then go back to applying the general gray edge method until the
parallel trees merge to form a spanning tree of $G$.  (This is 6.042's
parallel MST algorithm.)

\begin{solution}

Done in parallel:

T1@(0,0): $h_{0,0}h_{1,0}h_{2,0}v_{0,0}h_{0,1}h_{1,1}h_{2,1}$

T2@(0,3): $h_{0,3}h_{2,3}v_{0,2}h_{0,2}h_{1,2}h_{2,2}v_{0,1}$ (merges with T1)

T3@(2,3): $h_{1,3}$ (merges with T2)
\end{solution}

\ppart Verify that you got the same MST each time.

\begin{solution}
They are the same ---if no mistake was made.
Problem~\bref{PS_unique_MST} explains why there is a unique MST for
any finite connected weighted graph where no two edges have the same weight.
\end{solution}

\eparts

\inhandout{
\begin{staffnotes}
Propose a couple of figures that could be added to the text to help
make the proof of the ``gray edge'' Lemma~\bref{lem:edgeextends}
clearer.
\end{staffnotes}
}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
