\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_bipartite_matching_recitations}
  \pcomment{This is the 1st part of MQ_bipartite_matching_recitations}
  \pcomment{edited from MQ_bipartite_matching_recitations by ARM 3/15/11}
\end{pcomments}

\pkeywords{
  graphs
  bipartite
  matching
  Halls_theorem
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
  Overworked and over-caffeinated, the Teaching Assistant's (TA's)
  decide to oust the lecturer and teach their own recitations.  They
  will run a recitation session at 4 different times in the same room.
  There are exactly 20 chairs to which a student can be assigned in
  each recitation.  Each student has provided the TA's with a list of
  the recitation sessions her schedule allows and no student's
  schedule conflicts with all 4 sessions.  The TA's must assign each
  student to a chair during recitation at a time she can attend, if
  such an assignment is possible.

  Describe how to model this situation as a matching problem.  Be sure
  to specify what the vertices/edges should be and briefly describe
  how a matching would determine seat assignments for each student in
  a recitation that does not conflict with his schedule.  This is a
  \emph{modeling problem}---you need not determine whether a match is
  always possible.

\begin{staffnotes}
Of course no match is possible if there are more than 80 students, but
even with, say, 65 students, there isn't enough information to
determine whether a match is possible.  
\end{staffnotes}

 \begin{solution}
  There will be one vertex for each student, and 20 vertices for each
  recitation time slot (one for each chair). There is an edge between
  a student and all chair vertices for a particular recitation time
  slot if that time slot does not conflict with her schedule. A
  matching for the students assigns a student to a chair in a
  recitation that he can attend and assigns at most 20 students to
  any recitation.

  It is possible to assign the students to recitations iff a matching
  exists.
\end{solution}

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\endinput
