\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_congruence_no_prime_polynomial}
  \pcomment{by ARM 1/6/11}
\end{pcomments}

\pkeywords{
  primes
  polynomial
  nonconstant
  congruence
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{problem}
  An \term{integer polynomial} is a polynomial with integer coefficients.
  Let $q(x)$ be an integer polynomial.

\begin{problemparts}
\ppart\label{jknq>1} Prove that if $j \equiv k \pmod n$ then $q(j)
\equiv q(k) \pmod n$ for all $j, k, n \in \integers$ with $n>1$.

\hint If $q(x)$ is of degree $d>0$, then it equals $c + xp(x)$ for
some integer $c$ and integer polynomial $p(x)$ of degree $d-1$.

\begin{solution}
  The proof is by induction on $d$ with induction hypothesis
  that~\eqref{jknq>1} holds for all integer polynomials, $q$, of
  degree $d$.

  \textbf{Base case} ($d=0$): A polynomial $q$ of degree $0$ is a constant, so 
   $q(j) =q(k)$ for all $j,k$.

  \textbf{Inductive step}: We must prove that~\eqref{jknq>1} holds for
  any integer polynomial, $q(x)$, of degree $d+1$.

  So suppose $j \equiv k \pmod n$ where $n>1$.  Now by the hint, $q(x)
  = c + xp(x)$ for some integer polynomial, $p(x)$ of degree, $d$.
  Also,
\begin{equation}\label{}
p(j) \equiv p(k) \pmod n
\end{equation}
by induction hypothesis, so
\begin{equation}\label{jpjkpk}
jp(j) \equiv kp(k) \pmod n
\end{equation}
by multiplicativity of congruence
(Lemma~\bref{congruence_facts}.bref{congruence_multiplicativity}.
Finally,
\begin{align*}
  q(j) & = c + jp(j) \equiv c + kp(k) \pmod n
          & \text{(by~\eqref{jpjkpk} and congruence additivity
             Lemma~\bref{congruence_facts}.bref{congruence_additivity})}\\
       & = q(k).
 \end{align*}
\end{solution}

\ppart\label{qunbmd} Suppose that $q(j)$ grows unboundedly as $j$ goes to infinity.
Conclude that $q(\naturals)$ contains infinitely many composite
positive integers.  \hint Choose $j$ such that $q(j) = n > 1$ and consider 
$k$'s such that $j \equiv k \pmod n$.

\begin{solution}
Since $q$ grows unboundedly, there will be $j \in \naturals$ such that
$1 < n \eqdef q(j)$, and for any such $j$, there will be an infinite
sequence of nonnegative integers $j_1, j_2,j_3,\dots$ all $\equiv j
\pmod n$ such that $q(j_1), q(j_2), q(j_3), \dots$ is a strictly
increasing sequence of integers greater than $n$.  By All the numbers
in this sequence will be composite, since by part~\eqref{jknq>1}, they
all have $n$ as a divisor.
\end{solution}

Part~\eqref{qunbmd} shows if an integer polynomial has infinitely many
positive values on $\naturals$, then infinite number of these values
will not be prime.  This fact no longer holds true for multivariate
polynomials.  An amazing consequence of Matijesevich's solution to
Hilbert's Tenth Problem (INSERT REFERENCE) that multivariate
polynomials can be understood as \emph{general purpose} programs for
generating sets of integers.  In particular, there is an integer
polynomial $p(x_1,\dots,x_7)$ whose nonnegative values as
$x_1,\dots,x_7$ range over $\naturals$ are precisely the set of all
prime numbers!

\end{problemparts}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
