\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_CNF_from_DNF}
  \pcomment{by ARM 2/5/11, minor edits 9/11/17, solution restructured by zabel on 9/12/17 to separate the example from the proof}
\end{pcomments}

\pkeywords{
  proposition
  formula
  dnf
  cnf
  equivalence
  conjunt
  disjunct}

\begin{problem}
  Explain a simple way to obtain a \idx{conjunctive form} for a
  propositional formula directly from a \idx{disjunctive form} for its
  complement.

  \hint DeMorgan's Law does most of the job. Try working an illustrative example of your choice before describing the general case.

  \begin{solution}
    We'll follow the suggestion by starting with an example. As we know, ``proof by example'' is not a valid form of proof, and this worked example is not required for a formal write-up. We include it here for clarity.
    
    The idea is to negate the given disjunctive form and simplify as much as possible with DeMorgan's Law. For the example
    \[
      \bar{F} = (A \QAND B) \QOR (B \QAND \bar{C}),
    \]
    negating this formula and applying DeMorgan's Law yields
    \begin{align*}
      F &\iff \QNOT((A \QAND B) \QOR (B \QAND \bar{C})) \\
        &\iff \QNOT(A \QAND B) \QAND \QNOT(B \QAND \bar{C}).
    \end{align*}
    Now we can turn each of the negated products into a sum by
    applying DeMorgan's Law again:
    \begin{equation*}
      F \iff (\bar{A} \QOR \bar{B}) \QAND (\bar{B} \QOR \bar{\bar{C}}).
    \end{equation*}
    Then removing double-negations yields the desired conjunctive form:
    \[
      F \iff (\bar{A} \QOR \bar{B}) \QAND (\bar{B} \QOR C).
    \]

    Now for a fully general argument. Suppose $\bar{F}$ is given in disjunctive form as $$\bar{F} = C_1 \QOR C_2 \QOR \cdots \QOR C_n.$$ Formula $F$ is equivalent to $\bar{\bar{F}}$, which may be rewritten using DeMorgan's law for distributing \QNOT\ over \QOR:
    \begin{equation}\label{againDeMnotor}
      F \iff \bar{C_1} \QAND \bar{C_2} \QAND \cdots \QAND \bar{C_n}.
    \end{equation}
    Each clause $C$ from $\bar{F}$ (i.e., one of $C_1,\ldots,C_n$) has the form $C = (L_1 \QAND L_2 \QAND \cdots \QAND L_m)$, where each $L_j$ is either a variable or a negated variable. By DeMorgan's law again, $\bar{C}$ may be simplified into
    \begin{equation*}
      \bar{C} = (\bar{L_1} \QOR \bar{L_2} \QOR \cdots \QOR \bar{L_m}).
    \end{equation*}
    If any of these literals $L_j$ was already a negated variable, say $L_j = \bar{X}$, then $\bar{L_j}$ may be equivalently replaced with the original variable $X$, since $\bar{\bar{X}} \iff X$.

    Applying the transformation from the previous paragraph to each clause in~\eqref{againDeMnotor} results in the desired disjunctive form for $F$.
  \end{solution}

\end{problem}

\endinput

