\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_computable_reducibility}
  \pcomment{ARM 3/7/16}
\end{pcomments}

\pkeywords{
  halting_problem
  recognizable
  unrecognizable
}

%\asciibet}{\text{ASCII}}
%\asciistr}{\strings{\asciibet}}

%\newcommand{\strtyp}{\ensuremath{\textbf{string}}}

%\newcommand{\rcg}{\text{lang}} % \def\all0s{\textbf{all-0s}}
%\newcommand{\lsub}[1]{\ensuremath{L_{#1}}}
%\newcommand{\notself}{\text{\small{Not-self}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
\emph{String procedures} are one-argument procedures that apply to
strings over the \asciibet\ alphabet.  If application of procedure,
$P$, to string $s$ results in a computation that eventually halts, we
say that $P$ \term{recognizes} $s$.  We define $\rcg{P}$ to be the set
of strings or \emph{language} recognized by $P$:
\[
\rcg{P} \eqdef \set{s \in \asciistr \suchthat P\ \text{recognizes}\ s}.
\]
A language is \emph{unrecognizable} when it is not equal to $\rcg{P}$
for any procedure $P$.

A string procedure declaration is a text, $s \in \asciistr$, that
conforms to the grammatical rules for programs.  The declaration
defines a procedure $P_s$, which we can think of as the result of
compiling $s$ into an executable object.  If $s \in \asciistr$ is not
a grammatically well-formed procedure declaration, we arbitrarily
define $P_s$ to be the string procedure that fails to halt when
applied to any string.  Now every string defines a string procedure,
and every string procedure is $P_s$ for some $s \in \asciistr$.

An easy diagonal argument in \inbook{Section~\bref{halting_sec}}
\inhandout{the text} showed that
\[
\nohalt \eqdef \set{s \suchthat P_s\ \text{applied to $s$ does not halt}}
    = \set{s \suchthat s \notin \rcg{P_s}}
\]
is not recognizable.

It may seem pretty weird to apply a procedure to its own declaration.
Are there any less weird examples of unrecognizable set?  The answer
is ``many more.''  In this problem, we'll show three more:
\begin{align*}
\nohaltlam
         & \eqdef \set{s \suchthat P_s\
             \text{applied to $\emptystring$ does not halt}}
         = \set{s \suchthat \emptystring \notin \rcg{P_s}},\\
\finhalt & \eqdef \set{s \suchthat \rcg{P_s}\ \text{is finite}},\\
\allhalt & \eqdef \set{s \suchthat \rcg{P_s} = \asciistr}.
\end{align*}

Let's begin by showing how we could use a recognizer for
\nohaltlam\ to define a recognizer for \nohalt.  That is, we will
``reduce'' the weird problem of recognizing \nohalt\ to the more
understandable problem of recognizing \nohaltlam.  Since there is no
recognizer for \nohalt, it follows that there can't be one for
\nohaltlam\ either.

Here's how this reduction would work: suppose we want to recognize
when a given string $s$ is in \nohalt.  Revise $s$ to be the
declaration of a slightly modified procedure $P_{s'}$ which behaves as
follows:
\begin{quote}
 $P_{s'}$ applied to argument $t\in \asciistr$, ignores $t$, and
 simulates $P_s$ applied to $s$.
\end{quote}
So, if $P_s$ applied to $s$ halts, then $P_{s'}$ halts on every string
it is applied to, and if $P_s$ applied to $s$ does not halt, then
$P_{s'}$ does not halt on any string it is applied to.  That is,
\begin{align*}
s \in \nohalt
    & \QIMPLIES\ \rcg{P_{s'}} = \emptyset\\
    & \QIMPLIES \emptystring \notin \rcg{P_{s'}}\\
    & \QIMPLIES s' \in \nohaltlam,\\
s \notin \nohalt
    & \QIMPLIES\ \rcg{P_{s'}} = \asciistr\\
    & \QIMPLIES\ \emptystring \in \rcg{P_{s'}}\\
    & \QIMPLIES s' \notin \nohaltlam.
\end{align*}
In short,
\[
s \in \nohalt \QIFF\ s' \in \nohaltlam.
\]
So to recognize when $s \in \nohalt$ all you need to do is recognize
when $s' \in \nohaltlam$.

\bparts

\ppart Conclude that \finhalt\ is unrecognizable.

\hint Same $s'$.

\begin{solution}
For $s'$ as above, we know
\begin{align*}
s \in \nohalt
    & \QIMPLIES\ \rcg{P_{s'}} = \emptyset\\
    & \QIMPLIES s' \in \finhalt,\\
s \notin \nohalt
    & \QIMPLIES\ \rcg{P_{s'}} = \asciistr\\
    & \QIMPLIES s' \notin \finhalt.
\end{align*}
So to recognize when $s \in \nohalt$ all you need to do is recognize
when $s' \in \finhalt$.
\end{solution}

\eparts
Next, let's see how a reduction of \nohalt\ to \allhalt\ would work.
Suppose we want to recognize when a given string $s$ is in \nohalt.
Revise $s$ to be the declaration of a slightly modified procedure
$P_{s''}$ which behaves as follows:
\begin{quote}
  When $P_{s''}$ is applied to argument $t\in \asciistr$, it simulates
  $P_s$ applied to $s$ for up to $\lnth{t}$ ``steps'' (executions of
  individual machine instructions).  If $P_s$ applied to $s$ has
  \emph{not} halted in $\lnth{t}$ steps, then the application of
  $P_{s''}$ to $t$ halts.  If $P_s$ applied to $s$ \emph{has} halted
  within $\lnth{t}$ steps, then the application of $P_{s''}$ to $t$
  runs forever.
\end{quote}

\bparts
\ppart\label{allhaltunrec} Concude that \allhalt\ is unrecognizable.

\hint Explain why
\[
s \in \nohalt \QIFF s'' \in \allhalt.
\]

\begin{solution}
\begin{align*}
s \notin \nohalt
  & \QIMPLIES\ \rcg{P_{s''}} = \set{t \suchthat
                         \lnth{t} \leq \#\text{steps until $P_s$ halts on}\ s}\\
  & \QIMPLIES\ \rcg{P_{s''}}\ \text{is finite}\\
  & \QIMPLIES s'' \notin \allhalt,\\
s \in \nohalt 
   & \QIMPLIES\ \rcg{P_{s''}} = \asciistr\\
   & \QIFF\ s'' \in \allhalt.
\end{align*}
In short,
\[
s \in \nohalt \QIFF\  s'' \in \allhalt.
\]
So to recognize when $s \in \nohalt$ all you need to do is recognize
when $s'' \in \allhalt$.  But since there is no recognizer \nohalt,
there can't one for \allhalt.
\end{solution}

\ppart Explain why $\overline{\finhalt}$ is unrecognizable.

\hint Same $s''$.

\begin{solution}
We have from the solution to part~\eqref{allhaltunrec} that
\begin{align*}
s \notin \nohalt
  & \QIMPLIES\ \rcg{P_{s''}}
       = \set{t \suchthat
               \lnth{t} \leq \#\text{steps until $P_s$ halts on $s$}}\\
  & \QIMPLIES\ \rcg{P_{s''}}\ \text{is finite}\\
  & \QIMPLIES s'' \in \finhalt\\
  & \QIFF s'' \notin \overline{\finhalt},\\
s \in \nohalt
   & \QIMPLIES\ \rcg{P_{s''}} = \asciistr\\
   & \QIMPLIES\ s'' \in \overline{\finhalt}.
\end{align*}
In short,
\[
s \in \nohalt \QIFF\ s'' \in \overline{\finhalt},
\]
which implies that $\overline{\finhalt}$ is unrecognizable.
\end{solution}

\eparts

Note that it's easy to recognize when $P_s$ does halt on $s$: just
simulate the application of $P_s$ to $s$ until it halts.  This shows
that $\overline{\nohalt}$ \emph{is} recognizable.  We've just
concluded that $\finhalt$ is nastier: neither it nor its complement is
recognizable.
\end{problem}

\endinput


\iffalse
\begin{definition}
A procedure that takes string inputs and returns string outputs is
called a \emph{string I/O-procedure}.  The function $\phi: \asciistr
\to \asciistr$ where $\phi(s)$ is the output, if any, when the
procedure is applied to $s$, is called a \emph{partial computable
  function}.  An I/O-procedure that always returns an output---it
never runs forever or aborts no matter what string it is applied
to---defines a total function called a \emph{total computable
  function} $f:\asciistr \to \asciistr$.

If $L,M \subseteq \asciistr$ are languages, $L$ is defined to
\emph{reducible} to $M$,\footnote{There are many different notions of
  how one langauge can be reduced to another.  The particular one we 
  use is known in the literature as ``many-one'' reducibility.}
in symbols,
\[
L \preceq M
\]
iff there is a total computable procedure $f$ such that
\[
s \in L\ \QIFF\ f(s) \in M.
\]
for all $s \in \asciistr$.
\item $P_s \subseteq \set{ \emptystring}$          %type-error
\item $P_s \text{ recognizes } \emptystring}$ %OK
\item $\emptystring \in \rcg{P_s}}$           %OK
\item $P_s \text{halts when applied to the}\ \emptystring}$ %OK
\item $\mathcal{P}_0(\rcg{P_s})$              %OK
\item $\mathcal{P}_0(\set{\emptystring})$     %OK
\item $\mathcal{P}_0(\overline{\rcg{P_s}})}$  %probably OK
\item $\rcg{P_s} = \set{ \emptystring}}$      %NO
\item $\rcg{P_s} = \set{ \emptystring}}$      %NO
\end{enumerate}

oMore generally, for any language property $\mathcal{P}$, let
\[
L_{\mathcal{P}} \eqdef \set{s \in \asciistr \suchthat \mathcal{P}(\rcg{P_s})}
\]
Now analyzing string procedures to see if they halt on the emptystring
means recognizing the language $L_{\mathcal{P}_0}$.

Analyzing a string procedure to determine if it recognizes a language
with property.  This means we are asking whether the language
\[
L_{\mathcal{P}} \eqdef \set{s \in \asciistr \suchthat \mathcal{P}(\rcg{P_s})}
\]
is recognizable.
\fi

\iffalse

This means we are simply asking is the following
language recognizable:
\[
\set{s \in \asciistr \suchthat \emptystring \in \rcg{P_s}}.
\]



Now we are asking for a recognizer for the strings $s$ such that $P_s$
applied to $\emptystring$ halts.  This is the same as asking whether
$P_s$ recognizes $\emptystring$, which can further rephrased as
asking: is
\[
\set{s \in \asciistr \suchthat \mathcal{P_0}(\rcg{P_s})}
\]


Anyway, recognizing when a procedure applied to the
empty string halts means recognizing


That is, given the procedure declaration $s \asciistr$, does
$\rcg{P_s}$

 procedures  to find a procedure analyzer that recognizes
given any procedure declaration, whether the declared procedure halts
k  

That is, can we recognize


when a string procedure recognizes a language with property
$\mathcal{P}$.  This means we are asking whether the language
\[
L_{\mathcal{P}} \eqdef \set{s \in \asciistr \suchthat \mathcal{P}(\rcg{P_s})}
\]
is recognizable.

  We can rephrase
this question in the following simple way: is there a recognizer for
\[
\set{s \in \asciistr \suchthat P_s\ \text{does not halt when applied to the}\ \emptystring}\ ?
\]
A more concise way to phrase this would be: is the language
\[
s \in \asciistr \suchthat 

So we're asking
if we can find a recognizer for
\fi
