\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_divisible_by_24}
  \pcomment{first used SP12}
  \pcomment{from shri, 3/1/2012}
\end{pcomments}

\pkeywords{
  number_theory
  modular_arithmetic
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

Show that for all primes $p /geq 5$, $p^2-1$ is divisible by $24$:

\begin{solution}
A number is divisible by $24$ if it is divisble by $3$ and $8$.All primes in our range are odd (not $2$) and not divisible by $3$, meaning that $p$ is $1$ or $2$ mod $3$. 
If $p$ is $1$ mod $3$, $p^2-1$=$0$ mod $3$
if $p$ is $2$ mod $3$, $p^2-1$=$0$ mod $3$
so $p^2-1$ is divisible by $3$!

Since $p$ is odd, it can be expressed as $2x+1$ for an integer $x$. We plug it in to $p^2-1$ to get $4x^2+4x$. If $x$ is even, it is easy to see that the total is divisible by $8$. If $x$ is odd, then we use $x=2y+1$, $8y^2+16y+4+8y+4$=$8y^2+24y+8$ which is divisible by $8$.

Since $p^2-1$ is divisible by $8$ and $3$ and the two numbers are relatively prime; it is divisible by $24$, so we are done.
\end{solution}

\iffalse
\bparts

\ppart Ten pirates find a chest filled with gold and silver coins.
There are twice as many silver coins in the chest as there are gold.
They divide the gold coins in such a way that the difference in the
number of coins given to any two pirates is not divisible by $10$.
They will only take the silver coins if it is possible to divide them
the same way.  Is this possible, or will they have to leave the silver
behind?  Prove your answer.

\begin{solution}
They can divide the coins in such a way only if the number of coins is
$\equiv 45 = 0+1+2+\ldots+9 \pmod{10}$.  Hence the number of gold
coins is $ \equiv 5 \pmod{10}$, and the number of silver coins is
therefore $\equiv 0 \mmod {10}$.  As $0 \not\equiv 5 \pmod{10}$, they
cannot divide the silver coins in the same way.
\end{solution}

\ppart There are also 3 sacks in the chest, containing 5, 49, and 51
rubies respectively.  The treasurer of the pirate ship is bored and
decides to play a game with the following rules:
\begin{enumerate}
\item He can join any two piles together into one pile and
\item He can divide a pile with an even number of rubies into two
  piles of equal size.
\end{enumerate} 
He makes one move every day, and he will finish the game when he has
divided the rubies into 105 piles of one.  Is it possible for him to
finish the game?

\begin{solution}
From $(5,49,51)$ you have only three choices: $(54, 51)$, $(49, 56)$,
and $(5,100)$.  Consider $(54,51)$.  You can only combine two elements
or divide one by two (if it's divisble by 2).  Notice that 54 and 51
are divisible by 3.  Any set of numbers you can possibly obtain from
$(54,51)$ must all be divisible by 3 because adding two multiples of 3
produces a multiple of 3, and dividing an even multiple of 3 by 2
still produces a multiple of 3.  Similar reasoning rules out each of
the other pairs, because each has components that share a prime factor
beside 2.
\end{solution}

\eparts
\fi
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\endinput
