\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_gallup_poll}
  \pcomment{from: F05 cp14f, S08 cp13f}
  \pcomment{revised a lot by ARM 12/5/09}
\end{pcomments}

\pkeywords{
 pairwise_independent
 sampling
 variance
 estimate
 confidence
 probability
 uniform
 independent
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
	
  A recent Gallup poll found that 35\% of the adult population of the
  United States believes that the theory of evolution is ``well-supported
  by the evidence.''  Gallup polled $1928$ Americans selected uniformly
  and independently at random.  Of these, $675$ asserted belief in
  evolution, leading to Gallup's estimate that the fraction of Americans
  who believe in evolution is $675/1928 \approx 0.350$.  Gallup claims a
  margin of error of 3 percentage points, that is, he claims to be
  confident that his estimate is within 0.03 of the actual percentage.

\bparts

\ppart\label{varindicator} What is the largest variance an indicator
variable can have?

\begin{solution}
\[
\frac{1}{4}
\]

By Lemma~\bref{bernoulli-variance}, $\variance{H} = pq$.

\begin{staffnotes}
Here's a proof starting from the definition of variance:
\begin{align*}
\variance{H}
     & = \expect{(H-p)^2}
                   & \text{(def of $\variance{}$)}\\
     & = \expect{H^2-2pH+p^2}\\
     & = \expect{H^2} - 2p\expect{H} + p^2
                   & \text{(linearity of $\expect{}$)}\\
     & = p - p^2   & \text{(since $\expect{H} =p$ and $H^2 = H$)}\\
     & = pq        & (q \eqdef 1-p).
\end{align*}

\end{staffnotes}
Noting that $d\, p(1-p)/dp = 2p-1$ is zero when $p=1/2$, it follows that
the maximum value of $p(1-p)$ must be at $p = 1/2$, so the maximum value
of $\variance{H}$ is $(1/2)(1 -(1/2)) = 1/4$.
\end{solution}

\iffalse
\ppart What is the variance of a binomial variable, $B_{n,p}$?
\begin{solution}
\[
B_{n,p} = H_1 + H_2 + \cdots + H_n
\]
where the $H_i$'s are mutually independent indicator variables with
$\pr{H_i=1} = p$.  By independence and part~\eqref{varindicator},
\[
\variance{H_1 + H_2 + \cdots + H_n} = \variance{H_1} + \variance{H_2} + \cdots +
\variance{H_n} = npq.
\]

\end{solution}
\fi

\ppart Use the Pairwise Independent Sampling Theorem to determine a
confidence level with which Gallup can make his claim.

\begin{solution}
By the Pairwise Independent Sampling Theorem, the probability that a sample of
size $n=1928$ is further than $x = 0.03$ of the actual fraction is at most
\[
\paren{\frac{\sigma}{x}}^2 \cdot \frac{1}{n}
\leq \paren{\frac{1/2}{0.03}}^2 \cdot \frac{1}{1928} \leq 0.144
\]
so we can be confident of Gallup's estimate at the 85.6\% level.
\end{solution}

\ppart Gallup actually claims greater than 99\% confidence in his
estimate.  How might he have arrived at this conclusion?  (Just explain
what quantity he could calculate; you do not need to carry out a
calculation.)

\begin{solution}
The variance, $\sigma^2$, of a single sample might really be as bad as
1/4 (this happens when $p=1/2$) and the variance of $n$ samples might
be as bad as $n\sigma^2$, so there is no better mileage to be gotten
out of the Chebyshev bound.

Instead, Gallup uses the fact that the sample has a binomial
distribution $B_{1928,p}$, where $p$ is the unknown quantity to be
estimated. \iffalse We estimate that the unknown fraction $p$ is about
$0.35$, and the question is how much confidence to hold in that
estimate.\fi The tails of a binomial distribution decrease much more
rapidly than arbitrary distributions (see
Section~\bref{binomial_distribution_section}),\iffalse and this
decrease is exponential rather than quadratic\fi{} so confidence degrees
calculated using this distribution will be higher than calculations
based solely on variance.

So Gallup wants an upper bound on
\[
\Prob{\abs{\frac{B_{1928,p}}{1928} - p} > 0.03}
\]
By part~\eqref{varindicator}, the variance of $B_{n,p}$ is largest
when $p =1/2$, which suggests that the probability that a binomial
random variable differs from its mean will be largest
when $p=1/2$.  This is in fact the case.  So Gallup will calculate
\begin{eqnarray*}
\Prob{\abs{\frac{B_{1928,p}}{1928} - p} > 0.03}
    &=& \Prob{\abs{B_{1928,p} - 1928p} > 0.03(1928)}\\
    &\leq& \Prob{\abs{B_{1928,1/2} - 1928(1/2)} > 0.03(1928)}\\
    &=& \pr{906 \leq B_{1928,1/2} \leq 1021}\\
    &=& \frac{\sum_{i=906}^{1021} \binom{1928}{i}}{2^{1928}} \approx 0.9912.
\end{eqnarray*}
\emph{Mathematica} will actually calculate this sum exactly.  There
are also simple ways to use Stirling's formula to get a good estimate
of this value.
\end{solution}

\ppart Accepting the accuracy of all of Gallup's polling data and
calculations, can you conclude that there is a high probability that the
number of adult Americans who believe in evolution is $35 \pm 3$ percent?

\begin{solution}
No.  As explained in Notes and lecture, the assertion that
fraction $p$ is in the range $0.35\pm 0.03$ is an assertion of fact that
is either true or false.  The number $p$ is a \emph{constant}.  We
don't know its value, and we don't know if the asserted fact is true or
false, but there is nothing probabilistic about the fact's truth or
falsehood.

We \emph{can} say that either the assertion is true or else a 1-in-100
event occurred during the poll.  Specifically, the unlikely event is
that Gallup's random sample was unrepresentative.  This may convince you
that $p$ is ``probably'' in the range $0.35 \pm 0.03$, but this informal
``probably'' is not a mathematical probability.
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
