\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_Handshaking_Lemma}
  \pcomment{from: S09.cp6m, S06.cp5w}
\end{pcomments}

\pkeywords{
  handshaking
  degree
  graphs
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

\bparts

\ppart Prove that in every simple graph, there are an even number of
vertices of odd degree.

\hint The Handshaking Lemma~\bref{sumedges}.

\begin{solution}
\begin{proof}
Partitioning the vertices into those of even degree and those of odd
degree, we know
\[
{\sum_{v \in V} d(v)}\ =
\sum_{d(v) \mbox{ {\scriptsize is even}}} d(v)\ +\
\sum_{d(v) \mbox{ {\scriptsize is odd}}} d(v)
\]
By the Handshaking Lemma, the value of the lefthand side of this
equation equals twice the number of edges, and so is even.  The first
summand on the righthand side is even since it is a sum of even
values.  So the second summand on the righthand side must also be even.
But since it is entirely a sum of odd values, it must must contain an even
number of terms.  That is, there must be an even number of vertices with
odd degree.
\end{proof}
\end{solution}

\ppart\label{oddnumeven} Conclude that at a party where some people shake
hands, the number of people who shake hands an odd number of times is an
even number.

\begin{solution}
We can represent the people at the party by the vertices of a
graph.  If two people shake hands, then there is an edge between the
corresponding vertices.  So the degree of a vertex is the number of
handshakes the corresponding person performed.  The result in the first
part of this problem now implies that there are an even number of
odd-degree vertices, which translates into an even number of people who
shook an odd number of hands.
\end{solution}

\ppart Call a sequence of two or more different people at the party a
\emph{handshake sequence} if each person in the sequence has shaken
hands with the next person, if any, in the sequence.

Suppose George was at the party and has shaken hands with an odd number of
people.  Explain why, starting with George, there must be a handshake
sequence ending with a different person who has shaken an odd number of
hands.

\hint Just look at all the people who appear in handshake sequences
that start with George.

\begin{solution}
The handshake graph between just the people who appear in handshake
sequences that start with George is a graph, so by
part~\eqref{oddnumeven}, it must have an even number of people who
shake an odd number of hands.  In particular, there must be at least
one other person besides George, call him Harry, who has also shaken
an odd number of hands.  So the handshake sequence from George that
ends with Harry is what we were looking for.

  One might worry that Harry has even degree in the \emph{full} graph for
  the whole party but not in the subgraph of those with sequences from
  George.  However, clearly every edge in the George-sequence graph is also
  in the full graph.  Further, no person in the George-sequence graph may
  have shaken hands with anyone not in that graph, since otherwise that new
  person would belong in the George-sequence graph.  Thus, all vertices
  included in the new graph have the same degrees as they have in the full
  graph.
\end{solution}

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
