\documentclass[problem]{mcs}

\begin{pcomments}
\pcomment{CP_more-counting}
\pcomment{from: S09.cp10m commented out}
\pcomment{S07 cp10f?, maybe use S08 10f instead?}
\end{pcomments}

\pkeywords{
  counting
  counting_rules
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{problem}
Solve the following counting problems.  Define an appropriate mapping
(bijective or $k$-to-1) between a set whose size you know and the set
in question.

\bparts 

\ppart How many different ways are there to select a dozen donuts if
four varieties are available?

\begin{solution}
There is a bijection from selections of a
dozen donuts to 15-bit sequences with exactly 3 ones.  In particular,
suppose that the varieties are glazed, chocolate, lemon, and Boston
creme.  Then a selection of $g$ glazed, $c$ chocolate, $l$ lemon, and
$b$ Boston creme maps to the sequence:
%
\[
(g\ 0's)\ 1\ (c\ 0's)\ 1\ (l\ 0's)\ 1\ (b\ 0's)
\]
%
Therefore, the number of selections is equal to the number of 15-bit
sequences with exactly 3 ones, which is:
%
\[
\frac{15!}{3!\ 12!} = \binom{15}{3}
\]

\end{solution}

\ppart How many paths are there from $(0, 0)$ to $(10, 20)$ consisting
of right-steps (which increment the first coordinate) and up-steps
(which increment the second coordinate)?

\begin{solution}
There is a bijection from 30-bit sequences with 10 zeros and 20
ones.  The sequence $(b_1, \ldots, b_{30})$ maps to a path where the
$i$-th step is right if $b_i = 0$ and up if $b_i = 1$.  Therefore, the
number of paths is equal to $\binom{30}{10}$.
\end{solution}

\ppart An independent living group is hosting nine new candidates for
membership.  Each candidate must be assigned a task: 1 must wash pots, 2
must clean the kitchen, 3 must clean the bathrooms, 1 must clean the
common area, and 2 must serve dinner.  In how many ways can this be done?

\begin{solution}
There is a bijection from sequences containing one
$P$, two $K$'s, three $B$'s, a $C$, and two $D$'s.  In any such
sequence, the letter in the $i$th position specifies the task assigned to
the $i$th candidate.  Therefore, the number of
possible assignments is:
\[
\frac{9!}{1!\ 2!\ 3!\ 1!\ 2!}
\]

\end{solution}

\ppart In how many ways can Mr. and Mrs. Grumperson distribute 13
identical pieces of coal to their two--- no, three!--- children for
Christmas?

\begin{solution}
There is a bijection from 15-bit strings
with two ones.  In particular, the bit string $0^a10^b10^c$ maps to
the assignment of $a$ coals to the first child, $b$ coals to the
second, and $c$ coals to the third.  Therefore, there are
$\binom{15}{2}$ assignments.
\end{solution}

\ppart How many solutions over the natural numbers are there to the
equation:

\begin{eqnarray*}
x_1 + x_2 + \ldots + x_{10} & \leq & 100
\end{eqnarray*}

\begin{solution}
There is a bijection from 110-bit sequences
with 10 ones to solutions to this equation.  In particular, $x_i$ is
the number of zeros before the $i$-th one but after the $(i-1)$-st one
(or the beginning of the sequence).  Therefore, there are
$\binom{110}{10}$ solutions.
\end{solution}

\eparts
\end{problem}

\endinput
