\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_matching_more_boys}
  \pcomment{by ARM 3/15/11}
  \pcomment{similar to CP_stable_marriage_invariant_review,
            but requires slightly more insight}
  \pcomment{less general than PS_stable_matching_hospitals}
  \pcomment{may need mating ritual description at end of file}
\end{pcomments}

\pkeywords{
 stable_matching
 Mating_ritual
 invariant
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Suppose there are more boys than girls.

\bparts

\ppart Define what a stable matching should mean in this case.

\begin{solution}
The definition of rogue couple in a matching should now include
an unmarried boy and a married girl that likes the boy better than
her husband.
\end{solution}

\ppart Explain why applying the Mating Ritual in this case will yield a
stable matching in which every girl is married.

\begin{solution}
The same invariant holds, and the same proof applies, in this
case, except at the very last step: since there are more boys than girls,
the fact that all the girls are married does not imply that all the boys
are (and of course there have to be some unmarried boys at the end, since
polyandry can't happen).  The proof of stability then goes through
unchanged for married boys, and for an unmarried boy, Billy Bob, we need
only observe that the Invariant implies all the girls like their spouses
better than Billy Bob, so he can't be in a rogue couple either.
\end{solution}

\eparts
\end{problem}

\endinput

\inhandout{\instatements{
\newpage
\section*{Appendix: The Mating Ritual}

The \emph{Mating Ritual} takes place over several days.  The
following events happen each day:

{\bf Morning: } Each girl stands on her balcony.  Each boy stands under
the balcony of his favorite among the girls on his list, and he serenades
her.  If a boy has no girls left on his list, he stays home and does his
6.042 homework.

{\bf Afternoon: } Each girl who has one or more suitors serenading her,
says to her favorite suitor, ``We might get engaged.  Come back
tomorrow.''  To the others, she says, ``No.  I will never marry you!  Take
a hike!''

\textbf{Evening}: Any boy who is told by a girl to take a hike, crosses that
girl off his list.

\textbf{Termination condition}: When every girl has at most one suitor,
the ritual ends with each girl marrying her suitor, if she has one.
}}
