\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_min_weight_span_tree}
  \pcomment{needs work; maybe unsalvageable --ARM Oct '11}
  \pcomment{from F07.rec7t, S06.cp5f, S04.cp5w}
  \pcomment{special case of general theory in text}
\end{pcomments}

\pkeywords{
spanning_tree
weighted_tree
minimum_weight
MST
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
A \emph{weighted graph} is a graph with a function that assigns a
nonnegative real number to each edge.  The real number assigned to an
edge is called its \emph{weight}.  The weight of the graph is the sum
of the weights of its edges.

A \emph{minimum weight spanning tree} of a weighted graph is a
spanning tree with the smallest weight of any spanning tree of the
graph.

\begin{editingnotes}
An example would be nice.
\end{editingnotes}

Let $G$ be a weighted graph.

\bparts

\ppart Define an $n$-component subgraph of $T$ of $G$ is $n$-minimal
when $T$ has $n$ vertices, is connected, and has minimum weight among
all

Show that every minimum weight spanning tree must contain at least
one edge that has the smallest edge weight.

\begin{solution}
Suppose that an edge $e$ with smallest weight is not included in
some minimal spanning tree; in other words, suppose that the minimum
spanning tree $T$ contains only edges with weights larger than that of
$e$.  If we add $e$ to $T$, then we will obtain a
graph with this one cycle that includes $e$.  We can then delete
some other edge in the cycle, without disconnecting any of the
vertices in that cycle.  This results in a spanning tree with weight
strictly less than that of $T$ (since all the other edges have larger
weight than $e$ has), contradicting the minimality of $T$.  We conclude
that an edge with smallest weight must be included in $T$.
\end{solution}

\ppart Use the result of the previous part to describe a simple procedure
for contructing a minimum weight spanning tree.  You may assume for
simplicity that all edge weights are distinct.

\begin{solution}
Keep picking a smallest weight edge that does not create a cycle
of picked edges ---a classic ``greedy'' algorithm.

\begin{staffnotes}
Inadequate answer.  Not clear how part(a) implies greedy is OK.
\end{staffnotes}
\end{solution}

\eparts

\end{problem}

\endinput
