\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_no_odd_length_closed_walks}
  \pcomment{rephrasing of PS_no_odd_length_cycles}
  \pcomment{subsumed by text}
  \pcomment{from: S10?, S09.ps6, S07.ps5, S06.ps4 S04.quiz1, S04.cp5w}
\end{pcomments}

\pkeywords{
  closed_walk
  walk
  graph_coloring
  2-color
  bipartite
  trees
  spanning_trees
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
In this problem you will prove:
\begin{theorem*}
A graph $G$ is 2-colorable iff it contains no odd length closed walk.
\end{theorem*}

As usual with ``iff'' assertions, the proof splits into two proofs:
part~(a) asks you to prove that the left side of the ``iff'' implies the
right side.  The other problem parts prove that the right side implies the
left.

\bparts

\ppart Assume the left side and prove the right side.  Three to five
sentences should suffice.

\begin{solution}
  Assume $G$ is 2-colorable and select a 2-coloring of $G$.  Consider an
  arbitrary closed walk through successive vertices $v_1, v_2, \dots, v_k,
  v_1$.  Then the vertices $v_i$ must be one color for all even $i$ and
  the other color for all odd $i$.  (This is obvious, but could, of course,
  be proved by induction.)  Since $v_1$ and $v_k$ must be colored
  differently, $k$ must be even.  Thus, the walk has even length.
\end{solution}

\ppart Now assume the right side.  As a first step toward proving the
left side, explain why we can focus on a single connected component
$H$ within $G$.

\begin{solution}
If we can 2-color every connected component of $G$, then we can
2-color all of $G$.  Thus, it suffices to show that an arbitrary connected
component $H$ of $G$ is 2-colorable.
\end{solution}

\ppart\label{2colortree} As a second step, explain how to 2-color any tree.

\begin{solution}
A 2-coloring of a tree can be defined by selecting any fixed
vertex $v$, and coloring a vertex one color if the (unique) path to
it from $v$ has odd length, and coloring it with the other color if the
path has even length.

To verify that adjacent vertices in the tree get different colors, let
$e \eqdef \edge{x}{y}$ be an edge in the tree.  There is a unique path
from $v$ to $x$.  If this path includes $e$, it must consist of a path
from $v$ to $y$ followed by the edge $e$ and vertex $x$.  If this path
does not include $e$, then it can be extended to a path to $y$ by
adding $e$ and vertex $y$.  In either case, the paths to these
vertices from $v$ differ by a single traversal of $e$, and so the
lengths of the paths differ by 1; in particular, one must have odd
length and the other must have even length, so $x$ and $y$ are
differently colored.
\end{solution}

\ppart Choose any 2-coloring of a spanning tree, $T$, of $H$.  Prove that
$H$ is 2-colorable by showing that any edge \emph{not} in $T$ must also
connect different-colored vertices.

\begin{solution}
Let $v$ be any vertex in the spanning tree $T$.  Then vertices at an
even distance in $T$ from $v$ must have the same color as $v$ and
vertices at an odd distance in $T$ must have a different color from
$v$.

Let $\edge{x}{y}$ be an edge not in $T$.  Since $T$ is a tree, there
are unique paths in $T$ from $v$ to $x$ and from $v$ to $y$.  Exactly
one of these two paths must have odd length or else these two paths
together with the edge $\edge{x}{y}$ would form an odd length closed
walk.  So one path from $v$ has odd length and the other path from $v$
has even length, and hence $x$ and $y$ have different colors.
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
