\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{CP_prenex}
    \pcomment{by ARM 2/9/16}
\end{pcomments}

\pkeywords{
  prenex
  quantifier
  De_Morgan
  negation
  variable
  variable_convention
}

\begin{problem}
A predicate formula is in \emph{prenex form} when all its quantifiers come at the
beginning.  For example,

\TBA{prenex example}

It turns out that every formula is equivalent to one in prenex form.

Here's a simple recipe for converting a formula into an equivalent prenex form:

\begin{enumerate}

\item Rename variables to satisfy the \emph{variable convention}
  described in Problem~\bref{CP_variable_convention},

\item\label{toandor} Convert all connectives to \QAND, \QOR, \QNOT,

\item\label{appDeM} Use De Morgan Laws for negated quantifiers (Section~\bref{negquant_sec}) to push
  negations ``below'' all quantifiers,

\item\label{qtfront} Pull out all quantifiers to front of formula in any order that
  preserves nesting.  That is, if $\exists x$ occurs in a subformula starting with
  $\forall y$, then $\forall y$ has to come before $\exists x$ when the quantifiers
  are moved to the front.

\end{enumerate}

Use this method to find a prenex form for
\begin{staffnotes}
Nice to have an example where step~\ref{toandor}.\ needed a second
repeat andat a separate location.  Maybe start off without satisfying
variable convention.
\end{staffnotes}

\[
[\forall x\, \exists y_1.\, P(x) \QAND Q(x,y_1)]\ \QIMPLIES 
  (\exists x_1.\, R(x_1,z)) \QOR \exists x_2\, \forall z_1.\, S(x_2,y,w,z_1),
\]
which already satisfies the variable convention.

\begin{solution}
\begin{align*}
\lefteqn{[\forall x\, \exists y_1.\, P(x) \QAND Q(x,y_1)]\ \QIMPLIES}\\ 
  & \qquad\qquad   (\exists x_1.\, R(x_1,z)) \QOR \exists x_2\, \forall z_1.\, S(x_2,y,w,z_1)\\
 & \corresp
  \QNOT(\forall x\, \exists y_1.\, P(x) \QAND Q(x,y_1))\ \QOR\\
  & \qquad\qquad   (\exists x_1.\, R(x_1,z)) \QOR \exists x_2\, \forall z_1.\, S(x_2,y,w,z_1)
         &\text{(step~\ref{toandor}.)}\\
 & \corresp
  \exists x\, \QNOT(\exists y_1.\, P(x) \QAND Q(x,y_1)))\ \QOR\\
  & \qquad\qquad   (\exists x_1.\, R(x_1,z)) \QOR \exists x_2\, \forall z_1.\, S(x_2,y,w,z_1)
         &\text{(start step~\ref{appDeM}.)}\\
 & \corresp
  \exists x\,\forall y_1.\, \QNOT(P(x) \QAND Q(x,y_1))\ \QOR\\
  & \qquad\qquad  (\exists x_1.\, R(x_1,z)) \QOR \exists x_2\, \forall z_1.\, S(x_2,y,w,z_1)
         &\text{(repeat step~\ref{appDeM}.)}\\
 & \corresp
  \exists x\,\forall y_1\, \exists x_1.\,  \exists x_2\, \forall z_1.\,\\
  & \qquad\qquad  \QNOT(P(x) \QAND Q(x,y_1)) \QOR  R(x_1,z) \QOR  S(x_2,y,w,z_1).
         &\text{(step~\ref{qtfront}.)}\\
\end{align*}

\end{solution}

\end{problem}

\endinput
