\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{from: S08 cp9r}
\end{pcomments}

\pkeywords{
  pigeon hole
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%S08 cp9r

\begin{problem}
Solve the following problems using the pigeonhole principle.  For each
problem, try to identify the {\em pigeons}, the {\em pigeonholes}, and
a {\em rule} assigning each pigeon to a pigeonhole.

\bparts

\iffalse

\ppart In a room of 500 people, there exist two who share a birthday.

\begin{solution}The pigeons are the 500 people.  The
pigeonholes are 366 possible birthdays.  Map each person to his or her
own birthday.  Since there 500 people and 366 birthdays, some two
people must have the same birthday by the Pigeonhole Principle.
\end{solution}

\fi

\ppart Every MIT ID number starts with a 9 (we think).  Suppose that
each of the 75 students in 6.042 sums the nine digits of his or her ID
number.  Explain why two people must arrive at the same sum.

\begin{solution}The students are the pigeons, the possible sums are the
  pigeonholes, and we map each student to the sum of the digits in his or
  her MIT ID number.  Every sum is in the range from $9 + 8 \cdot 0 = 9$
  to $9 + 8 \cdot 9 = 81$, which means that there are 73 pigeonholes.
  Since there are more pigeons than pigeonholes, there must be two pigeons
  in the same pigeonhole; in other words, there must be two students with
  the same sum.
\end{solution}

\iffalse

\ppart In every set of 100 integers, there exist two whose difference
is a multiple of 37.

\begin{solution}The pigeons are the 100 integers.  The pigeonholes are the
  numbers 0 to 36.  Map integer $k$ to $\rem{k}{37}$.  Since there are
  100 pigeons and only 37 pigeonholes, two pigeons must go in the same
  pigeonhole.  This means $\rem{k_1}{37} = \rem{k_2, 37}$, which
  implies that $k_1 - k_2$ is a multiple of 37.
\end{solution}
\fi

\ppart For any five points inside a unit square (not on the boundary),
there are two points at distance \emph{less than} $1/\sqrt{2}$.

\begin{solution}
  The pigeons are the points.  The pigeonholes are the four
  subsquares of the unit square, each of side length 1/2.

  Pigeons are assigned to the subsquare that contains them, except that if
  the pigeon is on a boundary, it gets assigned to the leftmost and then
  lowest possible subsquare that includes it (so the point at $(1/2,1/2)$
  is assigned to the lower left subsquare).

  There are five pigeons and four pigeonholes, so more than one point must
  be in the same subsquare.  The diagonal of a subsquare is $1/\sqrt{2}$,
  so two pigeons in the same hole are at most this distance.  But pigeons
  must be inside the unit square, so two pigeons cannot be at the
  opposite ends of the same subsquare diagonal.  So at least one of them
  must be inside the subsquare, so their distance is less than the length
  of the diagonal.
\end{solution}

\iffalse

\ppart For any five points in an equilateral triangle of side length
2, there are two points at distance less than $1$.

\begin{solution}The pigeons are the points.  The pigeonholes are the four
  sub-equilateral triangles of side length 1.  There are five pigeons
  and four pigeonholes, so more than one point must be in the same
  sub-equilateral triangle.  Points in the same sub-equilateral
  triangle are at distance at most $1$. 
\end{solution}

\fi

\iffalse

\ppart Let $\{a_1, \ldots, a_{201}\}$ be a set of natural numbers less
than 300.  Then there are $i,j$ such that $\frac{a_i}{a_j} = 3^k$ for
some $k > 0$.

\begin{solution}The pigeons are the numbers.  The pigeonholes are the 200
  numbers less than 300 that are not divisible by 3.  Write each
  number $a_i = 3^{c_i} \cdot b_i$, where $b_i$ is not divisible by 3.
  Place $a_i$ in the pigeonhole $b_i$.  Then there are two numbers
  $a_i, a_j$ placed in the same pigeonhole, that is, $b_i = b_j$, so
  $\frac{a_i}{a_j} = 3^{c_i-c_j}$. (Assume $c_i > c_j$ without loss of
  generality)
\end{solution}

\fi

\ppart Show that if $n+1$ numbers are selected from $\set{1,2,3,
  \dots,2n}$, two must be consecutive, that is, equal to $k$ and $k+1$ for
some $k$.

\begin{solution}
  The pigeonholes will be the $n$ sets $\set{1,2}, \set{3,4},
  \set{5,6},\dots, \set{2n-1,2n}$.  The pigeons will be the $n+1$ selected
  numbers.  A pigeon is assigned to the unique hole of which it is a
  member.  By the Principle, two pigeons must assigned to some hole, and
  these are the two consecutive numbers required.  Notice that we've
  actually shown a bit more: there will be two consecutive numbers with
  the smaller being odd.
\end{solution}

\iffalse

\ppart Suppose $n+1$ numbers are selected from $\set{1,2,3, \dots,2n}$.
Show that there must be two selected numbers whose quotient is a power of
two.

\begin{solution}
  The $n+1$ selected numbers are the pigeons.  There are $n$ odd
  numbers between 1 and $2n$, namely, the numbers of the form $2k+1$ for
  $0 \leq k <2n$.  The pigeonholes will be the sets consisting of
  power-of-two multiples of each of these odd numbers.  For example, the
  third odd number is 5, so the third pigeonhole will be the set of
  multiples of 5 by a power of two: $\set{5,10,20,40,\dots}$.  Two pigeons
  must assigned to some hole, and these are the required two selected
  numbers, since the quotient of the larger of these two by the smaller
  will be a power of two by definition.
\end{solution}

  \fi

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
