\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_roots_of_polynomials}
  \pcomment{from: S09.cp1r, S08.cp2m(?), S99 Tutorial 1 Notes}
  \pcomment{edited ARM 1/16/12}
\end{pcomments}

\pkeywords{
  irrational
  primes
  polynomial
  root
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Here is a generalization of Problem~\bref{CP_generalize_root_2_proof}
that you may not have thought of:

\begin{lemma*}
Let the coefficients of the polynomial
\[
a_0+a_1x+a_2x^2+\cdots + a_{m-1}x^{m-1} + x^m
\]
be integers.  Then any real root of the polynomial is either integral
or irrational.
\end{lemma*}

\bparts

\ppart Explain why the Lemma immediately implies that $\sqrt[m]{k}$ is
irrational whenever $k$ is not an $m$th power of some integer.

\begin{solution}
Saying that an integer, $k$, is not the $m$th power of an integer, is
equivalent to saying that the equation $x^m = k$ has no integer
solutions.  Another way to say this is that the polynomial $x^m - k$
has no integer root.  The Lemma therefore implies that any root of
$x^m-k$ is irrational.  But $\sqrt[m]{k}$ is, by definition, a root of
this polynomial, so it is irrational.
\end{solution}

\ppart Carefully prove the Lemma.

You may find it helpful to appeal to:
\begin{fact*}
  If a prime, $p$, is a factor of some power of an integer, then it is a
  factor of that integer.
\end{fact*}
You may assume this Fact without writing down its proof, but see if
you can explain why it is true.

\begin{staffnotes}
Ask that a clear, textbook quality proof of the Lemma be written on
the team whiteboard.  Point out that besides clarity and correctness,
textbook quality requires good English with proper punctuation.  When
a real textbook writer does this, it usually takes multiple revisions.
If students are satisfied with their first draft, they're probably
misjudging.
\end{staffnotes}

\begin{solution}
%From S08 class problem 2M(?) and S99 Tutorial 1 Notes:

\begin{proof}

Let $r$ be a real root of the polynomial, so that
\[
a_0+a_1r+a_2r^2+\cdots+  a_{m-1}r^{m-1}+ r^m = 0.
\]
There are three cases: either $r$ is an integer, or $r$ is irrational, or
$r = s/t$ for integers $s$ and $t$ which have no common factors and such
that $t >1$.  We want to eliminate the last case, so assume for the sake
of contradiction that it held for some $r$.

Substituting $s/t$ for $r$ and multiplying both sides of the above
equation by $t^m$ yields:
\begin{align}
a_0t^m+a_1st^{m-1}+a_2s^2t^{m-2}+ \cdots + a_{m-1}s^{m-1}t+s^m & =  0,\\
a_0t^m+a_1st^{m-1}+a_2s^2t^{m-2}+ \cdots + s^{m-1}t & = -s^m.\label{e2}
\end{align}

Now since $t>1$, it must have a prime factor, $p$.  The prime, $p$,
therefore divides each term of the lefthand side of
equation~\eqref{e2}, so $p$ also divides the righthand side, $-s^m$.
This means that $p$ divides $s^m$, so by the Fact above, $p$ is also a
factor of $s$.  So $p$ is a common factor of $s$ and $t$,
contradicting the fact that $s$ and $t$ have no common factors.
\end{proof}

The Fact above about primes dividing powers is a simple consequence of
the \emph{Fundamental Theorem of Arithmetic} which says that every
integer $> 1$ factors into a product of primes that is \emph{unique}
except for the order in which the primes are multiplied.

For example, here are some ways to express 140 as a product of primes:
\[
140 = 2\cdot 2 \cdot 5 \cdot 7 = 2\cdot 5 \cdot 7 \cdot 2 = 7 \cdot 5 \cdot
2\cdot 2 = \cdots.
\]
By the Fundamental Theorem, every such product will have exactly two
occurrences of 2 and one each of 5 and 7.  Next, we can obviously get a
product of primes equal to, say, the third power of 140 by taking a
product that equals $140$ and repeating it three times.  For example,
\[
(140)^3 = 2\cdot 2 \cdot 5 \cdot 7 \ \ \cdot \ \  2\cdot 2 \cdot 7 \cdot 5
\ \ \cdot \ \ 2\cdot 2 \cdot 7 \cdot 5.
\]
The Fundamental Theorem now says that \emph{every} prime product equal to
the third power of 60 must have the same primes as this repeated product,
namely, six occurrences of 2 and three occurrences each of 5 and 7.  In
particular, the \emph{only} primes that are factors of $(140)^3$ are the
primes 2, 5 and 7 that are factors of 140.  This reasoning applies equally
well with any other integer greater than 1 in place of 140 and any power
greater than 0 in place of 3, proving that if $p$ is a prime factor of
$s^m$, then $p$ must have been a factor of $s$.

The Fundamental Theorem of Arithmetic is also know as the \emph{Unique
  Prime Factorization Theorem}.  It is one of those familiar mathematical
facts that is not exactly obvious.  We'll work out a proof of the
Fundamental Theorem in a later chapter.
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
