\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_sixteen_ions}
  \pcomment{from: S09.cp14m}
  \pcomment{rephrased by DIG from CP_sixteen_desks, 11/19/17}
\end{pcomments}

\pkeywords{
  expectation
  linearity
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem} A novice mad scientist is attempting to put ions together into a salt crystal. The crystal has sixteen ion locations in a $4 \times 4$ arrangement as shown below.
%
\begin{center}
\begin{picture}(300,160)
% \put(0,0){\dashbox(270,198){}} % bounding box
\multiput(0,  0)(72,0){4}{\framebox(40,25){}}
\multiput(0, 40)(72,0){4}{\framebox(40,25){}}
\multiput(0, 80)(72,0){4}{\framebox(40,25){}}
\multiput(0,120)(72,0){4}{\framebox(40,25){}}
\end{picture}
\end{center}
%
If there is a sodium ion in front, behind, to the left, or to the right of a
chloride ion, then the two \emph{bond}.  One ion may be in
multiple bonded pairs; for example, an ion in a corner of the
crystal can bond with up to two others, while an ion in the
center can bond with as many as four others.  Suppose that the locations are
occupied mutually independently by sodium and chloride ions with equal probability.  What is the expected number of bonded pairs?
\hint Linearity.

\begin{solution}
A natural first approach to this problem is to calculate the expected
number of bonds that each ion is involved in, add the
expectations for each ion, and then divide by two (since each
bond involves two ions).  This approach works fine, but it
requires finding the expectations for three different kinds of ion locations:
corner, side, and middle.

A more elegant approach is to note that the expected number of
bonds between adjacent ions is 1/2, and the number of pairs of
adjacent ions is 24---there are 12 pairs adjacent horizontally (3 in
each of 4 rows) and likewise 12 pairs adjacent vertically.  So by
linearity of expectation, the expected number of bonds is
$(1/2)\cdot 24 = 12$.

To be more explicit about this application of linearity, let's
arbitrarily number the pairs of adjacent ions from 1 to 24 and let
$B_i$ be an indicator random variable for the event that ions in the $i$-th pair bond.  Adjacent ions bond iff they are of different types, which happen with
probability 1/2, so $\pr{B_i = 1} = 1/2$.  The expectation of an
indicator variable is the same as the probability that it equals
1\inbook{(Lemma~\bref{expindic})}, so
\begin{equation}\label{fi=12}
\expect{B_i} = \frac{1}{2}.
\end{equation}
Now, if $B$ is the number of bonds, then $B =
\sum_{i=1}^{24} B_i$, so the expectation we want is
\begin{align*}
\expect{B}
        & = \Expect{\sum_{i=1}^{24} B_i}\\
	& = \sum_{i=1}^{24} \expect{B_i} 
			& \text{(linearity of $\expect{\cdot}$)}\\
	& = \sum_{i=1}^{24} \frac{1}{2} 
			& \text{(equation~\eqref{fi=12})}\\
        & = 24 \cdot \frac{1}{2} = 12.
\end{align*}

\end{solution}
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
