!\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_st_petersberg}
  \pcomment{F07.rec14h}
\end{pcomments}

\pkeywords{
  deviation
  pairwise_independent
  paradox
  roulette
  mean_time_to_failure
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
A gambler bets \$10 on ``red'' at a roulette table (the odds of red are
18/38 which slightly less than even) to win \$10.  If he wins, he gets
back twice the amount of his bet and he quits.  Otherwise, he doubles his
previous bet and continues.

\bparts

\ppart What is the expected number of bets the gambler makes before he
wins?

\begin{solution}

This is mean time to failure, with failure being a red numnber coming up.  So
the expected time (number of bets) is
\[
\frac{1}{18/38} = 2\ \frac{1}{9}.
\]
\end{solution}

\ppart What is his probability of winning?

\begin{solution}
He is certain to win, since $\pr{> k \text{ bets}} = (20/38)^{k}$ which
goes to zero as $k$ goes to infinity.  More fully,
\[
\pr{\text{win}} \geq \pr{\text{win in $\leq k$ bets}}
= 1 - \pr{> k \text{ bets}}
\]
and this last expression goes to 1 as $k$ goes to infinity.
\end{solution}

\ppart What is his expected final profit (amount won minus amount lost)?

\begin{solution}
His final profit is always \$10 whenever he finally wins, and he is
certain to win, so \$10 is also his expected final profit.
\end{solution}

\ppart The fact that the gambler's expected profit is positive,
despite the fact that the game is biased against him, is known as the
\term{St. Petersberg paradox}.  The paradox arises from an
unrealistic, implicit assumption about the gambler's money.  Explain.

\hint What is the expected size of his last bet?

\begin{solution}

It sounds plausible that, since his expected number of bets is less
than three, the expected size of his bet would be less than the size
of his third bet, that is, $10\cdot 2^2 = 40$ dollars.  But this is a
sloppy---and wrong---argument.  What it overlooks is that later bets,
though progressively less likely, grow tremendously, and so contribute
heavily to the expected bet size.

To get the answer, we go back to the definition of expected bets.  Let $B$
be the size of his last bet in dollars.  Now if he wins his $\$10$ final
profit on the $k$th bet, then $B=10\cdot 2^{k-1}$, so
\[
\pr{B=10\cdot 2^{k-1}} = (20/38)^{k-1}(18/38)
\]
So
\begin{align*}
\expect{B} & = \sum_{k \in \naturals^{+}} 10\cdot
      2^{k-1}\paren{\frac{20}{38}}^{k-1}(18/38)\\
  & = 10(18/38) \sum_{k \in \naturals} 2^k\paren{(20/38)^k}\\
  & > \sum_{k \in \naturals} \paren{1+ \frac{1}{19}}^k= \infty
\end{align*}
so the gambler has to have an infinite bank account to win \$10 with
certainty.

\end{solution}

\begin{staffnotes}

Maybe ask to verify that his expected gain is negative if he only has
a finite number, $n$, of dollars to bet.

\end{staffnotes}

\eparts

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
