\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{from: S09.cp1t}
%  \pcomment{}
%  \pcomment{}
\end{pcomments}

\pkeywords{
  faulty_reasoning
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Albert announces to his class that he plans to surprise them with a
quiz sometime next week.

His students first wonder if the quiz could be on Friday of next next.
They reason that it can't: if Albert didn't give the quiz
\emph{before} Friday, then by midnight Thursday, they would know the
quiz had to be on Friday, and so the quiz wouldn't be a surprise any
more.

Next the students wonder whether Albert could give the surprise quiz
Thursday.  They observe that if the quiz wasn't given \emph{before}
Thursday, it would have to be given \emph{on} the Thursday, since they
already know it can't be given on Friday.  But having figured that
out, it wouldn't be a surprise if the quiz was on Thursday either.
Similarly, the students reason that the quiz can't be on Wednesday,
Tuesday, or Monday.  Namely, it's impossible for Albert to give a
surprise quiz next week.  All the students now relax, having concluded
that Albert must have been bluffing.

And since no one expects the quiz, that's why, when Albert gives it on
Tuesday next week, it really is a surprise!

What do you think is wrong with the students' reasoning?

\begin{solution}
The basic problem is that ``surprise'' is not a mathematical
concept, nor is there any generally accepted way to give it a mathematical
definition.  The ``proof'' above assumes some plausible axioms about
surprise, without defining it.  The paradox is that these axioms are
inconsistent.  But that's no surprise \smiley---mathematically
speaking---we don't know what we're talking about.

Mathematicians and philosophers have had a lot more to say about what might
be wrong with the students' reasoning,---see Chow, Timothy Y.
\href{http://courses.csail.mit.edu/6.042/spring09/surprise-paradox.pdf}
{\emph{The surprise examination or unexpected hanging paradox}}, American
Mathematical Monthly (January 1998), pp.~41--51.
\end{solution}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
