\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_three_fair_coins}
  \pcomment{from: S09.cp13m}
\end{pcomments}

\pkeywords{
  conditional_probability
  independence
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% F09, S09

\begin{problem}
Suppose that you flip three fair, mutually independent coins.
Define the following events:
\begin{itemize}
\item Let $A$ be the event that \emph{the first} coin is heads.
\item Let $B$ be the event that \emph{the second} coin is heads.
\item Let $C$ be the event that \emph{the third} coin is heads.
\item Let $D$ be the event that \emph{an even number of} coins are heads.
\end{itemize}

\bparts

\ppart Use the four step method to determine the probability space for
this experiment and the probability of each of $A,B,C,D$.

\begin{solution}
The tree is a binary tree with depth 3 and 8 leaves.  The
  successive levels branching to show whether or not the successive events
  $A,B,C$ occur.  By definition of \emph{fair} and \emph{independent},
  each branch out of a vertex is equally likely to be followed.  So the
  probability space has as outcomes the eight length-3 strings of
  $\heads$'s and $\tails$'s, each of which has probability
  $(1/2)^3 = 1/8$.

  Each of the events events $A,B,C,D$ are true in four of the outcomes
  and hence has probability 1/2.
\end{solution}

\ppart Show that these events are not mutually independent.

\begin{solution}
\[
\pr{A \cap B \cap C \cap D} = 0 \neq (1/2)^4 = \pr{A} \cdot \pr{B} \cdot
\pr{C} \cdot \pr{D}.
\]
\end{solution}

\ppart Show that they are 3-way independent.

\begin{solution}
3-way independence implies that any subset of \emph{up to} three events must be independent.

Because the coin tosses are mutually independent, we know:
\[
\pr{A \cap B \cap C} = \pr{A} \cdot \pr{B} \cdot \pr{C}.
\]

and we know similarly that for any two of the events $A,B,C$,
\[
\pr{A \cap B} = \pr{A} \cdot \pr{B}
\]

with the remaining pairs holding as well by symmetry.

We must also check that the equality holds for the other subsets of
three events: $\set{A,B,D}$, $\set{A,C,D}$, and $\set{B,C,D}$.  By
symmetry, again, we need only check one, say the first one.
\[
\pr{A \cap B \cap D} = \pr{\set{HHT}} = \dfrac{1}{8}.
\]

This is equal to $\pr{A} \cdot \pr{B} \cdot \pr{D}$.

What remains to be checked is that the equality holds for any subset of two events including $D$, namely $\set{A,D}$, $\set{B,D}$, and $\set{C,D}$. By symmetry, once more, we need only 
check one, say the first one.

\[
\pr{A \cap D} = \pr{\set{HHT,HTH}} = \dfrac{1}{4}.
\]

This is equal to equal to $\pr{A} \cdot \pr{D}$.

We may now conclude that all four events are 3-way independent.
\end{solution}

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
