\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{exit-diagnostic-short-S14}
  \pcomment{JD 5/14/14; revised by CH}
\end{pcomments}

\pkeywords{
  probability
  counting
  logic
  relation
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}

\mbox{}

\bparts

\ppart %\mbox{}

\begin{romanlist}
\item The power set of the set $\set{a, b, \emptyset}$ is:
\begin{center}
\exambox{5.0in}{0.4in}{0.0in}
\end{center}

\begin{solution}
\[
\set{ \emptyset, \set{a}, \set{b}, \set{\emptyset}, \set{a,b},
  \set{\emptyset, a}, \set{\emptyset, b}, \set{a, b, \emptyset} }
\]
\end{solution}

\examspace[0.1in]

\item 
What is the number of elements in the power set of $\set{2, 4,
  \dots, 14 }$?\hfill \examrule[0.5in]

\begin{solution}
\[
2^7 = 128 .
\]
\end{solution}

\item Circle the condition(s) under which the size of the
  \textbf{intersection} $A \intersect B$ of two finite sets $A$ and
  $B$ is the sum of their sizes.

\begin{center}
Always \qquad $A \union B = \emptyset$ \qquad $A \subset B$ \qquad
$A \intersect B \neq \emptyset$ \qquad Never
\end{center}

\begin{solution}
\[
A \union B = \emptyset.
\]
\end{solution}

\end{romanlist}

\ppart There are twenty families in a small village.  Four families
each have exactly 1 child, seven families each have exactly 2 children
and among these five families have children of different ages and two
families have twins, three families each have exactly 3 children, and
a single family has exactly 4 children.  The remaining families have
no children.

A family was chosen uniformly at random.  What is the probability of
the family having:
\begin{romanlist}

\item no children at all?\hfill \examrule[0.5in]

\examspace[0.2in]

\begin{solution}
\[
\frac{5}{20} = \frac{1}{4} .
\]
\end{solution}

\item exactly 2 or 3 children?\hfill \examrule[0.5in]

\examspace[0.2in]

\begin{solution}
\[
\frac{1}{2}
\]
\end{solution}

\item fewer than 3 children?\hfill \examrule[0.5in]

\begin{solution}
\[
\frac{16}{20} = \frac{4}{5} .
\]
\end{solution}

\end{romanlist}

\ppart Explain the following concepts.

\begin{romanlist}
\item Walk relation in a digraph:

\begin{center}
\exambox{6.0in}{1.0in}{0in}
\end{center}

\begin{solution}
  The walk relation $G^*$ of a digraph $G$ is the binary relation on
  the vertices of the digraph such that $v\mrel{G^*}w$ if and only if
  there is a directed walk from vertex $v$ to $w$.
\end{solution}

\item Mutual independence:

\begin{center}
\exambox{6.0in}{1.0in}{0in}
\end{center}

\begin{solution}
Informally, events $A_1, A_2,\ldots,A_n$ are mutually independent when
the probability of each is unchanged given which of the other ones
occur.  Formally, they are mutually independent iff for all $i \in \Zintv{1}{n}$
and $J \subseteq \Zintv{1}{n} - \set{i}$
\[
\prob{A_i} = \prcond{A_i}{\lgintersect_{j \in J} \pr{A_j}}.
\]
An equivalent definition is that for every nonempty\footnote{Equality
  is guaranteed to hold when $J=\emptyset$ by the conventions for
  empty products and intersections.} subset $J \subseteq \Zintv{1}{n}$
\[
\Prob{\lgintersect_{j \in J} A_j} = \prod_{j \in J} \pr{A_j}.
\]
\end{solution}

\end{romanlist}

\ppart What propositional connective has the following truth table?  \hfill\examrule[0.7in]
\[\begin{array}{|c|c|c|}
\hline
P & Q & P\ \mathbf{?}\ Q\\
\hline
\true & \true & \false\\
\true & \false & \true\\
\false & \true & \true\\
\false & \false & \false\\
\hline
\end{array}\]

\begin{solution}
$\QXOR$.
\end{solution}

\ppart For each of the following formulas, state whether it is \textbf{V}alid
or \textbf{N}ot by writing \textbf{V} or \textbf{N}.

\begin{romanlist}
\item $\QNOT(\forall x \in A.\, P(x))\ \QIFF\ \exists x \in A.\, \QNOT(P(x))$.\hfill \examrule[0.5in]
\examspace[0.1in]

\item $P(x) \QAND \QNOT(Q(x))\  \QIFF\  \QNOT(P(x) \QIMPLIES Q(x))$.\hfill \examrule[0.5in]
\examspace[0.1in]

\item $\forall x \, \exists y. \, (P(x) \QOR Q(y))\ \QIFF\  \exists x \,
  \forall y \, \QNOT(P(x)) \QAND \QNOT(Q(y))$ . \hfill \examrule[0.5in]
\examspace[0.1in]

\item $\forall x \, \exists y. \ x^2 + y < 17$ where the domain of
  discourse is $\set{1, 2, 3, 4}$.\hfill \examrule[0.5in]
\end{romanlist}

\begin{solution}
\textbf{V,V, N, N}.

The third formula is not valid because it is false if $P$ and $Q$ are
true of all elements on the domain of discourse.

The fourth formula is false because if $x=4$, so $x^2 =16$, there is
no $y \in \Zintv{1}{4}$ such that $16 + y < 17$.
\end{solution}

\ppart The probability that a move in a certain game will succeed is
2/3.  Three such moves are done independently.  What is the
probability that:

\begin{romanlist}

\item Not all three moves will succeed?\hfill \examrule[0.7in]

\examspace[0.2in]

\begin{solution}
\[
1 - (2/3)^3 = 19/27 .
\]
\end{solution}

\item At most one move will succeed? \hfill \examrule[0.7in]

\examspace[0.2in]

\begin{solution}
\[
(1/3)^3 + 3 (2/3) (1/3)^2 = 7/27.
\]
\end{solution}

\end{romanlist}

\ppart Let $p(n) \eqdef n^2 - 16$.  What is the largest prime factor
of $p(51)$? \hfill \examrule[0.7in]

\begin{solution}
\textbf{47}.

$p(n) = (n - 4)(n + 4)$ So $p(51) = 47 \cdot 55 = 47 \cdot 11 \cdot 5 $.
\end{solution}

\ppart\iffalse
 Express the value of following sum in terms of a Harmonic number.\hfill \examrule[0.7in]
\[
2/1 + 2/2 + 2/3 + \dots + 2/53.
\]

\begin{solution}
\[
2 H_{53} .
\]
\end{solution}
\fi

What is the value of
\[
\sum_{n=0}^\infty \frac{2}{3^n} + \frac{2^n}{5^{n+1}} \,?
\] \hfill \examrule[0.7in]

\examspace[0.5in]

\begin{solution}
\[
\frac{2}{1-(1/3)} + \frac{1}{5} \cdot \frac{1}{1-(2/5)} = 3 + \frac{1}{3} = \frac{10}{3}.
\]
\end{solution}

\ppart Circle all the expressions below that equal the number of
size-5 subsets of the set $\set{1,2, 3, 4, 5, 6, 7}$.

\begin{center}
$7!/(2! 5!)$
\qquad $3 \cdot 7$
\qquad $5 \cdot 5!$
\qquad $7!/( 5! \cdot 5!)$
\qquad $7!/( 3! \cdot 4!)$
\end{center}

\begin{solution}
\[
7!/(2! 5!),~3 \cdot 7 .
\]
\end{solution}

\ppart A shooter shoots two bullets at a target.  The probability that
he will hit the target in the first shot is 0.70. If he hits the
target in his first shot, then the probability that he will hit the
target in the second shot is 0.80.  If he misses the target in his
first shot, then the probability that he will hit the target in the
second shot is $p$.  Given that the probability that he will hit the
target in the second shot is 0.62, what is $p$?

\hfill \examrule[0.5in]

\examspace

\iffalse
Circle the value of $p$ determined by this information.

\begin{center}
0.80 \qquad 0.62 \qquad 0.20 \qquad 0.56 \qquad Other
\end{center}
\fi

\begin{solution}
0.20 .

Let $T_1$ be the event that the shooter hits the target the first time, and let $T_2$ be the event that he hits the target the second time. We apply the rule of total probability.
\begin{align*}
  \pr{T_2}&= \prcond{T_2}{T_1}\pr{T_1} + \prcond{T_2}{\bar{T_1}}\pr{\bar{T_1}} \\
  0.62 &= 0.8\times 0.7 + p\times 0.3 
\end{align*}
We conclude that $p = (0.62-0.8\times 0.7)/0.3 = 0.20$.
\end{solution}

\ppart A jar contains an equal number of green and red balls, indistinguishable except for color.
George is blindfolded and asked to pick a ball from a jar and guess its color.  As soon as he picks,
his two friends, Anna and Sam, tell him what to answer.  Anna is color blind, so her answer is
random.  Sam can see the ball George draws and answers correctly 3/4 of the time.  If Sam and Anna
do not agree, George returns the ball to the jar without answering and then picks again.  If they do
agree, George will use their answer.

\begin{romanlist}

\item What is the probability that George will give the correct answer when he responds?\hfill
  \examrule[0.6in]

\examspace[0.5in]

\begin{solution}
\textbf{3/4}.

George responds when Anna and Sam agree, so the probability that
George is correct when he responds is the probability that both Anna
and Sam are correct, given that Anna and Sam agree, namely,
\begin{align*}
\lefteqn{\prcond{\text{both correct}}{\text{both agree}}}\\
& = \frac{ \pr{\text{Anna correct}} \pr{\text{Sam correct}}}{\pr{\text{both agree}}}\\
& = \frac{(1/2)(3/4)}{(1/2)(3/4) + (1/2)(1/4)} =\frac{3}{4}\, .
\end{align*}
\end{solution}

\item Suppose that there is only a single red ball in the jar.  Will
  your previous answer change?  Explain:

\begin{center}
\exambox{6.0in}{1.0in}{0in}
\end{center}
\begin{solution}
Doesn't matter.  The previous reasoning that led to the 3/4
probability did not depend on how many balls of either color were in
the jar.
\end{solution}

\end{romanlist}
\eparts

\end{problem}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
