\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{exit-diagnostic-short}
  \pcomment{by Judy Dori 12/15/13; revised ARM}
\end{pcomments}

\pkeywords{
  probability
  counting
  logic
  relation
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}

\bparts

\ppart Let $\power(A)$ be the power set of set $A$.

\begin{romanlist}
\item List the elements of $\power(\set{\emptyset, \set{\emptyset}})$.
\begin{center}
\exambox{3.0in}{0.4in}{0in}
\end{center}

\examspace[0.2in]

\begin{solution}
\[
\emptyset, \set{\emptyset}, \set{\set{\emptyset}}, \set{\emptyset, \set{\emptyset}}
\]

\end{solution}
\item How big is $\power(\set{1,2,\dots,10})$?\hfill \examrule[0.5in]

\examspace[0.2in]
\begin{solution}
\[
2^{10}
\]
\end{solution}

\item Circle the condition(s) under which the size of the union of
two finite sets, $A,B$, must equal to the sum of their sizes.

\begin{center}
Always \qquad $A \intersect B = \emptyset$ \qquad $A \subset B$ \qquad
$A \intersect B \neq \emptyset$ \qquad Never
\end{center}

\begin{solution}
\[
A \intersect B = \emptyset
\]\end{solution}

\end{romanlist}

\ppart In a certain place on Earth the probability that October will
be a rainy month is 1/5, and the probability that March will be a
rainy month is 1/4.  Assume these events are independent.
\begin{romanlist}

\item What is the probability that \textbf{both} October and March
  will \textbf{not} be rainy months?\hfill \examrule[0.5in]

\examspace[0.8in]

\begin{solution}
\[
\frac{3}{5} = \paren{1- \frac{1}{5}}\paren{1- \frac{1}{4}}
\]
\end{solution}

\item What is the probability that \textbf{only one} of these months will be
  rainy?\hfill \examrule[0.5in]

\examspace[0.8in]

\begin{solution}
\[
\frac{7}{20} =\frac{1}{5}\paren{1- \frac{1}{4}} + \paren{1 - \frac{1}{5}}\frac{1}{4} 
\]
\end{solution}

\item What is the probability that October \textbf{or} March will be rainy
  months?\hfill \examrule[0.5in]

\examspace[0.8in]

\begin{solution}
Inclusion-exclusion:
\[
\frac{2}{5} = \frac{1}{4} + \frac{1}{5} - \frac{1}{4}\cdot \frac{1}{5} =
\prob{\text{March}} + \prob{\text{October}} - \prob{\text{March and October}}
\]
\end{solution}

\end{romanlist}

\ppart Explain the following terms:

\begin{romanlist}
\item Injective (one-to-one) function.

\begin{center}
\exambox{6.0in}{1.0in}{0in}
\end{center}

\begin{solution}
In the notation of Section~\bref{leq1in_sec}:
\[
[\leq 1\ \text{out}] \QAND [\leq 1\ \text{in}]
\]

In words, each domain element is related (mapped) to at most one
codomain element, and each codomain element has at most one domain
element mapped to it.
\end{solution}


\item Proof by Contradiction.


\begin{center}
\exambox{6.0in}{1.0in}{0in}
\end{center}

\begin{solution}
To prove a proposition, assume it is false and reach a contradiction,
that is, deduce another proposition known to be false.
\end{solution}

\end{romanlist}

\ppart Circle each of the following formulas that are valid:

\begin{romanlist}
\item $\QNOT(P \QOR Q) \QIFF [\QNOT(P) \QAND \QNOT(Q)]$
%\item $\QNOT(P \QAND Q) \QIFF [\QNOT(P) \QOR \QNOT(Q)]$
\item $\forall x\, \exists y.\  P(x) \QOR \QNOT(P(y))$
\item $\exists x\, \forall y.\  P(x) \QOR \QNOT(P(y))$
%\item If $P$ is true, and $P$ implies $Q$, then $Q$ is true.
\item If $\QNOT(P)$ implies $\QNOT(Q)$, then $Q$ implies $P$.
\end{romanlist}

\begin{solution}
All are valid.
\end{solution}

\ppart What propositional connector has the following truth table? \hfill\examrule[0.5in]
\[\begin{array}{|c|c|c|}
\hline
P & Q & P\ \mathbf{?}\ Q\\
\hline
\true & \true & \true\\
\true & \false & \false\\
\false & \true & \false\\
\false & \false & \true\\
\hline
\end{array}\]

\begin{solution}
The $\QIFF$ connector; same as $\equiv$.
\end{solution}

\ppart An experiment consists of flipping three fair coins.

\begin{romanlist}

\item What is the probability that the number of heads at the second experiment\\
will be the same as that of the first one?\hfill \examrule[1.0in]

\examspace[1.0in]

\begin{solution}
\[
\frac{5}{16} = 2\paren{\paren{\frac{1}{2}}^3}^2 + 2\paren{3\paren{\frac{1}{2}}^3}^2
\]
\end{solution}

\item What is the total number of heads expected in 12 experiments? \hfill \examrule[1.0in]

\begin{solution}
\examspace[0.5in]
\[
18 = 12\cdot \frac{3}{2}.
\]
\end{solution}

\end{romanlist}

\ppart Let $p(n) \eqdef n^2+n+41$.  Circle the largest prime factor of
$p(41)$.

\begin{center}
37\qquad 41\qquad 43\qquad 1763\qquad Other
\end{center}

\begin{solution}
$43$
\end{solution}

\ppart Circle the value of $2 + 4 + 6 + 8 + \cdots + 2\cdot 2001 + 2 \cdot 2002$.

\begin{center}
2003001\qquad 2005003\qquad 4006002\qquad 4010006\qquad Other
\end{center}

\begin{solution}
\[
4010006 = \frac{(2+2\cdot 2002)(2002)}{2}
\]
\end{solution}

\ppart Circle all the expressions below that equal the number of
size-3 subsets of the set $\set{1,2, ..., 6}$.

\begin{center}
$5 \cdot 4$
\qquad $6 \cdot 5 \cdot 4$
\qquad $3 \cdot 6!$
\qquad $6!/( 3! \cdot 2!)$
\qquad $6!/( 3! \cdot 3!)$
\qquad $6!/( 3! \cdot 4!)$
\end{center}

\begin{solution}
\[
5 \cdot 4,\quad 6!/( 3! \cdot 3!) = \binom{6}{3}
\]
\end{solution}

\iffalse

\ppart Anna's toddler preschool class consists of three pairs of
non-identical twins.  Anna arbitrarily seats the six children at
two tables: table A, which can accommodate four children, and table B,
which can accommodate two children.

\begin{romanlist}
\item Circle the probability that exactly one pair of twins will sit at table A:

\begin{center}
1/5\qquad 4/15\qquad 1/3\qquad 4/5
\end{center}

\examspace[0.8in]

\begin{solution}
\[
\frac{7]{72}= \frac{1}{2}\cdot \frac{1}{6} + \frac{1}{2}\cdot \paren{\frac{1}{6}}^2
\]
\end{solution}

\item Circle the probability that at least one pair of twins will sit at table A:

\begin{center}
1/5\qquad 2/3\qquad 4/5\qquad 1
\end{center}

\examspace[0.8in]

\item Circle the probability that the twins Dan and Dina sit at different tables:

\begin{center}
1/5\qquad 4/15\qquad 8/15\qquad 4/5
\end{center}
\end{romanlist}

\examspace[0.8in]
\fi

\ppart First, we flip a fair coin. If heads comes up, then we roll one
die and take the result.  If tails comes up, then we roll two dice and
take the sum of the two results.  Circle the probability that this
process yields a 2.

\begin{center}
1/36\qquad 7/72\qquad 1/6\qquad 7/36\qquad Other
\end{center}

\begin{solution}

\[
\frac{7}{72} = \frac{1}{2}\cdot \frac{1}{6} + \frac{1}{2}\cdot
\paren{\frac{1}{6}}^2
\]
\end{solution}

\examspace

\ppart Tom and Jerry are two applicants for the same summer internship
at Google.  Since Tom and Jerry are equally qualified, Marla, who has
to hire one of them, decides to toss a coin exactly three times.  Tom
will get the internship if at least two heads are obtained, otherwise
Jerry will.  Marla's coin turns out to have a bias in favor of Tails.
Explain who has a higher probability of being accepted to the Google
Internship.

\begin{center}
\exambox{6.0in}{2.0in}{0in}
\end{center}

\begin{solution}
Jerry has the higher probability.  Since Tails are more likely than
Heads, two Tails and a Head is more likely than two Heads and a Tail,
and three Tails are more likely than three Heads.  Specifically,
$\prob{2 \text{ Tails}}= 3pq^2 > 3p^2q = \prob{2 \text{ Heads}}$ iff
$q>p$, where $p$ is the probability of Head and $q$ the probability of
a Tail.  Likewise $\prob{3 \text{ Tails}}> \prob{3 \text{ Heads}}$ iff
$q>p$.
\end{solution}
\eparts

\end{problem}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
