\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_AkraB3}
  \pcomment{formatted, but unedited from fall97 quiz1}
\end{pcomments}

\pkeywords{
  recurrence
  Akra-Bazzi
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Use the Akra-Bazzi formula to find $\Theta()$ asymptotic bounds for
the following recurrences.  For each recurrence $T(0) = 1$ and
$n\in\naturals$.

\bparts

\ppart
$T(n) = 2T\left(\floor{{n \over 4}}\right)+T\left(\floor{{n \over
3}}\right)+n$

\begin{solution}
$h_1(n) = \floor{{n \over 4}} - {n \over 4}$, $h_2(n) = \floor{{n
\over 3}} - {n \over 3}$.  $g(n) = n$.  $a_1 = 2, a_2 = 1$.  $b_1 = {1
\over 4}, b_2 = {1 \over 3}$.

Setting $p=1$ we get $2{1 \over 4} +{1\over 3} = {5 \over 6} < 1$. We
conclude that $p < 1$, therefore the recurrence is of the order of the
inhomogeneous part $g(n) = n$.  Therefore $T(n) = \Theta(n)$.
\end{solution}

\ppart

$T(n) = 4T\left(\floor{{n \over 2}+ \sqrt{n}}\right) + n^2$

\begin{solution}
$h(n) = \floor{{n \over 2}+ \sqrt{n}}-{n \over 2}$.  $a_1 = 4, b_1 =
{1 \over 2}, g(n) = n^2$

$4 \left( {1 \over 2} \right) ^2 = 1 \Rightarrow p = 2$

$$T(n) = \Theta(n^{2}(1+ \int_1^n \frac{u^2}{u^{2 + 1}} du)) =
\Theta(n^{2}(1+ \int_1^n u^{-1} du)) = \Theta(n^2 \log n).$$
\end{solution}

\ppart
A society of devil-worshipers meets every week in a catacomb to
initiate new members.  Members who have been in the society for two or
more weeks initiate four new members each and members who have been in
the society for only one week initiate one new member each.  On week 0
there is one devil-worshiper.  There are two devil-worshipers on week
1.

Write a recurrence relation for the number of members $D(n)$ in the
society on the $n$th week.

{\bf You do NOT need to solve the recurrence.  Be sure to include the
base cases.}

\begin{solution}
The number of devil-worshipers on week $n$ is the number of
devil-worshipers that were around 2 weeks before plus their new
recruits, so $5D(n-2)$, and the number of devil worshipers that were
around only last week, $D(n-2) - D(n-1)$, plus their new recruits.  So
the total number of devil-worshipers on the $n$th week is

$D(n) = 5D(n-2) + 2(D(n-1)-D(n-2)) = 3D(n-2) + 2D(n-1)$

The initial conditions are

$D(0) = 1$

$D(1) = 2$
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endinput


\iffalse
from fall98 final:

\begin{displaymath}
T(n) = 
\begin{cases}
1  & \text{if $n \leq 4$}\\
T\left(\floor{{n\over 4}}\right)+T\left(\floor{{n\over 3}}\right)+n^2
&  \text{if $n > 4$}\\
\end{cases}
\end{displaymath}


\answer{}

We will use the Akra-Bazzi formula.

We observe that $p < 1$.

$\Theta \left( n^p\left(1+ \int_1^n {x^2 \over x^{p+1}} dx \right)\right) = 
\Theta \left(  n^p n^{2-p} \right) = \Theta(n^2)$

\ppart{\bf [2 points]}

\begin{displaymath}
T(n) = 
\begin{cases}
1 & \text{if $n \leq 4$}\\
2T\left(\floor{{n\over 4}+ \sqrt{n}}\right)+\sqrt{n}
&  \text{if $n > 4$}\\
\end{cases}
\end{displaymath}

\answer{}

We use the Akra-Bazzi formula. $p$ is the solution to
$2 \left({1 \over 4}\right)^p =  1$, so 
$p = 1/2$

\begin{eqnarray*}
\Theta \left( n^{1/2} + n^{1/2} \int_1^n {\sqrt{x} \over x^{3/2}} \right)
 & =  & \Theta \left( n^{1/2} + n^{1/2} \int_1^n {1 \over x} \right)\\
& = & \Theta ( n^{1/2} +  n^{1/2}\ln n)\\
& = & \Theta (n^{1/2}\ln n) 
\end{eqnarray*}
\fi
