\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_binary_relations_on_01_S12}
  \pcomment{excerpted from FP_binary_relations_on_01}
  \pcomment{revised format from CP_binary_relations_on_01}
\end{pcomments}

\pkeywords{
  binary
  relations
  relational_properties
  partial_orders
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
How many binary relations are there on the set $\set{0,1}$? \hfill
\examrule[0.7in] \insolutions{16}

\vspace{0.1in}
\noindent How many are there that are:

transitive? \hfill \examrule[0.7in]  \insolutions{13}

\iffalse

\examspace[0.2in]
asymmetric? \hfill \examrule[0.7in] \insolutions{3}

\examspace[0.2in]
reflexive? \hfill \examrule[0.7in] \insolutions{4}

\examspace[0.2in]
irreflexive? \hfill \examrule[0.7in] \insolutions{4}
 
\examspace[0.2in]
strict partial orders? \hfill \examrule[0.7in] \insolutions{3}
\fi

\examspace[0.2in]
weak partial orders? \hfill \examrule[0.7in] \insolutions{3}

\examspace[0.2in]
equivalence relations?  \hfill \examrule[0.7in] \insolutions{2}

\begin{solution}
There are $2^4 = 16$ such relations, since in any such relation
  there are four possible arrows between $\set{0,1}$ and itself, each of
  which may or may not be there.

There are 3 \textbf{in}transitive relations, because the only way
transitivity can fail in a relation on two elements is when there is
an arrow in both directions between the elements, but one or the other
or both the elements are missing a \term{self-loop}, that is, an arrow
that starts and ends at the element.  So there are $13=16-3$
transitive relations.

There are 3 asymmetric relations.  Asymmetry implies no self-loops, and at
most one of the two possible arrows between 0 and 1.  So the only
3 possibilities are no arrows, arrow from 0 to 1, arrow from 1 to 0.

There are 4 reflexive relations, because two of the four possible arrows
(the self-loops) must be present, the remaining two arrows can be either
present or not present, which yields $2^2$ relations.  There are 4
irreflexive relations for the same reason.

There are 3 strict partial orders, because the 3 asymmetric relations are
all transitive.

There are 3 weak partial orders, because the 3 strict partial orders
remain distinct after adding self-loops to both elements.

There are only two equivalence relations: either 0 \& 1 are equivalent
or not.

\end{solution}


\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
