\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{new problem for final exam by YingZ, May 2014}
\end{pcomments}

\pkeywords{
  expectation
  conditional expectation
  linearity of expectation
  deviation
  sampling
  confidence
  tree diagram (4-step method)
  sample_space
  outcome
  probability space
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{problem}

%\begin{staffnotes}
% Point distribution goes here!
%\end{staffnotes}

Members of MIT's Computer Science and Artificial Intelligence
Laboratory (CSAIL) are known to thrive on caffeine.  Sixty percent are
devoted coffee drinkers, who drink coffee and water exclusively and
with equal probability in the lab.  Ten percent are devoted tea
drinkers, who drink tea 100\% of the time.  Another ten percent of lab
members are devoted soda drinkers, who drink soda 40\% of the time and
water or juice with equal probability the rest of the time.  The
remaining twenty percent drink all three types of caffeinated
beverages randomly 60\% of the time, and water and juice with equal
probability the rest of the time.

\bparts

\ppart As a 6.042 student, you go to CSAIL for office hours.  As soon
as you get off the elevator, you see two people each holding a cup in
front of the white board in discussion of something mod something.  What
is the probability that both of them are drinking coffee?

\hint Draw a tree diagram first.

\examspace[2in]

\begin{solution}
$0.3+0.04=0.34$
\end{solution}

\ppart Assuming the two filled their cups independently.  What is the
probability that their cups contain the same kind of beverage?

\begin{solution}
\end{solution}

\ppart Next, you bump into someone holding a cup of hot tea, due to
either your disorientation or his sleep deprivation, or both.  If he
is a devoted tea drinker, he will offer you a cup of fine tea as he
makes another cup for himself; else, you will have to find him an
unopened tea bag.  What is the probability that you get to enjoy a cup
of fine tea?

\begin{solution}
$\frac{0.1}{0.1+0.2 \cdot 0.6 \cdot 1/3} = \frac{0.1}{0.14} = \frac{5}{7}$
\end{solution}   

\ppart At office hours, as other students ask about a problem you
already solved, you imagine pulling all 500 cups that people in CSAIL
are drinking from and then counting the number of cups with just
water.  Let $W$ be a random variable for such number.  What is the
Chebyshev bound on the probability that $W \ge 100$?

\begin{solution}
\end{solution}

\ppart Not surprisingly, every student gains from office hours
differently.  Let $K_{i}$ be the random variable for the amount of
knowledge gained by the $i$-th student at the end of the office hours.
Assuming $n$ students are at the office hours, and $K_{i}$'s are
pairwise independent but all have mean $\mu$ and standard deviation
$\sigma$.  What is the upper bound on the probability that the average
gain among the $n$ students deviates from the expected gain of
individual students by at least $g$?  You could provide your answer in
an expression containing any or none of $g$, $\mu$, $\sigma$, $n$.

\begin{solution}
Let $\frac{S_{n}}{n}$ denotes the average gain among the $n$ students,
where $S_{n}::=\sum_{i=1}^{n} K_{i}$.  By Pairwise Independent
Sampling Theorem, $\pr{|\frac{S_{n}}{n} - \mu| \ge g} \le \frac{1}{n}
\cdot \frac{\sigma^{2}}{g^{2}}$.

\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
