\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_classlist_expectation}
  \pcomment{from: S01.practice final, prob3; modified from S07.ps13, prob3}
  \pcomment{adapted by Steven F09}
\end{pcomments}

\pkeywords{
  expectation
  random
  permutation
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}
There are $n$ students who are both taking Math for Computer Sience
and Introduction to Signal Processing this term.  To make it easier on
themselves, the Professors in charge of these classes have decided to
randomly permute their class lists and then assign students grades
based on their rank in the permutation ---just as many students have
suspected \smiley.  Assume all permutations are equally likely and
that the ranking in each class is independent of the other.

\bparts

\ppart
What is the expected number of students that have a higher rank
in Math for CS than in Signal Processing.?

\begin{solution}

Let $X_i$ be random variable whose value is 1 if student i has higher
rank in Math for CS than Signal Processing and 0 otherwise.  If
student $i$ has rank of $r$ for aMath for CS than the probability that
student $i$ has higher rank on Math for CS, given that the student has
rank $r$ in Math for CS, is $(r-1)/n$.  The probability student $i$
has rank $r$ in Math for CS is $1/n$.  Thus, the probability that
student $i$ has higher rank in Math for CS is
\[
\sum_{1\leq r \leq n}  \frac{1}{n} \frac{r-1}{n} = \frac{n-1}{2n}.
\]
By linearity of expectation, the expected number of students with a higher
rank on Math for CS is
\[
\expect{x} = \expect{x_1}+\expect{x_2}+...+\expect{x_n} = \frac{n-1}{2}.
\]

Another way to solve this problem is to observe that students with higher
rank in one class have lower rank in the other class, so by symmetry, the
expected number of higher- and lower-ranked students in either class is
the same.  The expected number of students with the same rank equals 1 by
linearity of expectation, since probability student ranked the same in
both schools is $1/n$.  These observations now yield same answer as above.

\end{solution}

\ppart What is the expected number of students that have a ranking at
least $k$ higher in Math for CS than in Signal Processing?

\begin{solution}
This part can be done in the same way as the previous one.  Let $X_i$
be the random variable whose value is 1 if student $i$ is ranked at
least $k$ higher in Math for CS than Signal Processing and 0
otherwise.  If student $i$ has rank of $r$ $(r>k)$ for Math for CS
than the probability that student $i$ has at least $k$ higher rank in
Math for CS given student ranked $r$ in Math for CS is $(r-k)/n$.  The
probability student $i$ has rank $r$ in Math for CS is $1/n$.  Thus,
the probability that student $i$ has at least $k$ higher rank in Math
for CS is
\[
\sum_{k+1\leq r\leq n} \frac{1}{n} \frac{r-k}{n} = \frac{(n-k)(n-k+1)}{2n^2}.
\]
By linearity of expectation, the expected number of students with a higher
rank on Math for CS is
\[
\expect{x}=\expect{x_1}+\expect{x_2}+...+\expect{x_n} = \frac{(n-k)(n-k+1)}{2n}.
\]

\end{solution}

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
