\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_divisibility_tests}
  \pcomment{by kazerani on 12/5/11 verbatim from F11 CP5F P1}
\end{pcomments}


\pkeywords{
series
modular arithmetic
number theory
divisibility
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

\textbf{This is currently identical to a current-term class problem.
As a short question, we could say that the alternating sum test tests
for divisibility by 11.  Divisibility by three prime numbers can
be checked via alternating sums involving chunks of three digits.
What are they (7,11,13) and why (prime divisors of 1001)?  Could
also ask about other divisibility tests.}\\\\

\begin{problemparts}

\problempart Why is a number written in decimal evenly divisible by 9 if and
only if the sum of its digits is a multiple of 9?  \hint $10 \equiv 1
\pmod{9}$.

\begin{solution}
Since $10 \equiv 1 \pmod{9}$, so is
\begin{equation}\label{10k}
10^k \equiv 1^k \equiv 1 \pmod{9}.
\end{equation}
Now a number in decimal has the form:
\[
d_k \cdot 10^k + d_{k-1} \cdot 10^{k-1} + \ldots + d_1 \cdot 10 + d_0.
\]

From~\eqref{10k}, we have
\begin{equation*}
d_k \cdot 10^k + d_{k-1} \cdot 10^{k-1} + \ldots + d_1 \cdot 10 + d_0
     \equiv  d_k + d_{k-1} + \ldots + d_1 + d_0 \pmod{9}
\end{equation*}

This shows something stronger than what we were asked to show, namely, it
shows that the remainder when the original number is divided by 9 is equal
to the remainder when the sum of the digits is divided by 9.  In
particular, if one is zero, then so is the other.
\end{solution}

%S04 ps7

\problempart Take a big number, such as 37273761261.  Sum the digits, where
every other one is negated:
\[
3 + (-7) + 2 + (-7) + 3 + (-7) + 6 + (-1) + 2 + (-6) + 1  =  -11
\]
Explain why the original number is a multiple of 11 if and only
if this sum is a multiple of 11.
%\hint $10 \equiv -1 \pmod{11}$.

\begin{solution}
A number in decimal has the form:
\[
d_k \cdot 10^k + d_{k-1} \cdot 10^{k-1} + \ldots + d_1 \cdot 10 + d_0
\]

Observing that $10 \equiv -1 \pmod{11}$, we know:
\begin{align*}
\lefteqn{d_k \cdot 10^k + d_{k-1} \cdot 10^{k-1} + \cdots + d_1 \cdot 10 + d_0} \\
    & \equiv d_k \cdot (-1)^k + d_{k-1} \cdot (-1)^{k-1} + \cdots  + d_1 \cdot (-1)^1 + d_0 \cdot (-1)^0 
\pmod{11} \\
& \equiv d_k - d_{k-1} + \cdots - d_1 + d_0  \pmod{11}
\end{align*}
assuming $k$ is even.  The case where $k$ is odd is the same with signs
reversed.

The procedure given in the problem computes $\pm$ this alternating sum of
digits, and hence yields a number divisible by 11 ($\equiv 0 \pmod {11}$)
iff the original number was divisible by 11.
\end{solution}

\end{problemparts}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
