\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_graph_100_dilworth}
  \pcomment{CH, spring 2014}
  \pcomment{forked from TP_subsequence_of_101}
\end{pcomments}

\pkeywords{chain
           anti-chain
           increasing
           decreasing
           partial order
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

Let $G$ be an arbitrary directed acyclic graph with $100$ vertices. 
%No other\emph{a priori} information is available regarding its edges.

\bparts

\ppart What is the size of the largest chain that $G$ could possibly have? Justify your reasoning.

\begin{solution}
Arrange the  vertices of $G$ in linear order. This gives a chain of size 100.
\end{solution}

\examspace[1in]

\ppart What is the size of the largest antichain that $G$ could possibly have? Justify your reasoning.

\begin{solution}
 The empty graph contains no edges and the vertices are fully
 unordered. In this case, the size of the largest antichain is also 100.
\end{solution}

\examspace[1in]

\ppart Argue that for any such $G$, either the maximum-size chain or
the maximum-size antichain has a size of at least 10.

\begin{solution}
Dilworth's Lemma states that any DAG with $n$ vertices must have either a chain of
size at least $t$, or an antichain of size at least $n/t$. Substitute $n=100$ and $t=10$ to obtain the result.
\end{solution}

\examspace[1in]

\ppart Give an example of $G$ where the maximum-size chain and the
maximum-size antichain \emph{both} have size 10. Justify your reasoning.

\begin{solution}

Partition the vertices $V$ into 10 blocks of 10 vertices
each. Linearly order vertices within each block, and leave different
blocks disconnected. 

Clearly, the length of the maximum-size chain is 10. We can construct
a set containing all vertices from any one block, forming a chain of
10 vertices. Any vertex from another block is incomparable with the
vertices we have chosen, and therefore no chain of bigger size is possible.

On the other hand, the length of the maximum-size
antichain is also 10. We can construct a set consisting of exactly one vertex from each of
the 10 blocks, forming an antichain of 10 vertices. Adding any other
vertex in this set would result in a pair of comparable vertices, and
therefore no antichain of bigger size is possible.

\end{solution}

\eparts

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
