\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_graph_colorable}
  \pcomment{Related to FP_graph_colorable}
  \pcomment{ARM 4/11/14}
\end{pcomments}

\pkeywords{
 graph theory
 colorable
 width
}

\begin{problem}
A simple graph, $G$, is said to have \emph{width} $w$ iff there is a
way to list all its vertices so that each vertex is adjacent to at
most $w$ vertices that appear earlier in the list.  All the graphs
mentioned below problem are assumed to be finite.

\bparts

\ppart Prove that every graph with width one is a forest.

\hint By induction, removing the last vertex.

\begin{solution}

\begin{proof}
By induction on the number of vertices, $n$.  The induction hypothesis
is

$P(n) \eqdef$ all $n$-vertex graphs $G$ with width one are forests.

\inductioncase{Base case}: ($n=1$).  A graph with one vertex is
acyclic and therefore is a forest.

\inductioncase{Induction step}.  Assume that that $P(n)$ is true for
some $n \geq 1$ and let $G$ be an $(n+1)$-vertex graph with width one.
We need only show that $G$ is acyclic.

The vertices of $G$ can be listed with each vertex adjacent to at most
one vertex earlier in the list.  Let $v$ be the last vertex in the
list.  Since \emph{all} the vertices adjacent to $v$ appear earlier in
the list, it follows that $\degr{v} \leq 1$.

Now $G-v$ obviously still has width one, so it is acyclic by Induction
Hypothesis.  But no degree one vertex is in a cycle, so adding $v$
back to $G-v$ will not create a cycle.  Hence $G$ is acyclic, as
claimed.

This proves $P(n+1)$ and completes the induction step.

\end{proof}

\end{solution}

\ppart Prove that every finite tree has width one.  Conclude that a
graph is a forest iff it has width one.

\begin{solution}
By induction on the number of vertices, $n$.  The induction hypothesis
is

$Q(n) \eqdef$ all $n$-vertex trees $T$ have width one.

\inductioncase{Base case}: ($n=1$).  Trivial.

\inductioncase{Induction step}.  Assume that that $Q(n)$ is true for
some $n \geq 1$ and let $T$ be an $(n+1)$-vertex tree.  We need only
show that $T$ has width one.

Let $v$ be a leaf of $T$.  Then $T-v$ has width one by Induction
Hypothesis, so its vertices can be listed with each vertex adjacent to
at most one vertex earlier in the list.  Since $v$ has degree one, we
can add it to the end of list of vertices of $T-v$ to obtain the
required list for $T$.  Hence $T$ has width one, as claimed.

This proves $Q(n+1)$ and completes the induction step.

Now 

\end{solution}


\end{problem}
\endinput
