\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_increasing_sequences_diag}
  \pcomment{variant of PS_increasing_sequences_bijection}
  \pcomment{zabel for s18 midterm3}
  \pcomment{ARM format 4/13/18, but will not include in book}
\end{pcomments}

\pkeywords{
  infinite sets
  sequences
  increasing
  diagonalization
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\Inc{\text{Inc}}

\begin{problem}
  \inhandout{As we did in PSET 4 Problem \#2, l}\inbook{L}et
  $\infstrings{\nngint}$ be the set of infinite sequences of
  nonegative integers, and say that a sequence
  $(a_0,a_1,a_2,\dots)\in\infstrings{\nngint}$ is \emph{strictly
    increasing} if $a_0 < a_1 < a_2 < \cdots$.  Define the subset
  $\Inc \subset \infstrings{\nngint}$ to be the set of strictly
  increasing sequences.

  Carefully prove that $\Inc$ is uncountable by a direct
  \textbf{diagonalization argument over \boldmath$\Inc$}.  \inhandout{Your proof
  should be self-contained, that is, written in a way that doesn't assume
  the reader is familiar with similar diagonalization proofs.}

\inhandout{  \medspace
  \noindent\emph{Note}: PSET 4 Problem \#2 provides a different way to
  prove that $\Inc$ is uncountable, by showing $\Inc \bij
  \infstrings{\nngint}$.  Please do \emph{not} cite or duplicate that
  bijection here, as it isn't helpful in constructing a proof that
  diagonalizes $\Inc$.
}

  \begin{staffnotes}
    PSET 4 Problem \#2 refers to
    \verb+PS_increasing_sequences_bijection+ in spring18.  Remove or
    replace this reference if using this problem in later semesters
    and/or in the book.
  \end{staffnotes}
  
\begin{solution}
  The proof is by contradiction, so assume that $\Inc$ is
  countable. $\Inc$ is not finite (for example, there are infinitely
  many arithmetic sequences in $\Inc$ of the form $(n,n+1,n+2,\ldots)$
  for $n\in\nngint$), so $\Inc$ must instead be countably
  infinite. Therefore there must exist a function $f:\nngint\to\Inc$
  that is a \emph{bijection}.

  Define a new sequence $s\in \infstrings{\nngint}$ as follows: choose
  $s[0] \eqdef 1 + f(0)[0]$, and thereafter, for each integer $k \ge
  1$, define
\[
s[k] \eqdef 1 + \max(s[k-1], f(k)[k]).
\]
  This defines a sequence $s\in \infstrings{\nngint}$.

  If $s = f(n)$ for some integer $n \ge 0$, then these sequences must have the same $n$th entry: $s[n] = f(n)[n]$. But by the definition of $s$, we have
\[
s[n] \ge 1 + f(n)[n] > f(n)[n]
\]
  (even in the $n = 0$ case), so $s[n] = f(n)[n]$ cannot happen. This means $s \ne f(n)$ for \emph{every} $n\in\nngint$, i.e., $s$ is not in the image of $f$.

  We also claim that $s\in\Inc$. The definition of $s$ shows that
\[
s[k] \ge 1 + s[k-1] > s[k-1]
\]
  for each $k\ge 1$, so $s$ is indeed \emph{strictly increasing}, that is, $s\in\Inc$.

  Because we have found a sequence $s\in\Inc$ that is not in the image
  of $f$, it follows that $f$ is not surjective and therefore cannot
  be a bijection.  This is a contradiction, as desired.
\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
