\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_induction_plus_2}
  \pcomment{from: F05.final}
  \pcomment{by adamc, 2012/01/13}
  \pcomment{wordsmithed by ARM 2/14/12}
\end{pcomments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

Suppose $P(n)$ is a predicate on natural numbers and suppose
\begin{equation}\label{Sk2} \notag
\forall k. \; P(k) \QIMPLIES P(k+2). \qquad (*)
\end{equation}

For $P$'s that satisfy~(*), some of the assertions below \emph{can}
(\textbf{C}) hold for some, but not all such $P$, other assertions
\emph{always} (\textbf{A}) hold no matter what the $P$ may be, and
some \emph{never} (\textbf{N}) hold or any such $P$.  Indicate which
case applies for each of the assertions \inbook{and briefly explain
  why}\inhandout{by \textbf{circling} the correct letter}.

\begin{problemparts}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart  $\forall n \ge 0.\; P(n)$ \inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
\textbf{C}.  The assertion means that $P$ is always true.  So $P(k+2)$
is always true, and therefore $P(k) \QIMPLIES P(k+2)$ is always true.
So this case is possible.  But~\ref{Sk2} also holds when $P$ is always
false, so the asertion does not always hold when \ref{Sk2} does.
\end{solution}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart  $\QNOT(P(0)) \QAND \forall n \ge 1.\; P(n)$ \inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
\textbf{C}.  This time $P$ is false at 0, but true everywhere else.
So $P(k) \QIMPLIES P(k+2)$ still always holds because $P(k+2)$ is
still always true.  So this assertion can hold, but not always,
since~\eqref{Sk2} can hold when $P(0)$ is true.
\end{solution}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart $\forall n \ge 0.\; \QNOT(P(n))$  \inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
\textbf{C.}  Now $P$ is always false.  So $P(k)
\QIMPLIES P(k+2)$ is always true because $P(k)$ is false.  So this case is
possible, but again does not always hold.
\end{solution}

\problempart $(\forall n \le 100.\; P(n)) \QAND (\forall n > 100.\; \QNOT(P(n)))$
\inhandout{\hfill \textbf{C\quad A\quad N}}


\begin{solution}
  \textbf{N}.  In this case, $P$ is true for $n$ up to 100 and false
  from 101 on.  So $P(99)$ is true, but $P(101)$ is false.  That means
  that $\QNOT[P(k) \QIMPLIES P(k+2)]$ for $k = 99$.  This case is
  impossible.
\end{solution}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart $(\forall n \le 100.\; \QNOT(P(n))) \QAND (\forall n > 100.\; P(n))$
\inhandout{\hfill \textbf{C\quad A\quad N}}


\begin{solution}
\textbf{C}.  In this case, $P$ is false for $n$ up to 100 and true
from 101 on.  So $P(k) \QIMPLIES P(k+2)$ for $k \le 100$ because
$P(k)$ is false, and $P(k) \QIMPLIES P(k+2)$ for $k \ge 99$ because
$P(k+2)$ is true.  So this case is possible, but again does not always
hold.
\end{solution}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart $P(0) \QIMPLIES \forall n.\, P(n+2)$
\inhandout{\hfill \textbf{C\quad A\quad N}}


\begin{solution}
\textbf{C}.  In this case, $P$ is false for $n$ up to 100 and true
from 101 on.  So $P(k) \QIMPLIES P(k+2)$ for $k \le 100$ because
$P(k)$ is false, and $P(k) \QIMPLIES P(k+2)$ for $k \ge 99$ because
$P(k+2)$ is true.  So this case is possible, but again does not always
hold.
\end{solution}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart $[\exists n\, P(2n)] \QIMPLIES \forall n. \ P(2n+2)$
\inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
\textbf{C}.  If $P(n)$ is always true this assertion holds.  So this
case is possible.  If $P(n)$ is true only for even $n$~\eqref{Sk2}
still holds, but $P(1+2)$ is false.  So this case does not always
hold.
\end{solution}

\problempart $P(1) \QIMPLIES \forall n.\, P(2n+1)$
\inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
  \textbf{A}.  This assertion says that if $P(1)$
  holds, then $P(n)$ holds for all odd $n$.  This case is always true.
\end{solution}

\problempart $[\exists n\, P(2n)] \QIMPLIES \forall n. \ P(2n+2)$
\inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
  \textbf{C}.  If $P(n)$ is always true, this assertion holds.  So
  this case is possible.  If $P(n)$ is true only for even $n$ greater
  than 4,~\eqref{Sk2} holds, but this assertion is false.  So this
  case does not always hold.
\end{solution}

\problempart $\exists n\, \exists m > n\, [P(2n) \QAND \QNOT(P(2m))]$
\inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
  \textbf{N}.  This assertion says that $P$ holds for some even
number ,$2n$, but not for some other larger even number, $2m$.  However,
if $P(2n)$ holds, we can apply~\eqref{Sk2} $n-m$ times to conclude $P(2m)$
also holds.  This case is impossible.
\end{solution}

\problempart $[\exists n.\, P(n)] \QIMPLIES \forall n.\, \exists m>
n.\, P(m)$
\inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
  \textbf{A}.  This assertion says that if $P$
holds for some $n$, then for every number, there is a larger number, $m$,
for which $P$ also holds.  Since~\eqref{Sk2} implies that if there is one
$n$ for which $P(n)$ holds, there are an infinite, increasing chain of
$k$'s for which $P(k)$ holds, this case is always true.
\end{solution}

\begin{staffnotes}
WAS COMMENTED OUT.
\end{staffnotes}

\problempart $\QNOT(P(0)) \QIMPLIES \forall n. \ \QNOT(P(2n))$
\inhandout{\hfill \textbf{C\quad A\quad N}}

\begin{solution}
\TBA{solution}
\end{solution}

\end{problemparts}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
