\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_martingale}
  \pcomment{very close to PS_fair_martingale}
  \pcomment{F07.rec14h}
  \pcomment{may be subsumed by sec:martingale}
\end{pcomments}

\pkeywords{
  deviation
  pairwise_independent
  paradox
  roulette
  mean_time_to_failure
  martingale
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
A gambler bets \$10 on ``red'' at a roulette table (the odds of red are
18/38, slightly less than even) to win \$10.  If he wins, he gets
back twice the amount of his bet, and he quits.  Otherwise, he doubles his
previous bet and continues.

For example, if he loses his first two bets but wins his third bet,
the total spent on his three bets is $10+20+40$ dollars, but he gets
back $2 \cdot 40$ dollars after his win on the third bet, for a net
profit of\$10.

\bparts

\ppart What is the expected number of bets the gambler makes before he
wins?

\begin{solution}
This is mean time to failure, with failure being a red number coming up.  So
the expected time (number of bets) is
\[
\frac{1}{18/38} = 2\ \frac{1}{9}.
\]
\end{solution}

\ppart What is his probability of winning?

\begin{solution}
He is certain to win, since $\pr{> k \text{ bets}} = (20/38)^{k}$ which
goes to zero as $k$ goes to infinity.  More fully,
\[
\pr{\text{win}} \geq \pr{\text{win in $\leq k$ bets}}
= 1 - \pr{> k \text{ bets}}
\]
and this last expression goes to 1 as $k$ goes to infinity.
\end{solution}

\ppart What is his expected final profit (amount won minus amount lost)?

\begin{solution}
His final profit is always \$10 whenever he finally wins, and he is
certain to win, so \$10 is also his expected final profit.
\end{solution}

\ppart It seems absurd that the gambler can expect to reap a positive
profit after just a few bets in a game that is biased against him, yet
probability theory leads to this conclusion.  The absurdity comes
about because it takes an infinite bankroll to follow the bet-doubling
strategy, so it's absurd to assume that bet-doubling is a feasible
strategy.

\iffalse It sounds plausible that, since his expected number of bets
is less than three, the expected size of his bet would be less than
the size of his third bet, that is, $10\cdot 2^2 = 40$ dollars.  But
this is a sloppy---and wrong---argument.  What it overlooks is that
later bets, though progressively less likely, grow tremendously, and
so contribute heavily to the expected bet size.\fi

Verify that bet-doubling requires an infinite bankroll by proving that
the expected size of the gambler's last bet is infinite.

\begin{solution}
Let $B$ be the size of his last bet in dollars.  Now if he wins his
$\$10$ final profit on the $k$th bet, then $B=10\cdot 2^{k-1}$, so
\[
\pr{B=10\cdot 2^{k-1}} = (20/38)^{k-1}(18/38)
\]
Therefore
\begin{align*}
\expect{B} & = \sum_{k \in \naturals^{+}} 10\cdot
      2^{k-1}\paren{\frac{20}{38}}^{k-1}(18/38)\\
  & = 10(18/38) \sum_{k \in \naturals} 2^k\paren{(20/38)^k}\\
  & > \sum_{k \in \naturals} \paren{1+ \frac{1}{19}}^k= \infty.
\end{align*}

\end{solution}

\eparts

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
