\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_MST_min_out.tex}
  \pcomment{by ARM 11/3/15}
\end{pcomments}

\pkeywords{
  tree
  spanning_tree
  MST
  weight
  minimum
  cycle
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
\bparts

\ppart
Give an example of a connected, weighted simple graph with four vertices
and distinct edge weights that contains a cycle and in which the edge with
the largest weight belongs to the minumum weight spanning tree (MST) of the graph.

\examspace[1.25in]

\begin{solution}
$V = \set{a,b,c,d}$, $E = \set{\edge{a}{b}, \edge{b}{c}, \edge{b}{d}, \edge{c}{d}}$, $W = \set{4,3,2,1}$.
\end{solution}

\ppart
Can you construct such an example with three vertices? Explain. %Why is it not possible to construct such an example with three vertices?

\examspace[1.5in]

\begin{solution}
No.
A cycle must contain at least three vertices. Since there are only three vertices, they must be all connected
in order to form a cycle. It follows that the edge with the largest weight must also be on that cycle.
But then it is not possible for the largest edge to be in MST, since taking the other two edges would
result in an MST with smaller weight.
\end{solution}

\ppart
Let $G$ be a connected, weighted simple graph.  Let $v \in \vertices{G}$
be an arbitrary vertex whose incident edges have distinct weights.
Suppose that edge $\edge{v}{w} \in \edges{G}$ has minimum weight among
the edges incident to $v$.  Prove that $\edge{v}{w}$ must be an edge
of some\footnote{Actually, the minimum weight edge incident to $v$
  must be in \emph{every} MST, but you need not prove this.}  minumum
weight spanning tree (MST) of $G$.

\hint Appeal to the gray edge construction \inhandout{in the text}
\inbook{of Lemma~\bref{lem:edgeextends}}.  Alternatively, consider a
path in an MST from between $v$ and $u$.

\examspace[3in]

\begin{solution}
Using the gray edge Lemma, color vertex $v$ black and all other
vertices white.  Then the only gray edges are the edges incident to
$v$, so $\edge{v}{w}$ will be the minimum weight gray edge, and
therefore will be in the MST built starting with this coloring.

Alternatively, if $\edge{v}{w}$ is not in some MST, $M$, then there is
a path in $M$ from $v$ to $u$, and this path must start with some edge
$g$ incident to $v$ that is heavier than $\edge{v}{w}$.  Then
$M-g+\edge{v}{w}$ would be an MST with smaller weight than $M$, a
contradiction.  So $\edge{v}{w}$ must be a member of every MST for $G$.
\end{solution}

\eparts

\end{problem}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endinput
