\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_marriage_modify_S16}
  \pcomment{variant of FP_marriage_modify}
  \pcomment{ARM 5/6/16}
\end{pcomments}

\pkeywords{
 stable_matching
 stable_marriage
 mating_ritual
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Four boys and four girls have the following preference rankings:

\begin{center}
\begin{tabular}{r|c} 
Alfred:    & Helen, Emily, Fiona, Grace    \\
Billy:     & Helen, Grace, Fiona, Emily    \\
Calvin:    & Grace, Helen, Emily, Fiona    \\
David:     & Helen, Grace, Emily, Fiona
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{r|c} 
Emily:     & Alfred, Calvin, David,  Billy    \\
Fiona:     & Calvin, Billy,  Alfred, David    \\
Grace:     & Calvin, Alfred, David,  Billy    \\
Helen:     & Calvin, Billy,  David,  Alfred
\end{tabular}
\end{center}

\bparts

\ppart\label{GraceCalvin} Explain why Grace and Calvin must be married
in every stable matching.

\examspace[0.6in]

\begin{solution}
Each is the other's first choice, so they would be a rogue couple if
they were not married to each other.
\end{solution}

\ppart Show that Helen amd Billy must must be married in every
stable matching.

\examspace[0.6in]

\begin{solution}
With Grace and Calvin matched, Helen and Billy are each other's
remaining first choice, so they must similarly be matched.
\end{solution}

\ppart Conclude that there is only one possible stable matching.

\begin{solution}
With the four above out of the way, Alfred and Emily are each other's
remaining first choice, so they must similarly be matched, forces Fiona to marry David.

\end{solution}
\iffalse

\ppart A machine is programmed to output three different matchings (not
necessarily stable) randomly.  What is the probability that at least one
of the three is stable?

\examspace[3in]
\begin{solution}
  There are $4!$ possible matchings, so there are $\binom{4!}{3}$ sets
  consisting of three matchings.  We know that the solution to
  part~\eqref{AlfredEmily} is unique, so there is only one possible
  sequence out of the $4!$ sequences that is stable.
 
  The total number of sets of three matchings that include one stable matching
  is therefore $\binom{4!-1}{2}$ (choose 2 unstable matching), so the
  probability is:
\[
\frac{\binom{4!-1}{2}}{\binom{4!}{3}} = \frac{1}{8}
\]
\end{solution}
\fi

\iffalse too much for a final

\ppart Suppose that Billy, being a noble and good friend of David, decides
to modify his preference ranking in order to help match David with Helen.
Find a new preference ranking for Billy such that any stable matching will
have David and Helen be a couple.  Explain why there is no other stable
matching with Helen being matched with someone else.

\begin{solution}
  Any ranking with Fiona in front of Helen will make it work.
\end{solution}
\fi

\eparts
\end{problem}

\endinput
