\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP-_partial_orders}
  \pcomment{based on CP_partial_orders}
  \pcomment{changed from to TP from CP, 3/16/10 by ARM}
\end{pcomments}


\pkeywords{
  partial_orders
  weak_partial_order  
  strict_partial_order
  transitive
  reflexive
  antisymmetric
  maximal
  minimal
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[quiz]{mcs}\begin{problem}

% FROM: Spring-07 MQ-2/28-1
%
% COMMENTS: Change to ask if they are weak/strict partial/total orders?
%           f=O(g), f=o(g), f~g, and subsets 

\TBA{Merge the parts; shorten}

\bparts

%not including the path total stuff
\iffalse
\ppart For each of the relations below, indicate whether it is
\emph{transitive} but not a partial order (\textbf{Tr}), a \emph{path-total
order} (\textbf{Tot}), a \emph{strict partial order} that is not path-total
(\textbf{S}), a \emph{weak partial order} that is not path-total (\textbf{W}),
or \emph{none} of the above (\textbf{N}).


\begin{solution}big Oh is \textbf{Tr}, little Oh is \textbf{S}, sim is
\textbf{Tr}, and subgraph is \textbf{W}.
\end{solution}

\fi


%\set{(a,b) | a<b} \text{ on the set of integers }\quad & \examrule{1in}\\
%% Added
%\set{(a,b) | a^2 \leq b^2} \text{ on the set of real numbers }\quad & \examrule{1in}\\



\ppart
For each of the binary
  relations below, select all of the properties terms that apply to it from the list given.

\TBAf{need to add: multiple choice reflexive, transitive, symmetric,
  anti-symmetric, strict partial, weak partial, maybe on a case by case basis}

\begin{itemize}

\item The relation $\mrel{R}$ between any two integers, $a$, $b$ such that $(a,b) \in \mrel{R}$
if and only if $a = b + 1$.

\begin{solution}
not reflexive, not symmetric, not transitive (hence not partial anything), but it is anti-symmetric vacuously.
\end{solution}

\item The superset relation, $\supseteq$ on the power set of the integers.

\begin{solution}
reflexive, transitive, not symmetric, but anti-symmetric. not strict partial. weak partial.
\end{solution}

\item The relation $\mrel{Q}$ between random variables with a finite expectation, where $(R_1, R_2) \in Q$ if and only if $\expect{R_1} < \expect{R_2}$

\begin{solution}
not reflexive, transitive, not symmetric, anti-symmetric (vacuously), strict partial.
\end{solution}

\item The \idx{empty relation} on the set of rationals.

\begin{solution}
reflexive, transitive, symmetric, anti-symmetric, strict-partial, weak-partial.
\end{solution}

\item the ``is a subgraph of'' relation on finite graphs.  (Note that every graph
is a considered a subgraph of itself.) \hfill \examrule[0.5in]

\begin{solution}
reflexive, transitive, not symmetric, anti-symmetric, weak partial.
\end{solution}

For the next three parts, let $f,g$ be nonnegative functions from the integers to the real numbers.

\item  the ``Big Oh'' relation, $f=O(g)$,  \hfill \examrule[0.5in]

\begin{solution}
reflexive, transitive, not symmetric, not anti-symmetric, weak partial.
\end{solution}

\item  the ``Little Oh'' relation, $f=o(g)$,  \hfill \examrule[0.5in]

\begin{solution}
not reflexive, transitive, not symmetric, not anti-symmetric, strong partial
\end{solution}

\item the ``asymptotically equal'' relation, $f \sim g$.  \hfill \examrule[0.5in]

\begin{solution}
reflexive, transitive, symmetric, weak partial.
\end{solution}

\end{itemize}
\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
