\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_random_graphsF99}
  \pcomment{textbf{Random Graphs.}}
  \pcomment{from F99.final}
\end{pcomments}

\pkeywords{
  probability
  random_variables
  expectation
  binomial
}

A simple graph with 10 vertices is constructed by randomly placing an edge
between every two vertices with probability $p$.  These random edge
placements are performed independently.
\begin{problem}
\bparts

\ppart What is the probability that a given vertex, $v$, of the graph
has degree two?

\exambox{1in}{0.6in}{}

\begin{solution}
\[
\binom{9}{2} \cdot p^2\cdot ({p-1})^7
\]
\end{solution}

\examspace[1in]

\ppart 
What is the expected number of nodes with degree two?

\exambox{2in}{1in}{}

\begin{solution}
10 times the previous answer by linearity of expectations.
\end{solution}

\examspace[2in]

\iffalse
\ppart Are $X_i$ and $X_j$ independent for $i\neq j$?  Prove
your answer.

\begin{solution}
No.

Prob($X_i=9$ and $X_j=0$)=0 $\neq Prob(X_i=9)\cdot Prob(X_j=0)$

\end{solution}

\fi

\ppart 
What is the expected number of cliques of size $k$, that is,
subgraphs isomorphic to the complete graph with $k$ vertices?  Express
your answer as a closed form formula involving binomial
coefficients.

\exambox{1.5in}{0.7in}{}

\begin{solution}
The strategy for computing the expected number of cliques uses linearity
of expectation.  Define an indicator random variable for every possible
$k$-clique.  Every $k$-subset of the $n$ nodes is a possible $k$-clique,
so there are $\binom{n}{k}$ such random variables.  The random variable
for a particular $k$-clique is 1 if the $k$-clique appears in the graph
and 0 if not, so its expected value is just the probability that a
particular $k$-clique occurs in the graph.  This happens if and only if
all $k$ nodes are connected by edges; there are $\binom{k}{2}$ such edges,
so the probability of a particular $k$-clique is $p^{\binom{k}{2}}$.
Summing the expected values of all the $k$-cliques gives the expected
number of $k$-cliques:
\[
\binom{n}{k} p^{\binom{k}{2}}.
\]

\end{solution}
\examspace[1in]

\ppart  The previous result impliies an interesting fact which you may
assume without proof: if $p=1/2$, then the expected number of cliques of
size $2 \log n$ is greater than one, where $n$ is the total number of
vertices in the graph.  (In the previous parts we had $n=10$; now $n$ is
arbitrary.)  Oddly, even though we can expect there to be such a clique,
no procedure is known to \emph{find} some clique of size $2\log n$ in the
graph---when there is one---which is much more efficient than exhaustively
searching through all possible subsets of vertices of size $2 \log n$ and
checking each subset to see if it is a clique.  Circle each of the
following expressions which correctly bounds the number of subsets such an
exhaustive search would examine:
\begin{enumerate}
\item $O(2^n)$
\item $O(n^2)$
\item $O(n^{\log n})$
\item $O(n^{2 \log n})$
%\item $\Theta(e^{\log^2 n})$
%\item $\Theta((\log n)^n)$
\end{enumerate}

\begin{solution}

(a),(d).

The number of such subsets is
\[
\binom{n}{2 \log n} = \frac{P(n, 2 \log n)}{(2 \log n)!} < P(n,2 \log n) <
n^{2 \log n},
\]
and 
\[
\frac{P(n, 2 \log n)}{(2 \log n)!} > \left(\frac{(n- \log n)}{2 \log n}\right)^{2 \log
n} > (n^{0.9})^{2 \log n} = n^{1.8 \log n} \neq O(n^{\log n}).
\]

\end{solution}

\eparts
\end{problem}

\endinput
