\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{FP_stable_marriage_S16}
  \pcomment{Lisa, Emanuele 3/10/13}
\end{pcomments}

\pkeywords{
  stable_marriage
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
We want a stable matching between $21$ boys and $21$ girls.

\bparts

\ppart Specify preference rankings for the boys and girls that allow
only one possible stable matching.  Briefly justify your answer.

\examspace[1.5in]

\begin{solution}
Let the boys all have different girls as first choices.  Let each
girl's first choice be the boy who ranks her first.  All other
rankings can be assigned arbitrarily.

Since the boys' first choices differ, it is possible to match them all
with their first choices.  Such a match is always stable.  Also, if
Bob and Alice rank each other first, then they would be a rogue couple
in any set of marriages where they were not married.  So this matching
of first choices to each other is the only possible stable matching.
\end{solution}

\ppart Mark each of the following predicates about the Stable Marriage
Ritual \textbf{P} if it is a \textbf{P}reserved Invariant, \textbf{N}
if it is \textbf{n}ot, and ``\textbf{?}'' if you are very unsure.
``Bob's list'' refers to the list of the women he has not crossed off.

\textbf{Scoring}: \emph{This part worth 10 pts.  Each correct answer
  gets 1 point. Each incorrect answer is -1 point.  Do \textbf{not}
  include any explanations.}

\begin{enumerate}[(i)]
\item Alice is not on Bob's current list.                      \hfill\examrule
\item No girl is on Bob's current list.                        \hfill\examrule
%\item Some boy is serenading Alice.                    \hfill\examrule
\item Bob is the only boy serenading Alice.            \hfill\examrule
\item Bob has fewer than 5 girls on his current list.          \hfill\examrule
\item Bob prefers Alice to his favorite remaining girl.\hfill\examrule 
\item Alice prefers her favorite current suitor to Bob.         \hfill\examrule
\item Bob is serenading his optimal spouse.            \hfill\examrule
\item Bob is serenading his pessimal spouse.           \hfill\examrule  
\item Alice's optimal spouse is serenading her.        \hfill\examrule
\item Alice's pessimal spouse is serenading her.       \hfill\examrule
\end{enumerate}

\begin{solution}
PPPPPPPPNP(?)
\end{solution}

\eparts
\end{problem}

\endinput

\iffalse
\begin{solution}
NPNNPPPPPP.
\end{solution}

\begin{center}
\begin{tabular}{lr}
Every girl appears on Bob's list.                & \examrule\\
No girl is on Bob's list.                        & \examrule\\
Every boy is serenading Alice.                   & \examrule\\
No boy is serenading Alice.                      & \examrule\\
Bob has fewer than 5 girls on his list.          & \examrule\\
Alice is Bob's optimal spouse.                   & \examrule\\
Alice's optimal spouse is serenading her.        & \examrule\\
Bob is serenading his current favorite girl.     & \examrule\\
Bob prefers Alice to his current favorite girl.  & \examrule\\
Alice prefers her current suitor to Bob.         & \examrule
\end{tabular}
\end{center}
\fi
