\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{TP_Stable_Marriage_Invariants}
  \pcomment{Converted from
  ./00Convert/probs/practice6/stable-marriage-invariants.scm by
  scmtotex and drewe on Thu 28 Jul 2011 01:05:16 PM EDT}
  \pcomment{similar to TP_mating_ritual_invariant}
  \pcomment{Converted to FP by AC 5/17/15}
\end{pcomments}

\begin{problem}
We are interested in invariants of the Mating
Ritual~\bref{stablemarriagesec} for finding stable marriages.  Let
Angelina and Jen be two of the girls, and Keith and Tom be two of the
boys.

Which of the following predicates are preserved invariants of the Mating Ritual
no matter what the preferences are among the boys and girls?

\bparts

\ppart Tom is not serenading Jen.

\begin{solution}
Not invariant: Tom might serenade Angelina, get rejected by
her, and then serenade Jen next.
\end{solution}

\ppart Tom's list of girls to serenade is empty.

\begin{solution}
Invariant: No girl will ever get added to Tom's list, so once his list
is empty, it stays empty.  If we have at least as many girls as boys,
Tom's list will never be empty, since the Ritual guarantees he will be
married in the end.  That is, this predicate is an invariant because
it is always false.  But it would still be an invariant even if we ran
the Mating Ritual with more boys than girls, in which case Tom's list
might become empty, but then would stay that way.
\end{solution}

\ppart
 All the boys have the same number of girls left uncrossed in their lists.

\begin{solution}
 Not invariant: Suppose all the boys like Angelina best.  Then she will
 reject all but her favorite, say it's Tom, on the first day. Now, on
 the second day, Tom's list will be one longer than all the other
 boys' lists.
\end{solution}

\ppart Jen is crossed off Keith's list and Keith prefers Jen to anyone
he is serenading.

\begin{solution}
Invariant: Keith crosses off girls in order of preference, so if Jen
is crossed off, Keith likes her better than anybody that's left.
\end{solution}

\eparts

\end{problem}

\endinput
