\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_4_way_dependent}
  \pcomment{new in spring 11}
\pcomment{some of the text has references to other spring11 problems}
\end{pcomments}

\pkeywords{
  independence
  pairwise_independence
  distribution
}

\begin{problem}

Suppose there are 4 desks in a classroom, laid out such that each is in a different 
corner of a square. Like in the flirting couples problem of last week, each seat a student
that is male or female with equal probability and independently of the gender of the 
other 3 students.  Also like in the class problem, two students located
either directly side by side or one in front of the other  will {\em flirt} if and only if they are
of different gender. Label each of the 4 eligible couples with the numbers 1,2,3,4.

Let $I_i$ be a random variable indicating whether the ith pair is flirting.

Prove or disprove each of the following statements.

\ppart $\{I_i\}_i=1^4$ are pairwise independent
\begin{solution}
true. can do by cases: 2 adjacent edges, 2 separate edges.
\end{solution}

\ppart $\{I_i\}_i=1^4$ are 3-wise independent
\begin{solution}
 true. Any three couples form a 'line' with 4 nodes. Assume WLOG we chose nodes 123. $\pr{I_1 and I_2 and I_3}$
 then we have  M F M F and  F M F M as unique posibilities, out of $2^4$, that is $1/8 = 1/2^3$
\end{solution}

\ppart $\{I_i\}_i=1^4$ are 4-wise independent
\begin{solution}
false. In this case we have the full square. Still only two pos. M F M F and F M F M,
out of $2^4$, so $1/8 != (1/2)^2$.
\end{solution}

\ppart What is the expected number of flirting couples?
\begin{solution}
2. linearity of expectation, or possibly from doing the next part.
\end{solution}

\ppart What is the probability distribution function for the random variable $N$, the number
of flirting pairs.
\begin{solution}
$p_N(0) = 2/16, p_N(1) = 0 p_N(2) = 12/16, p_N(3) = 0, p_N(4) = 2/16$
\end{solution}

\end{problem}

\endinput