\documentclass[problem]{mcs}

\begin{pcomments}
\pcomment{MQ_bipartite_matching_recitatations}
\pcomment{from S090.MQmar19}
\end{pcomments}

\pkeywords{
  graphs
  bipartite
  matching
  Halls_theorem
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

  Overworked and over-caffeinated, the Teaching Assistant's (TA's)
  decide to oust the lecturer and teach their own recitations.  They
  will run a recitation session at 4 different times in the same room.
  There are exactly 20 chairs to which a student can be assigned in
  each recitation.  Each student has provided the TA's with a list of
  the recitation sessions her schedule allows and no student's
  schedule conflicts with all 4 sessions.  The TA's must assign each
  student to a chair during recitation at a time she can attend, if
  such an assignment is possible.

  \bparts

  \ppart Describe how to model this situation as a matching problem.
  Be sure to specify what the vertices/edges should be and briefly
  describe how a matching would determine seat assignments for each
  student in a recitation that does not conflict with his schedule.
  (This is a \emph{modeling problem}; we aren't looking for a description
  of an algorithm to solve the problem.)
\begin{staffnotes}
4pts
\end{staffnotes}

 \begin{solution}
  There will be one vertex for each student, and 20 vertices for each
  recitation time slot (one for each chair). There is an edge between
  a student and all chair vertices for a particular recitation time
  slot if that time slot does not conflict with her schedule. A
  matching for the students assigns a student to a chair in a
  recitation that he can attend and assigns at most 20 students to
  any recitation.

  It is possible to assign the students to recitations iff a matching
  exists.
\end{solution}

  \ppart Suppose there are 65 students.  Given the information provided
  above, is a matching guaranteed?  Briefly explain.

\begin{staffnotes}
1pt. 
\end{staffnotes}

  \begin{solution}
   NO---TBA
  \end{solution}

  \eparts

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
