\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_cut_edges}
  \pcomment{from S00.ps4}
  \pcomment{revised by ARM 3/12/10}
\end{pcomments}

\pkeywords{
  graphs
  cut_edge
  simple_cycle
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
An edge of a connected simple graph is called a \term{cut-edge} when
removing the edge disconnects the graph.  It turns out that an edge in
a connected graph is a cut-edge iff no simple cycle traverses that
edge.  Prove one direction or the other of this ``iff'' assertion (you
choose which direction to prove).

\textbf{Warning}: Your proof may depend on the fact that simple cycles
are defined to be of length at least three.  If so, you should
indicate where your proof uses this fact.

\begin{solution}
  First, we prove that if $e$ is contained in a simple cycle, then $e$
  is not a cut-edge.  Since $G$ is connected, there exists a path $P$
  between every pair of distinct vertices.  We must show that this
  remains true after $e$ is removed.  We will do this by considering
  the two nodes $u$ and $v$ on each side of $e$, and showing that $u$
  and $v$ must still be connected by a different path $P_2$ when $e$
  is removed.  This shows that all original paths between any two
  vertices in the path can be changed to include the path $P_2$ where
  the edge $e$ used to be.  Let us write out the simple cycle that
  contains $e$ as $u,v,v_1,v_2,\dots,v_n,u$.  Because it is a simple
  cycle, all vertices are distinct (except for the double occurrence
  of $u$).  This means that the edge $e$ only occurs once in the
  cycle.  When $e$ is removed we can't travel along $(u,v)$, however
  we can use the path $u,v_n,\dots,v_2,v_1,v$.  This implies that all
  vertices are still connected in the graph, because we have a
  ``detour'' around $e$ when it is cut.

Next, we prove that if $e$ is not a cut-edge, then $e$ is contained in
a simple cycle.  We will make use of Theorem 1 on page 469 of Rosen,
which states that there is a {\em simple} path between every pair of
distinct vertices in a connected graph.  Since $e$ is not a cut-edge,
the graph obtained by removing edge $e$ is connected.  So by the
theorem, there exists a simple path $P$ connecting the endpoints $u$
and $v$ of edge $e$, even after $e$ is removed from the graph.  Let
$P=v,v_1,v_2,\dots,v_n,u$.  The path is simple so adjoining edge $e$
to the end of $P$ gives a simple cycle containing edge $e$.
Specifically, $v,v_1,v_2,\dots,v_n,u,v$.
\end{solution}
\end{problem}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
