\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_lehman_eq_modified_S15}
  \pcomment{Shen, Elizabeth S15}
\end{pcomments}

\pkeywords{
  WOP
  well_ordering
  Lehman
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here %%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Use the Well Ordering Principle to prove that the equation
$$3a^2 + 9b^2 = c^4$$ has no positive integer solutions. 
\noindent
\inhandout{\hint Consider the minimum value of $a$ amongst all possible solutions}
\end{problem}

\begin{solution}
The proof is by the Well Ordering Principle. Suppose that there exists a solution. Then, there must be a solution in which $a$ has the smallest possible value. We will show that in this solution, $a, b$, and $c$ must all be multiples of 3. However, then there exists a valid solution with even smaller $a$, namely $(\frac{a}{3}, \frac{b}{3}, \frac{c}{3})$! This is a contradiction, and therefore no solution exists.
\noindent
Now, we will show that $a, b$, and $c$ must be multiples of three. The left hand side is a multiple of 3, so $c$ must be a multiple of 3 and we can write $c = 3c'$. Substituting back gives 
$$3a^2+ 9b^2 = 81{c'}^4$$
$3a^2$ must be a multiple of 9, since the other terms are multiples of 9. This implies that $a^2$ is divisible by 3, and therefore that $a$ is a multiple of 3. Substituting $a = 3a'$ gives: 
$$27{a'}^2 + 9b^2 = 81{c'}^4$$
Similarly, we see that $b$ must be a multiple of 3, since the other terms are multiples of 27.  Therefore, $a$, $b$, and $c$ are all multiples of 3, as claimed. 
\end{proof}
\end{solution}
