\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_mating}
\end{pcomments}

\pkeywords{
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

In the Mating Ritual, suppose Tiger is one of the boys and Elin is one of the girls.
Which of the following are preserved invariants \textbf{in general}?
\begin{enumerate}
\item Tiger is Elin's only suitor.
\item Tiger's optimal wife is the most preferred girl on his list.
\item Elin has been crossed off by everyone above and including Tiger on her list.
\end{enumerate}
\examspace[2in]
\begin{solution}
The statements that are preserved invariants in general appear in boldface below:
\begin{enumerate}
\item Tiger is Elin's only suitor.  (This would certainly make Tiger Elin's favorite that day, but one or more of the boys who got rejected by another girl that day may visit Elin the following day.)
\item \textbf{Tiger's optimal wife is the most preferred girl on his list.}  (The Mating Ritual gives each boy his optimal wife.  Tiger must therefore ultimately marry his optimal wife, so once she becomes the most preferred girl 
on his list -- and thus the girl he is serenading -- she must remain the top girl on his list.)
\item \textbf{Elin's name has been crossed off by everyone whom she finds at least as desirable as Tiger.} (Note that this is a preserved invariant because it cannot ever be true.  Were it true on some day, Tiger would have 
crossed Elin's name off his list, so he would end up marrying a woman he finds less desirable.  She would also have removed from contention everyone she finds more desirable than Tiger.  So she would end up marrying someone she 
finds less desirable than Tiger.  Consequently, Tiger and Elin would constitute would a rogue couple.  Another way to think about it is this: If Elin's name was 
crossed off by Tiger and all the boys Elin prefers to him, then she must have a current favorite whom she prefers to all of them.  But Tiger and his betters in Elin's eyes are the top boys on her list: there is no one she 
prefers to them.) 
\end{enumerate}
\end{solution}

\end{problem}

\endinput
