\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_predicate_jections_S13}
  \pcomment{subset of MQ_predicate_jections}
  \pcomment{CH 2/20/14}
\end{pcomments}

\pkeywords{
  surjection
  bijection
  injection
  predicate
  formula
  total
  function
}

\begin{problem}

\bparts

\ppart Five assertions about a binary relation $R:A \to B$ are
bulleted below.  There are five predicate formulas that express some of
these assertions.  Write the number of each formula next to the
bulleted assertion it expresses, if there is one.  For example, you
should write ``2'' next to the last bulleted assertion, since formula
2.\ expresses that $R$ is the identity relation.  Write ``none'' next
to an assertion that no formula expresses.

Variables $a,a_1,\dots$ range over the domain $A$ and $b,b_1,\dots$
range over the codomain $B$.  More than one formula may express the
same bulleted assertion, and some formulas may not express any
bulleted assertion.

\begin{quote}

\begin{itemize}
\item $R$ is a surjection \hfill\examrule[1in]
\begin{solution}
3
\end{solution}

\item $R$ is an injection \hfill\examrule[1in]
\begin{solution}
5
\end{solution}

\item $R$ is a function \hfill\examrule[1in]
\begin{solution}
4
\end{solution}

\item $R$ is total \hfill\examrule[1in]
\begin{solution}
1
\end{solution}

\item $R$ is the identity relation.\hfill\examrule[1in]
\begin{solution}
2
\end{solution}

\end{itemize}

\end{quote}


\begin{enumerate}

\item $\forall a.\, \exists b.\ a \mrel{R} b$.   %total

%\item $\forall a.\ a \mrel{R} a$.     %none

\item $\forall a,b.\ a \mrel{R} b \QIFF a = b$.     %identity

%\item $\forall a,b.\ a \mrel{R} b \QOR a \neq b$.     %none

\item $\forall b.\, \exists a.\ a \mrel{R} b$.   %surj

%\item $\forall b_1,b_2,a.\ (a \mrel{R} b_1 \QAND a \mrel{R} b_2) \QIMPLIES b_1 = b_2$. %function
\item $\forall a_1,a_2,b_1,b_2.\ (a_1\mrel{R}b_1 \QAND a_2\mrel{R}b_2 \QAND b_1\neq b_2) \QIMPLIES a_1\neq a_2$.  %function

\item $\forall a_1,a_2,b.\ (a_1 \mrel{R} b \QAND a_2 \mrel{R} b) \QIMPLIES a_1 = a_2$.  %injection

%\item $\forall a_1,a_2,b_1,b_2.\ (a_1 \mrel{R} b_1 \QAND a_2 \mrel{R}
%  b_2 \QAND a_1 \neq a_2) \QIMPLIES b_1 \neq b_2$.  %injection

\end{enumerate}


\ppart Give an example of a relation $R$ that satisfies three of the
properties --- surjection, injection, total, and function (you indicate
which) --- but is \emph{not} a bijection.

\begin{solution}
Let
\[
A \eqdef \set{1,2}, B \eqdef \set{1}, \graph{R} \eqdef \set{(1,1)}.
\]

Then $R$ is not a bijection because it is not total, and indeed
$\card{A} \neq \card{B}$.  But $R$ is an injective, surjective
function.

There can be several other possible answers.
\end{solution}

\eparts

\end{problem}

\endinput
