\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_pascals_triangle.tex}
  \pcomment{from: S10}
\end{pcomments}

\pkeywords{
  combinatorial proof
  Pascal's Triangle
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

The following combinational identity is known as Pascal's Triangle: 
\begin{eqnarray*}
{{n-1}\choose{k-1}} + {{n-1}\choose k} = {n \choose k}
\end{eqnarray*}

\bparts

\ppart Give a \emph{combinatorial proof} that for Pascal's Triangle. 

\examspace[3in]
\begin{solution}

Both sides are equal to:
%
\[
\frac{n!}{k!\ (n-k)!}
\]


\end{solution}

\ppart Verify this combinational proof by giving an algebraic proof of this same fact. 

\examspace[3in]
\begin{solution}
This is identical to problem 16.12.1 in the notes, reproduced below:

Jay figures that $n$ people (including himself)
are competing for spots on a team and only $k$ will be selected.  As
part of maneuvering for a spot on the team, he needs to work out how
many different teams are possible.  There are two cases to consider:

\begin{itemize}

\item Jay \emph{is} selected for the team, and his $k - 1$
  teammates are selected from among the other $n - 1$ competitors.
  The number of different teams that can be formed in this way is:
%
\[
\binom{n-1}{k-1}
\]

\item Jay is \emph{not} selected for the team, and all $k$ team
members are selected from among the other $n - 1$ competitors.  The
number of teams that can be formed this way is:
%
\[
\binom{n - 1}{k}
\]

\end{itemize}

All teams of the first type contain Jay, and no team of the second
type does; therefore, the two sets of teams are disjoint.  Thus, by
the Sum Rule, the total number of teams is:
%
\[
\binom{n-1}{k-1} + \binom{n - 1}{k}
\]

Another way of thinking about it is that $n$ people (including
himself) are trying out for $k$ spots.  Thus, the number of ways to
select the team is simply:
%
\[
\binom{n}{k}
\]

Each method correctly counts the number of teams, so the 
 answers must be equal.  So we know:
%
\[
\binom{n-1}{k-1} + \binom{n - 1}{k} = \binom{n}{k}
\]
%
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
