\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_predicate_jections}
  \pcomment{ARM 3/17/13}
\end{pcomments}

\pkeywords{
  surjection
  bijection
  injection
  predicate
  formula
  total
  function
}

\begin{problem}

\bparts

\ppart Five assertions about a binary relation $R:A \to B$ are
bulleted below.  There are nine predicate formulas that express some
of these assertions.  Write the numbers of the formulas next to the
assertions they express.  For example, you should write ``4'' next to
the last assertion, since formula (4) expresses the assertion that $R$
is the identity relation.

Variables $a,a_1,\dots$ range over the domain $A$ and $b,b_1,\dots$
range over the codomain $B$.  More than one formula may express one
assertion.

\begin{quote}

\begin{itemize}
\item $R$ is a surjection \hfill\examrule[1in]

\item $R$ is an injection \hfill\examrule[1in]

\item $R$ is a function \hfill\examrule[1in]

\item $R$ is total \hfill\examrule[1in]

\item $R$ is the identity relation.\hfill\examrule[1in]

\end{itemize}

\end{quote}


\begin{enumerate}
\item $\forall b.\, \exists a.\ a \mrel{R} b$.   %surj

\item $\forall a.\, \exists b.\ a \mrel{R} b$.   %total

\item $\forall a.\ a \mrel{R} a$.     %none

\item $\forall a,b.\ a \mrel{R} b \QIFF a = b$.     %identity

\item $\forall a,b.\ a \mrel{R} b \QOR a \neq b$.     %none

\item $\forall b_1,b_2,a.\ (a \mrel{R} b_1 \QAND a \mrel{R} b_2) \QIMPLIES b_1 = b_2$. %function

\item $\forall a_1,a_2,b.\ (a_1 \mrel{R} b \QAND a_2 \mrel{R} b) \QIMPLIES a_1 = a_2$.  %injection

\item $\forall a_1,a_2,b_1,b_2.\ (a_1 \mrel{R} b_1 \QAND a_2 \mrel{R}
  b_2 \QAND a_1 \neq a_2) \QIMPLIES b_1 \neq b_2$.  %injection

\item $\forall a_1,a_2,b_1,b_2.\ (a_1\mrel{R}b_1 \QAND a_2\mrel{R}b_2 \QAND b_1\neq b_2) \QIMPLIES a_1\neq a_2$.  %function

\end{enumerate}


\ppart Give an example of a relation $R$ that satisfies three of the
properties surjection, injection, total, and function (you indicate
which) but is not a bijection.

\begin{solution}
Let
\[
A \eqdef \set{1,2}, B \eqdef \set{1}, \graph{R} \eqdef \set{(1,1)}.
\]

Then $R$ is not a bijection because it is not total, and indeed
$\card{A} \neq \card{B}$.  But $R$ is an injective, surjective
function.

There are lots more possible answers.
\end{solution}

\eparts

\end{problem}

\endinput
