
\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{CP_stable_matching_unique_morning}
  \pcomment{contributed by Rich Chan, revised and last part by ARM 10/18/09}
\end{pcomments}

\pkeywords{
 stable_matching
 stable_marriage
 invariant
 mating_ritual
 optimal
 pessimal
 }


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
  Consider a \idx{stable marriage} problem with 4 boys and 4 girls. Here
  are their preference rankings:

\begin{center}
\begin{tabular}{r|c} 
Alfred:    & Helen, Emily, Fiona, Grace    \\
Billy:     & Emily, Grace, Fiona, Helen    \\
Calvin:    & Grace, Helen, Emily, Fiona    \\
David:     & Helen, Grace, Emily, Fiona
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{r|c} 
Emily:     & Alfred, Calvin, David,  Billy    \\
Fiona:     & Calvin, Billy,  Alfred, David    \\
Grace:     & Calvin, Alfred, David,  Billy    \\
Helen:     & Calvin, Billy,  David,  Alfred
\end{tabular}
\end{center}

\bparts

\ppart
Exhibit a stable matching between the boys and girls.

\examspace[2in]
\begin{solution}
(Alfred, Emily), (Billy, Fiona), (Calvin, Grace), (David, Helen)
\end{solution}

\ppart Explain why this is the only stable matching possible.

\examspace[2in]
\begin{solution}
  The Mating Ritual reveals that the boy-optimal and boy-pessimal
  (obtained by switching the roles of boys and girls) matchings are the
  same, so the matching in the previous part is unique.
\end{solution}

\ppart Suppose that Harry is one of the boys and Alice is one of the girls
when a \idx{Mating Ritual} is performed.  Circle the properties
below that must be preserved invariants.
\renewcommand{\theenumi}{\roman{enumi}}
\renewcommand{\labelenumi}{(\theenumi)}

\begin{enumerate}

\item Harry is serenading Alice.

\item Harry is serenading a girl he likes better than Alice.

\item Harry is serenading a girl he likes less than Alice.

% \item Alice is the only girl on Harry's list.

% \item Three girls have boys serenading them.

\item Alice is crossed off Harry's list.

\item Alice likes her favorite better than Harry.

%\item Alice has no suitors.

\item Alice has at least one suitor.

%\item Alice has exactly one suitor.



\end{enumerate}

\begin{solution}
\mbox{ }

\begin{enumerate}

\item NOT preserved.  Harry will be rejected when Alice gets a better suitor.

\item NOT preserved; if the girl he is serenading rejects him, Alice
  might be next on Harry's list.

\item Preserved: the quality of the girl Harry serenades is is weakly decreasing.

\item Preserved.  No girls ever get added to Harry's list, so if Alice is
  off, she stays off.

\item Preserved: the quality of Alice's favorites is weakly increasing.

\item Preserved: the quality of Alice's favorites is weakly increasing.

\end{enumerate}
\end{solution}

\eparts

\end{problem}

\endinput
