\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{MQ_stable_matching_unique}
  \pcomment{contributed by Rich Chan, reivsed and last part by ARM 10/18/09}
\end{pcomments}

\pkeywords{
 stable_matching
 stable_marriage
 invariant
 mating_ritual
 optimal
 pessimal
 }


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
  Consider a \idx{stable marriage} problem with 4 boys and 4 girls. Here
  are their preference rankings:

\begin{center}
\begin{tabular}{r|c} 
Alfred:    & Grace, Helen, Emily, Fiona    \\
Billy:     & Emily, Grace, Fiona, Helen    \\
Calvin:    & Helen, Emily, Fiona, Grace    \\
David:     & Helen, Grace, Emily, Fiona
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{r|c} 
Emily:     & Calvin, Alfred, David,  Billy    \\
Fiona:     & Alfred, Billy,  Calvin, David    \\
Grace:     & Alfred, Calvin, David,  Billy    \\
Helen:     & Alfred, Billy,  David,  Calvin
\end{tabular}
\end{center}

\bparts

\ppart
Exhibit a stable matching between the boys and girls.

\examspace[2in]

\begin{solution}
(Alfred, Grace), (Billy, Fiona), (Calvin, Emily), (David, Helen)
\end{solution}

\ppart Explain why this is the only stable matching possible.

\examspace[2in]
\begin{solution}
  The Mating Ritual reveals that the boy-optimal and boy-pessimal
  (obtained by switching the roles of boys and girls) matchings are the
  same, so the matching in the previous part is unique.
\end{solution}

\ppart Suppose that Harry is one of the boys and Alice is one of the girls
when a \idx{Mating Ritual} is performed.  Circle the properties
below that must be preserved invariants.
\renewcommand{\theenumi}{\roman{enumi}}
\renewcommand{\labelenumi}{(\theenumi)}

\begin{enumerate}

\item Harry is serenading Alice.

% \item Alice is the only girl on Harry's list.

% \item Three girls have boys serenading them.

\item Alice is crossed off Harry's list.

\item Alice likes her favorite better than Harry.

%\item Alice has no suitors.

\item Alice has at least one suitor.

%\item Alice has exactly one suitor.

\item Harry is serenading a girl he likes better than Alice.

\item Harry is serenading a girl he likes less than Alice.


\end{enumerate}

\begin{solution}
\mbox{ }

\begin{enumerate}

\item NOT preserved.  Harry will be rejected when Alice gets a better suitor.

\item Preserved.  No girls ever get added to Harry's list, so if Alice is
  off, she stays off.

\item Preserved: the quality of Alice's favorites is weakly increasing.

\item Preserved: the quality of Alice's favorites is weakly increasing.

\item NOT preserved; if the girl he is serenading rejects him, Alice
  might be next on Harry's list.

\item Preserved: the quality of the girl Harry serenades is is weakly decreasing.

\end{enumerate}
\end{solution}

\eparts

\end{problem}

\endinput
