\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_13_card_hand}
  \pcomment{from: F03.ps5}
\end{pcomments}

\pkeywords{
  probability
  cards
  conditional_probability
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
A 52-card deck is thoroughly shuffled and you are dealt a hand of 13
cards.

\bparts

\ppart If you have one ace, what is the probability that you have a
second ace?

\begin{solution}
The number of possible hands is $N = \binom{52}{13}$, all of
which are equally likely.  Let $A_k$ be the event that you have
exactly $k$ aces.  Then:

\begin{eqnarray*}
\pr{A_k} & = & \frac{\binom{4}{k} \binom{48}{13 - k}}{N}
\end{eqnarray*}

This holds because there are $\binom{4}{k}$ ways to choose the $k$
aces and $\binom{48}{13 - k}$ ways to choose the remaining $13 - k$
non-aces.  Now we can compute the desired probability as follows:

\begin{eqnarray*}
\pr{\text{at least two aces} \mid \text{at least one ace}}
	& = &	\frac{\pr{\text{at least two aces} \cap
			\text{at least one ace}}}
		{\pr{\text{at least one ace}}} \\
	& = &	\frac{\pr{\text{at least two aces}}}
		{\pr{\text{at least one ace}}} \\
	& = &	\frac{\pr{A_2 \cup A_3 \cup A_4}}
		{\pr{A_1 \cup A_2 \cup A_3 \cup A_4}} \\
	& = &	\frac{\pr{A_2} + \pr{A_3} + \pr{A_4}}
		{\pr{A_1} + \pr{A_2} + \pr{A_3} + \pr{A_4}} \\
	& = &	\frac{	\binom{4}{2} \binom{48}{11} +
			\binom{4}{3} \binom{48}{10} +
			\binom{4}{4} \binom{48}{9}
			}	
		{	\binom{4}{1} \binom{48}{12} +
			\binom{4}{2} \binom{48}{11} +
			\binom{4}{3} \binom{48}{10} +
			\binom{4}{4} \binom{48}{9}
			} \\
	& \approx &	0.369
\end{eqnarray*}

The first step uses the definition of conditional probability.  In the
second step, note that if you have at least two aces, then you surely
have at least one.  In the third, we express the events of interest in
terms of the events $A_k$.  The fourth step uses the sum rule.  In the
fifth step, we substitute in values for each $\pr{A_k}$ and cancel
$N$'s.
\end{solution}

\ppart If you have the ace of spades, what is the probability that you
have a second ace?

\begin{solution}
The total number of hands with the ace of spaces is $M =
\binom{51}{12}$, since this is the number of ways to choose the
remaining 12 cards.  Let $B_k$ be the event that you have exactly $k$
\emph{additional} aces.  Then:

\begin{eqnarray*}
\pr{B_k} & = & \frac{\binom{3}{k} \binom{48}{12 - k}}{M}
\end{eqnarray*}

This equation follows from the fact that there are $\binom{3}{k}$ ways
to choose the $k$ additional aces and then $\binom{48}{12 - k}$ ways
to choose the remaining $12 - k$ non-aces.  Now we compute the desired
probability much as before:

\begin{eqnarray*}
\pr{\text{$A \spadesuit$ and at least one more ace} \mid
	\text{$A \spadesuit$}}
	& = &	\frac{\pr{\text{$A \spadesuit$ and at least one more ace}
			\cap \text{$A \spadesuit$}}}
		{\pr{\text{$A \spadesuit$}}} \\
	& = &	\frac{\pr{B_1 \cup B_2 \cup B_3}}
		{\pr{\text{$A \spadesuit$}}} \\
	& = &	\frac{\pr{B_1} + \pr{B_2} + \pr{B_3}}
		{\pr{\text{$A \spadesuit$}}} \\
	& = &	\frac{
			\binom{3}{1} \binom{48}{11} +
			\binom{3}{2} \binom{48}{10} +
			\binom{3}{3} \binom{48}{9}}
			{\binom{51}{12}} \\
	& \approx &	0.561
\end{eqnarray*}
\end{solution}

\eparts

Remarkably, the two answers are different.  This problem will test
your counting ability!

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
