\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_bipartite_matching_virtues}
  \pcomment{from: S06.ps4, commented out}
  \pcomment{minor edits by ARM, 10/16/09}
\end{pcomments}

\pkeywords{
  graphs
  bipartite
  matching
  Halls_theorem  
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
  Scholars through the ages have identified \emph{twenty} fundamental
  human virtues: honesty, generosity, loyalty, prudence, completing
  the weekly course reading-response, etc.  At the beginning of the
  term, every student in Math for Computer Science possessed exactly
  \emph{eight} of these virtues.  Furthermore, every student was
  unique; that is, no two students possessed exactly the same set of
  virtues.  The Math for Computer Science course staff must select
  \emph{one} additional virtue to impart to each student by the end of
  the term.  Prove that there is a way to select an additional virtue
  for each student so that every student is unique at the end of the
  term as well.

Suggestion: Use \idx{Hall's theorem}.  Try various interpretations for
the vertices on the left and right sides of your \idx{bipartite graph}.

\begin{solution}
Construct a bipartite graph $G$ as follows.  The vertices on
on the left are all students and the virtues on the right are all
subset of nine virtues.  There is an edge between a student and a set
of 9 virtues if the student already has 8 of those virtues.

Each vertex on the left has degree 12, since each student can learn one of
12 additional virtues.  The vertices on the right have degree at most 9,
since each set of 9 virtues has only 9 subsets of size 8.  So this
bipartite graph is \idx{degree-constrained}, and therefore, by
Lemma~\bref{degree-constrained_lemma}, there is a matching for the
students.  Thus, if each student is taught the additional virtue in the
set of 9 virtues with whom he or she is matched, then every student is
unique at the end of the term.
\end{solution}

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
