\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_cases_sum_of_squares}
  \pcomment{F16.ps1 and earlier even-year Fall terms}
  \pcomment{edited by ARM and fncv 2/5/17}
\end{pcomments}

\pkeywords{
  cases
  sum
  square
  even
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}

Suppose that 
\[
x^2 + y^2 = z^2,
\]
where $x, y, z$ are nonnegative integers.\\  

Consider the following sets of cases for parities of $x$ and $y$

\begin{enumerate}
    \item $x$ odd, $y$ even; $x$ even, $y$ odd; $x,y$ different parities
    \item $x,y$ both even; $x$ even, $y$ odd; $x$ odd, $y$ even; $x,y$ different parities; $x,y$ same parity
    \item $x,y$ different parities; $x,y$ both even; $x,y$ both odd
    \item $x,y$ both odd; $x,y$ same parity; $x,y$ both even
    \item $x,y$ both even; $x,y$ both odd; $x$ even, $y$ odd; $x$ odd, $y$ even; $x,y$ same parity; $x,y$ different parity 
\end{enumerate}

\noindent Pick one of the set of cases above and use it to prove the claim that $z$ is even if and only if $x$ and $y$ are even. (circle the set you picked)\\


\hint An odd number equals $2m+1$ for some integer $m$, so its square
equals $4(m^2 + m) + 1$.

\begin{solution}
Use set 3\\
Note that a number $n$ is even iff $n^2$ is even. In addition, the square of any even member must be a multiple of 4 \\

\inductioncase{Case 1}: if $x,y$ are different parities, $\Rightarrow$ one of $x,y$ is even, and the other one is odd.  Since the square of an even number is even, the square of an odd number is odd, and the sum of an odd number and a even number is odd $\Rightarrow x^2+y^2=z^2$ is odd $\Rightarrow$ $z$ cannot be even.

\inductioncase{Case 2}: if  $x,y$ are both even $\Rightarrow$ both $x^2$ and $y^2$ are multiples of 4. $\Rightarrow$ $x^2+y^2=z^2$ is also a multiple of 4 $\Rightarrow$ $z$ is even. 


\inductioncase{Case 3}: if $x,y$ are both odd, let $x=2m+1, y=2n+1$ for some nonnegative integers $m,n$. Then,  
\[
z^2 = (2m+1)^2 + (2n+1)^2
\].
Using the hint and rearranging this equation, we get
\[
z^2 = 4(m^2+m+n^2+n) + 2
\]
But the square of any even number must be a multiple of 4. Since $z^2$ leaves a remainder of 2 on division by 4, it is not a multiple of 4, so this is a contradiction. Therefore, no such $z$ can exist. That is, there exist no integers $x,y,z $ such that $x,y$ are odd, $z is even$, and $x^2+y^2=z^2$. So the claim is vacuously true.\\

We have showed that in all cases the claim is true.
\end{solution}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endinput