\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_counting_robot_paths}
  \pcomment{from: S07PS9}
\end{pcomments}

\pkeywords{
  counting
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
There is a robot that steps between integer positions in 3-dimensional
space.  Each step of the robot increments one coordinate and leaves the
other two unchanged.

\bparts

\ppart Let $m,n$ be nonnegative integers.  Describe a bijection
between the length $m+n$ binary strings with exactly $m$ ones and the
paths the robot can take to go from $(0,0,0)$ to $(m,n,0)$.  How many
such paths are there?

\begin{solution}
The bijection is defined by letting the $i$th bit of the binary
string corresponding to a path be 1 iff the $i$th step in the path
increments the 1st coordinate.  The number of such strings is
\[
\binom{m+n}{m}
\]
so this is also the number of paths.
\end{solution}

\ppart How many different paths are there from point $(0, 0, 0)$ to $(12,
24, 36)$?

\begin{solution}
There is a bijection between the set of all such paths and
the set of strings containing 12 X's, 24 Y's, and 36 Z's.  In
particular, we obtain a path by working through a string from left to
right.  An $X$ corresponds to a step that increments the first
coordinate, a $Y$ increments the second coordinate, and a $Z$
increments the third.  The number of such strings is:
%
\[
\frac{72!}{12!\ 24!\ 36!}
\]
%
Therefore, this is also the number of paths.
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
