\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_fib_induction}
  \pcomment{from: S03.ps5 (by wing)}
\end{pcomments}

\pkeywords{
  induction
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

  Fibonacci numbers are defined as follows:

  \begin{eqnarray*}
  F(0)	& = &	0 \\
  F(1)	& = &	1 \\
  F(n)	& = &	F(n - 1) + F(n - 2) \hspace{0.5in} \text{(for $n \geq 2$)}
  \end{eqnarray*}

  Thus, the first few Fibonacci numbers are 0, 1, 1, 2, 3, 5, 8, 13, and
  21.  Prove that for all $n \geq 1$, $F(n-1) \cdot F(n+1) - F(n)^2 =
  (-1)^n$.

\begin{solution}
  
  The proof is by induction on $n$.  Let $P(n)$ be the
  proposition that $F(n-1) \cdot F(n+1) - F(n)^2 = (-1)^n$.

  First, we must show that $P(1)$ holds.  In that case, we have:

  \begin{eqnarray*}
  F(1-1) \cdot F(1+1) - F(1)^2
  	& = &	0 \cdot 1 - 1^2 \\
  	& = &	-1^1
  \end{eqnarray*}

  Therefore, $P(1)$ is true.

  Next, we must show for all $n \geq 1$ that $P(n+1)$ is true, assuming that $P(n)$ is
  true.  We transform the left side of the equation
  in proposition $P(n+1)$ into the right side as follows.

  \begin{eqnarray*}
  && F((n+1) - 1) \cdot F((n+1) + 1) - F(n+1)^2 \\
  	& = &	F(n) \cdot F(n+2) - F(n+1)^2 \\
  	& = &	F(n) \cdot (F(n+1) + F(n)) - F(n+1) \cdot (F(n) + F(n-1)) \\
  	& = &	F(n) \cdot F(n+1) +
  		F(n) \cdot F(n) -
  		F(n+1) \cdot F(n) -
  		F(n+1) \cdot F(n-1) \\
  	& = &	F(n) \cdot F(n) -
  		F(n+1) \cdot F(n-1) \\
  	& = &	- (-1)^n \\
  	& = &	(-1)^{n+1}
  \end{eqnarray*}

  We begin by simplifying.  Then we use the definitions of $F(n+2)$ and
  $F(n+1)$.  In the next two steps, we multiply out and cancel terms.
  In the fifth step, we use the induction hypothesis.  In the last step,
  rewriting the expression gives the right side of the equation in
  $P(n+1)$.

  Since we have proved $P(1)$ and showed that $P(n)$ implies $P(n+1)$
  for all $n \geq 1$, the proposition $P(n)$ is true for all $n \geq 1$
  and the claim is proved.

\end{solution}

\end{problem}
