\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_logic_and_probability} 
  \pcomment{from S06, pset10}
\end{pcomments}

\pkeywords{
  probability
  random variable
  expectation
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}
Here are seven propositions:
%
\[
\begin{array}{rcrcr}
x_1 & \vee & x_3 & \vee & \neg x_7 \\
\neg x_5 & \vee & x_6 & \vee & x_7 \\
x_2 & \vee & \neg x_4 & \vee & x_6 \\
\neg x_4 & \vee & x_5 & \vee & \neg x_7 \\
x_3 & \vee & \neg x_5 & \vee & \neg x_8 \\
x_9 & \vee & \neg x_8 & \vee & x_2 \\
\neg x_3 & \vee & x_9 & \vee & x_4
\end{array}
\]
%
Note that:

\begin{enumerate}

\item Each proposition is the OR of three terms of the form $x_i$ or
the form $\neg x_i$.

\item The variables in the three terms in each proposition are all
different.

\end{enumerate}

\noindent Suppose that we assign true/false values to the variables
$x_1, \ldots, x_9$ independently and with equal probability.

\bparts

\ppart What is the expected number of true propositions?

\begin{solution}Each proposition is true unless all three of its terms are
false.  Thus, each proposition is true with probability:
%
\[
1 - \paren{\frac{1}{2}}^3 = \frac{7}{8}
\]

Let $T_i$ be an indicator for the event that the $i$-th proposition is
true.  Then the number of true propositions is $T_1 +
\ldots + T_7$ and the expected number is:
%
\begin{align*}
\expect{T_1 + \ldots + T_7}
    & = \expect{T_1} + \ldots + \expect{T_7} \\
    & = 7/8 + \ldots + 7/8 \\
    & = 49/8 = 6 \frac{1}{8}
\end{align*}
\end{solution}

\ppart Use your answer to prove that there exists an assignment to the
variables that makes \textit{all} of the propositions true.

\begin{solution}A random variable can not always be less than its
expectation, so there must be some assignment such that:
%
\[
T_1 + \ldots T_7 \geq 6 \frac{1}{8}
\]
%
This implies that $T_1 + \ldots + T_7 = 7$ for at least one outcome.
This outcome is an assignment to the variables such that all of the
propositions are true.
\end{solution}

\eparts

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
