\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{from: S09.ps1}
  \pcomment{from: S04 (Eric Lehman)}
  \pcomment{minor edits by ARM in S09}
  \pcomment{Fix hard reference to Week 1 notes.}
\end{pcomments}

\pkeywords{
  well-ordering
  WOP
  contradiction
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
  Week Notes 1 described a conjecture made by the great Mathematician Euler
  in 1769: there are no positive integer solutions to the equation
\[
a^4 + b^4 + c^4 =  d^4.
\]
The Notes also gave integer values for $a,b,c,d$ that do satisfy this
equation, discovered by Noam Elkies at Harvard more than two hundred years
later.  So Euler guessed wrong.

Now let's consider Lehman's\footnote{Suggested by Eric Lehman, a former
  6.042 Lecturer.} equation, similar to Euler's but with some
coefficients:
\begin{eqnarray}\label{wc}
8 a^4 + 4 b^4 + 2 c^4 & = & d^4
\end{eqnarray}

Prove that Lehman's equation~\eqref{wc} really does not have any positive
integer solutions.

\hint Consider the minimum value of $a$ among all possible solutions
to~\eqref{wc}.

\solution{Suppose that there exists a solution.  Then there must be a
  solution in which $a$ has the smallest possible value.  We will show
  that, in this solution, $a$, $b$, $c$, and $d$ must all be even.
  However, we can then obtain another solution over the positive integers
  with a smaller $a$ by dividing $a$, $b$, $c$, and $d$ in half.  This is
  a contradiction, and so no solution exists.

All that remains is to show that $a$, $b$, $c$, and $d$ must all be even.
The left side of Lehman's equation is even, so $d^4$ is even, so $d$ must
be even.  Substituting $d = 2 d'$ into Lehman's equation gives:

\begin{eqnarray}
8 a^4 + 4 b^4 + 2 c^4 & = & 16 d'^4
\end{eqnarray}

Now $2 c^4$ must be a multiple of 4, since every other term is a
multiple of 4.  This implies that $c^4$ is even and so $c$ is also
even.  Substituting $c = 2 c'$ into the previous equation gives:

\begin{eqnarray}
8 a^4 + 4 b^4 + 32 c'^4 & = & 16 d'^4
\end{eqnarray}

Arguing in the same way, $4 b^4$ must be a mutliple of 8, since every
other term is.  Therefore, $b^4$ is even and so $b$ is even.
Substituting $b = 2 b'$ gives:

\begin{eqnarray}
8 a^4 + 64 b'^4 + 32 c'^4 & = & 16 d'^4
\end{eqnarray}

Finally, $8 a^4$ must be a multiple of 16, $a^4$ must be even, and so
$a$ must also be even.  Therefore, $a$, $b$, $c$, and $d$ must all be
even, as claimed.}
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
