%PS_min_edge_DAG

\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{notes problem from S09 DAG notes}
\end{pcomments}

\pkeywords{
  DAG
  path relation
  covering edge
  digraph
  acyclic
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}%\label{mindag}

If $a$ and $b$ are distinct nodes of a digraph, then $a$ is said to
\emph{cover} $b$ if there is an edge from $a$ to $b$ and there is no other
path from $a$ to $b$.  If $a$ covers $b$, the edge from $a$ to $b$ is
called a covering edge.

\bparts

\ppart Show that if two DAG's have the same positive path relation, then
they have the same set of covering edges.

\ppart Describe two graphs with vertices $\set{1,2}$ which have the same
set of covering edges, but not the same positive path relation (\hint They
can't be DAG's.)

\begin{solution}
Let one graph have edges $\set{(1,2), (1,1)}$ and the other
$\set{(1,1),(2,1)}$.  They have the same set of covering edges, namely,
none.  The reason is that if a vertex is on a positive length cycle, then
no edge incident to it can be a covering edge.
\end{solution}

\ppart For any DAG, $D$, let $\widehat{D}$ be the subgraph of $D$
consisting of only the covering edges.  Show that if $D$ is finite, then
$D$ and $\widehat{D}$ have the same positive path relation, that is $D^+ =
\widehat{D}^+$.
 
\ppart Conclude that if $D$ is a finite DAG, then $\widehat{D}$ is the
unique DAG with the smallest number of edges among all digraphs with the same
positive path relation.

\ppart Show that the previous result is not true for the infinite DAG
corresponding to the total order on the rational numbers.

\begin{solution}
In the DAG for $<$ on the $\rationals$, there are no covering
edges, so $\widehat{<}$ has no edges.
\end{solution}

\eparts
\end{problem}

\endinput
