\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_monochromatic_rectangle_4}
  \pcomment{edit of PS_monochromatic_rectangle to add 4 colors}
  \pcomment{Author: Justin Venezuela (jven@mit.edu)}
  \pcomment{Generalized from USAMTS Problem 4/1/18 (usamts.org)}
  \pcomment{edited 11/3/11 by ARM}
  \pcomment{last part suggested by CH, added by ARM, 4/26/14}
\end{pcomments}

\pkeywords{
  counting
  pigeonhole
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

\bparts

\ppart Let $R$ be an $1025 \times 5$ rectangular matrix each of whose
entries are colored red, white, blue or black.  Explain why at least two of
the 1025 rows in $R$ must have identical color patterns.

\examspace[1.25in]

\begin{solution}
There are only $4^5=1024$ ways to color the points in a row of length 5,
and there are 1025 rows, so by the Pigeonhole Principle, some pair of rows
must have identical color patterns.
\end{solution}

\ppart\label{ppart:rect} Conclude that $R$ contains four points with the same color that
form the corners of a rectangle.

\examspace[1.5in]

\begin{solution}
Each row is length 5 and there are only 4 colors, so some color must
occur twice.  So the two rows with identical color patterns have the same
color occurring twice in corresponding positions, and these four
points form the corners of a rectangle.
\end{solution}

\ppart Now show that the conclusion from part (\ref{ppart:rect}) holds even when $R$ has only 41
rows.

\hint How many ways are there to pick two positions in a row of length
five and color them the same?

\begin{solution}
Define a match-pattern to be a set of two of the five positions in a
row, together with a color.  Say a row \emph{fits} the match pattern,
if in that row, the two positions specified by the pattern have the
color given by the pattern.

For example, if the pattern is $(\set{2, 4}, \text{white})$ then the
first two row colorings below fit but the last two do not.

\begin{tabular}{lccr}
(red,  &white, &blue,&  white, &black)\\
(blue, &white, &black,& white, &white)\\
(white,&blue,  &white,& red,   &white)\\
(white,&black, &white,& black, &red)
\end{tabular}

If two rows fit the same match pattern, then the positions where the
rows match will form the corners of a rectangle.  Now we will argue
that there must be two rows that fit the same match pattern.

The rows are length five and are four colors, so every row must fit
at least one match-pattern.  But there are $\binom{5}{2} \cdot 4 = 40$
match-patterns and 41 rows, so two rows are sure to fit the same match
pattern.
\end{solution}

\iffalse

\ppart Generalize the above argument to a coloring using the rainbow
colors Red, Orange, Yellow, Green, Blue, Indigo, Violet along with
White and Black.

\begin{solution}
There are now 9 colors, so a row of 10 points will have a color
repeated, say red, and a rectangle with $9^{10} +1$ rows will have two
rows with matching colors, so again, the first two matching colors in
each of the matched rows will form the corners of a rectangle.
\end{solution}
\fi

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
