\documentclass[problem]{mcs}

\begin{pcomments}
\pcomment{PS_putnam_tournament_problem}
\end{pcomments}

\pkeywords{
halls_theorem
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

Suppose $2m$ teams play in a round-robin tournament. Over a period of $2m - 1$ days, every team plays every other team exactly once. There are no ties. Show that for each day we can select a winning team, without selecting the same team twice.

\begin{solution}

We make use of Hall’s Theorem. We can model this problem as a bipartite graph $G$. Let $L(G)$ be the set of days. And $R(G)$ be the set of teams. An edge from $l \in L(G)$ to $r \in R(G)$ exists if team $r$ won on day $l$. If we can find a matching that covers $L(G)$, we will be done. 

Using Hall’s Theorem, we know that such a matching exists iff for every $ S \in L(G)$, $ |S| \le |N(S)|$. Assume that there exists some subset $S$ in $L(G)$ such that $|N(S)| < |S|$(which means $|N(S)| < 2m$). Let $R(G) \backslash N(S)$ be the set of teams that have not on won for any day $d \in S$. Consider a team $t \in R(G) \backslash N(S)$. For every day in $S$, t must have lost to a different opponent. This means that at least $S$ teams must have won some game. This would imply that $|N(S)| \ge |S|$. Contradiction. 

Thus, since we have satisfied Hall’s Theorem, it must always be possible to find a matching that covers $L(G)$. By extension, this means that for each day, we can select a winning team without selecting the same team twice. 

\end{solution}


\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
 
