\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{from: S08 ps9}
\end{pcomments}

\pkeywords{
  Pigeon Hole Principle
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% S08 ps9
\begin{problem}[4] {\bf Pigeon Huntin'}
\bparts

\ppart Show that any odd integer $x$ in the range $10^9 < x < 2 \cdot 10^9$ 
containing all ten digits $0,1,\ldots,9$ must have consecutive even digits. 
\hint What can you conclude about the parities of the first and last digit?

\begin{solution}In any such number, the first digit must be 1, the last must be 
odd and each digit from 0 to 9 must appear exactly once.  The only way for 
there to be no consecutive even digits is if every even digit has an odd 
digit on either side of it.  Treating the 4 positions between a sequence 
of the 5 odd digits beginning with 1 as the pigeonholes and the 5 even 
digits as pigeons, the pigeonhole principle implies that there must exist 
consecutive even digits.\end{solution}

\ppart Show that there are 2 vertices of equal degree in any finite 
undirected graph with $n \geq 2$ vertices. \hint Cases conditioned upon 
the existence of a degree zero vertex. 

\begin{solution}In the case where there are no degree 0 vertices, 
the domain for the degrees is the set $\{1,2,\ldots,n-1\}$ 
since each vertex can be adjacent to at most all of the 
remaining $n-1$ vertices.  If there is a degree zero vertex 
then it is not possible for any vertex to have degree $n-1$
and so the domain is $\{0,1,2,\ldots,n-2\}$.  In either case
we treat the $n$ vertices as the pigeons and the $n-1$ possible
degrees as the pigeonholes, the pigeonhole principle implies 
that there must exist a pair of vertices with the same degree.\end{solution}

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
