\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{from: F09.ps2}
  \pcomment{from: S03.ps5}
\end{pcomments}

\pkeywords{
  relations
  relational_properties
  mapping_lemma
  functions
  injections
  surjections
  bijections
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

Let $A$, $B$, and $C$ be finite sets, and let $f : B \to
C$ and $g : A \to B$ be functions.  Let $h$ be the function with
domain $A$ and range $C$ that maps $x \in A$ to $f(g(x))$.  Prove or
disprove the following claims:

\bparts
\ppart
If $h$ is surjective, then $f$ must be surjective.

\begin{solution}
True.

For all $x$ in $C$: Since $h$ is surjective, there exists $y$ in $A$ such that 
$h(y) = x$. Therefore, by definition of $h$, $f(g(y)) = x$, so $x$ is in
the image of $f$.

Therefore, all of $C$ is in the image of $f$, so $f$ is surjective.
\end{solution}

\ppart
If $h$ is surjective, then $g$ must be surjective.

\begin{solution}
False.

Suppose $A = C = \{1\}$ and $B = \{1, 2\}$. Let $f$ be such that $f(1) = f(2) = 1$, and
$g$ such that $g(1) = 1$. In this case $h$ is indeed surjective, as
$h(1) = 1$, but $g$ is not surjective as it doesn't map anything to 2.
\end{solution}

\ppart
If $h$ is injective, then $f$ must be injective.

\begin{solution}
False.

Taking the same example as in the previous case. $h$ is injective, because
only 1 maps to 1. However, $f$ is not injective as $f(1) = f(2).$
\end{solution}

\ppart
If $h$ is injective, then $g$ must be injective.

\begin{solution}
True.

For all $x$ and $y$: If $g(x) = g(y)$ then $h(x) = f(g(x)) = f(g(y)) = h(y)$ so $x = y$ because
$h$ is injective.

Therefore, $g$ is injective.
\end{solution}

\eparts
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
