%PS_stable_matching_non-optimal

\documentclass[problem]{mcs}

\begin{pcomments}

  \pcomment{from F07.ps4}
\end{pcomments}

\pkeywords{
 stable_matching
 optimal
 non-optimal
 Mating_ritual}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
\bparts

\ppart Describe a simple procedure to determine whether or not a stable
marriage problem has a unique solution, that is, only one possible stable
marriage assignment.

\begin{solution}

See if the Mating algorithm with Boys as suitors yields the same solution
as the algorithm with Girls as suitors.  These two marriage assignments
are boy-optimal and girl-optimal, respectively, so they agree iff there is
a unique solution.

\end{solution}

\ppart Construct a set of marriage preferences in which there is a stable
marriage assignment which differs from the boy-optimal matching that the
Mating Ritual would yield and also differs from the girl-optimal matching
the Mating Ritual would yield when the roles of boys and girls are
reversed.

Describe such a non-optimal stable matching, and briefly explain why it has
these properties.

\hint Divide the boys into two groups and also the girls into two groups.
Assign preferences so the first group of boys and girls prefer each other
to anyone in the second group.  Use a boy-optimal matching between the
first group, and a girl-optimal matching between the second group.

\begin{solution}

Consider a stable marriage problem with 4 boys and 4 girls and the
following partial information about their preferences:

\begin{center}
\begin{tabular}{|ccccc|}
\hline
B1: & G1 & G2 & --  & --\\
B2: & G2 & G1 & --  & --\\
B3: & --  & --  & G4 & G3\\
B4: & --  & --  & G3 & G4\\
\hline
G1: & B2 & B1 & --  & --\\
G2: & B1 & B2 & --  & --\\
G3: & -- & -- & B3 & B4\\
G4: & -- & -- & B4 & B3\\
\hline
\end{tabular}
\end{center}

The assignment
\[
 (B1, G1), (B2, G2), (B3, G3), (B4, G4)
\]
will be a stable matching whatever the unspecified preferences may be:

\begin{itemize}

\item $B1$ and $B2$ get their 1st choice, so won't be in a rogue couple.

\item $G1$ and $G2$ get their 2nd choices, so won't be in a rogue couple
  with the other two boys, $B3$ or $B4$.  So $G1$ and $G2$ won't be in any
  rogue couple, either.

\item $G3$ and $G4$ get their best remaining choices, so will never be in
  a rogue couple.

\item This leaves no possible rogue partners for $B3$ and $B4$.

\end{itemize}
So the marriages are sure to be stable.

Notice that giving $G1$ and $G2$ their first choices, that is, marrying
$(B1, G2)$ and $(B2, G1)$ would also be stable for the same reason.  But
with this switch, $B1$ does worse.  So the stable matching above is not
boy-pessimal.

Likewise, after marrying off the first two boys and girls, giving $B3$ and
$B4$ their best remaining choices, that is, marrying $(B3, G4), (B4, G3)$,
will also be stable.  But with this switch, $B3$ does better.  So the
stable matching above is not boy-optimal.

This implies that the stable matching above would not be produced by the
Mating Ritual.  
\end{solution}

\ppart \iffalse (Optional) \fi

Describe how to define a set of marriage preferences among $n$ boys and
$n$ girls which have more than $2^{n/4}$ stable assignments.

\begin{solution}
To find a set of preferences which have many stable matchings,
  arrange the boys into a list of $n/2$ pairs, and likewise arrange the
  girls into a list of $n/2$ pairs of girls.  Choose preferences so that
  the $k$th pair of boys ranks the $k$th pair of girls just below the
  previous pairs of girls, and likewise for the $k$th pair of girls.  This
  ensures that any matching in which the corresponding pairs of boys and
  girls are matched will be stable, as long as the submatches involving
  each pair separately is stable.  But the preferences of the $k$th pairs
  for each other can be chosen so that either of the two ways of matching
  them to each other will be stable in isolation.  
\end{solution}

\eparts

\end{problem}

\endinput
