\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_Sammy_the_shark} 
  \pcomment{from S07, pset7}
\end{pcomments}

\pkeywords{
  asymptotics
  geometric_sum
  geometric_series
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}
Sammy the Shark is a financial service provider who offers loans on
the following terms.

\begin{itemize}

\item Sammy loans a client $m$ dollars in the morning.  This puts the
client $m$ dollars in debt to Sammy.

\item Each evening, Sammy first charges a ``service fee'', which
increases the client's debt by $f$ dollars, and then Sammy charges
interest, which multiplies the debt by a factor of $p$.  For example,
if Sammy's interest rate were a modest 5\% per day, then $p$ would be
$1.05$.

\end{itemize}

\bparts

\ppart What is the client's debt at the end of the first day?

\examspace[0.5in]

\begin{solution}At the end of the first day, the client owes
  Sammy $(m + f) p = m p + f p$ dollars.
  \end{solution}

\ppart What is the client's debt at the end of the second day?

\examspace[0.5in]

\begin{solution}
\[
((m + f) p + f) p = m p^2 + f p^2 + f p
\]
\end{solution}

\ppart Write a formula for the client's debt after $d$ days and find
an equivalent closed form.

\begin{solution}
The client's debt after three days is
%
\[
(((m + f) p + f) p + f) p  =  m p^3 + f p^3 + f p^2 + f p.
\]
%
Generalizing from this pattern, the client owes
%
\[
m p^d + \sum_{k=1}^d fp^k
\]
%
dollars after $d$ days.  Applying the formula for a geometric sum gives:
%
\[
m p^d + f \cdot \paren{\frac{p^{d+1} - 1}{p - 1} - 1}
\]
\end{solution}

\eparts

\end{problem}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
