\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_strict_PO_irreflexive}
  \pcomment{from: S06.ps2 (by wing)}
\end{pcomments}

\pkeywords{
  partial_orders
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem} 
  
  Prove that a binary relation, $R$, on a set, $A$, is a strict partial
  order iff it is transitive and irreflexive.

  \begin{solution}
    For the left-to-right direction of the ``iff'': Suppose $R$
    is a strict partial order; namely, $R$ is transitive and asymmetric. We
    need to prove that it is transitive and irreflexive. That $R$ is
    transitive we know already. To prove that it is irreflexive, we need
    to prove that
    \[
    \text{for every $a\in A$:\quad  $aRa$ is false.}
    \]
    So, pick any $a\in A$. Towards a contradiction, assume that $aRa$ is
    true.  Then, by the asymmetry of $R$, we know that $\neg(aRa)$ is also
    true. So, we have both $aRa$ and $\neg(aRa)$. This is a
    contradiction. Hence, $R$ must be irreflexive.

    For the right-to-left direction of the ``iff'': 
    Suppose $R$ is transitive and irreflexive.  To show that $R$ is a
    strict partial order, we need to show that it is transitive and
    asymmetric.  That $R$ is transitive we know already. To prove that it
    is asymmetric, we need to prove that 
    \[
    \text{for every $a,b\in A$:\quad   $aRb\implies\neg(bRa)$.}
    \]
    So, pick any $a,b\in A$. We need to prove that the implication
    $aRb\implies\neg(bRa)$ holds. Towards a contradiction, assume the
    implication is false. In other words, assume that both $aRb$ and $bRa$
    are true. Then we would also have $aRa$ (by transitivity). But this
    would contradict the fact that $R$ is irreflexive. Therefore, $R$ must
    be asymmetric.
  \end{solution}
  
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
