\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_stable_marriage_better_worse_count}
  \pcomment{simpler alternative to PS_stable_marriage_better_worse}
  \pcomment{by ARM 5/6/15}
  \pcomment{Lemma suggested by Leonid Levin; proof by ARM}

\end{pcomments}

\pkeywords{
 stable_matching
 optimal_spouse
 pessimal
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Suppose there are two stable sets of marriages\iffalse , a first set
and a second set\fi.  So each man has a first wife and a second wife
\iffalse(they may be the same)\fi, and likewise each woman has a first
husband and a second husband.

Someone in a given marriage is a \emph{winner} when they prefer their
current spouse to their other spouse, and they are a \emph{loser} when
they prefer their other spouse to their current spouse.  (If someone
has the same spouse in both of their marriages, then they will be
neither a winner nor a loser.)

We will show that
\begin{equation}\tag{WL}
\text{In each of the marriages, someone is a winner iff their spouse is a loser.}
\end{equation}

\bparts

\ppart\label{2l2w} The left to right direction of~(WL) is
equivalent to the assertion that married partners cannot both be
winners.  Explain why this follows directly from the definition of
rogue couple.

\begin{solution}
If both partners in a marriage were winners, then they would, by
definition, be a rogue couple for the other set marriages,
contradicting stability.
\end{solution}

\medskip
\eparts

The right to left direction of~(WL) is equivalent to the
assertion that a married couple cannot both be losers.  This will
follow by comparing the number of winners and losers among the
marriages.

\bparts

\ppart Explain why the number of winners must equal the number of
losers among the two sets of marriages.

\begin{solution}
If a person has the same spouse in both marriages, then they add
nothing to the numbers of winners or losers.  Otherwise, a person is a
loser in one marriage and a winner in the other.  So the overall
number of winners and losers among all the marriages must be equal.
\end{solution}

\ppart Complete the proof of~(WL) by showing that if some
married couple were both losers, then there must be another couple who
were both winners.

\begin{solution}
Since the number of winners and losers is the same, if some marriage
contributed two losers---and hence no winners---some other marriage
would have to contribute two winners, contradicting part~\eqref{2l2w}.
\end{solution}

\ppart Conclude that in a stable set of marriages, someone's spouse is
optimal iff they are pessimal for their spouse.

\begin{solution}
A man is not married to his optimal wife in a stable set of marriages
iff he has a better wife in some second set of stable marriages.
By~(WL), this holds iff his first wife has a worse second husband,
which means he is not his first wife's pessimal husband.  By symmetry,
this also holds when genders are reversed.

Note that Lemma~\bref{optispess} already showed that everyone was
pessimal for their optimal spouse, but now we have the converse that
everyone is optimal for their pessimal spouse.
\end{solution}

\eparts
\end{problem}

\endinput
