\documentclass[problem]{mcs}

\begin{pcomments}
\pcomment{PS_tournament_matching_problem}
\pcomment{soln by Prekshah Naik & ARM 4/9/17}
\pcomment{edited ARM 11/6/17}
\end{pcomments}

\pkeywords{
halls_theorem
bottleneck
tournament
bipartite
matching
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Suppose $n$ teams play in a round-robin tournament.  Each day, each
team will play a match with another team.  Over a period of $n - 1$
days, every team plays every other team exactly once.  There are no
ties.  Show that for each day we can select a winning team, without
selecting the same team twice.\footnote{Based on 2012 Putnam Exam
  problem B3.}

\hint Define a bipartite graph $G$ with $\leftbi{G}$ the set of days
and $\rightbi{G}$ the set of teams.  For any set $D$ of days, there
may, or may not, have been a team that lost on all of those days.

\begin{solution}
\iffalse We make use of Hall's Theorem.  We can model this problem as
a bipartite graph $G$.  Let $\leftbi{G}$ be the set of $n-1$ days, and
let $\rightbi{G}$ be the set of $n$ teams.\fi  If team $t$ wins its
match on day $d$, let there be an edge from $d$ to $t$.  Selecting a
different winner for each day of the tournament now simply means
finding a bipartite matching that covers $\leftbi{G}$.

For any set of days $D \subseteq \leftbi{G}$, let $\neighbors{D}$ be
the set of teams that won on at least one of those days.  Using Hall's
Theorem, we know the desired matching will exist iff the number of
winners during each of those days is at least as large as the number
of days, that is,
\begin{equation}\tag{DND}
\card{D} \leq \card{\neighbors{D}}.
\end{equation}

We now prove~(DND) by cases:

\begin{proof}
If every team won on one or more of the days in $D$, then
$\neighbors{D}$ is the set of all $n$ teams, so 
\[
\card{\neighbors{D}} = n > n-1 \geq \card{D}.
\]
Therefore,~(DND) holds.

On the other hand, suppose there is a team $t$ that did not win on any
of the days in $D$.  By the rules of the tournament, a different team
must have won against $t$ on each day in $D$.  Let $W_t \subseteq
\neighbors{D}$ be this set of teams.  So we have
\[
\card{D}  = \card{W_t}  \leq \card{\neighbors{D}},
\]
and (DND) also holds.

So~(DND) holds in any case, and we conclude by Hall's Theorem that
the desired matching exists.
\end{proof}

Note that $n$ must be even and each day there must be $n/2$ winners
and $n/2$ losers.  But the graph $G$ in general will not be degree
constrained (Definition~\bref{degree-constrained_def}), and these
numbers are not needed in the proof.
\end{solution}

\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\endinput
