\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{from: F09.ps3}
  \pcomment{from: S03.ps4}
\end{pcomments}

\pkeywords{
  induction
  strong_induction
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}

  A group of $n \geq 1$ people can be divided into disjoint teams, each
  containing either 4 or 7 people.  What are all the possible values of
  $n$?  Use induction to prove that your answer is correct.

\begin{solution}

Based on the following observations
\begin{eqnarray*}
\lefteqn{4 = 4} \\
\lefteqn{7 = 7} \\
\lefteqn{8 = 4 + 4} \\
\lefteqn{11 = 4 + 7} \\
\lefteqn{12 = 4 + 4 + 4} \\
\lefteqn{14 = 7 + 7} \\
\lefteqn{15 = 4 + 4 + 7} \\
\lefteqn{16 = 4 + 4 + 4 + 4} \\
\lefteqn{18 = 4 + 7 + 7} \\
\lefteqn{19 = 4 + 4 + 4 + 7} \\
\lefteqn{20 = 4 + 4 + 4 + 4 + 4} \\
\lefteqn{21 = 7 + 7 + 7} \\
\end{eqnarray*}
we claim and will prove the statement ``A group of $n \geq 18$ people can be divided into disjoing teams, each containing either 4 or 7 people''.

\begin{proof}
  The proof is by strong induction on $n$. Let $P(n)$ be the proposition
  that a group of $n \geq 18$ people can be divided into disjoing teams,
  each containing either 4 or 7 people.

Base cases:
$P(18)$ is true since $18 = 4 + 7 + 7$. 
$P(19)$ is true since $19 = 4 + 4 + 4 + 7$. 
$P(20)$ is true since $20 = 4 + 4 + 4 + 4 + 4$.
$P(21)$ is true since $21 = 7 + 7 + 7$.

Inductive steps:
For all $n \geq 21$, we assume that $P(18)$, $P(19)$, $\ldots$, $P(n)$ are true in order to prove that $P(n+1)$ is true.
Since $n+1 = (n-3) + 4$ and $n-3 \geq 18$, $n+1$ people can be first divided into 2 disjoint teams with 4 people and $n-3$ people. The team with $n-3$ people can be further divided into disjoint teams with 4 or 7 people by the induction hypothesis. That is, $P(n+1)$ is true.

Therefore, $P(n)$ holds for all $n \geq 18$ by strong induction.

All the possible values of $n$ are 4, 7, 8, 11, 12, 14, 15, 16, and $\geq 18$.
\end {proof}
\end{solution}

\end{problem}
