\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{TP_6042_TEAL_Table}
    \pcomment{Converted from round-table.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 04:04:28 PM EDT}
    \pcomment{revised by drewe 2 Aug 2011}
\end{pcomments}

\begin{problem}

%% type: short-answer
%% title: 6.042 TEAL Table

8 students---Anna, Brian,
Caine,\dots--are to be seated around a circular table in a circular room.  Two seatings
are regarded as defining the same \emph{arrangement} if each student
has the same student on his or her right in both seatings: it does not matter which way they face.  We'll be
interested in counting how many arrangements there are of these 8
students, given some restrictions.

\bparts

\ppart
As a start, how many different arrangements of these 8 students
around the table are there without any restrictions?

\begin{solution}
$7!$

There are $8!$ ways to seat the 8 around the table, but all the 8
cyclic shifts of a seating define the same arrangement, giving a total
of $8!/8 = 7!$ seating arrangements.
\end{solution}

\ppart
How many arrangements of these 8 students are there with Anna
sitting next to Brian?

\begin{solution}
$2*6!$

Consider Anna and Brian to be 1 person, say AB.  Now, there are 6!
arrangements of these 7 people to seat around the table.  Now, break
up AB to be Anna and Brian---there are 2 ways to do this---Brian could
be on Anna's left or Brian could be on Anna's right, giving a total of
$2*6!$  seating arrangements.
\end{solution}

\ppart
How many arrangements are there with if Brian sitting
next to both Anna AND Caine?

\begin{solution}
$2*5!$

Consider Anna, Brian and Caine to be 1 person, say ABC, to ensure they
are seated together.  Now, you have 6 people you want to seat around a
table.  There are 5! ways of doing this.  Now, break up ABC so Brian
is in between Anna and Caine; there are 2 ways to do this: Anna is on
Brian's left and Caine on his right or vice versa.  That gives $2*5!$
seating arrangements.
\end{solution}

\ppart
How many arrangements are there with Brian sitting next to
Anna OR Caine?

\begin{solution}
$22*5!$

The inclusion-exclusion principle and the previous parts can be used
to solve this question.  Number of arrangements with Brian sitting
next to Anna OR Caine is the number of arrangements with Brian next to
Anna ($2*6!$) plus number of arrangements with Brian next to Caine
($2*6!$) minus number of arrangements with Brian next to both Anna AND
Caine ($2*5!$).  That gives us $2*6! + 2*6! - 2*5! = 4*6! - 2*5!  =
(4*6)5! - 2*5!= 22*5!$.
\end{solution}

\eparts

\end{problem}

\endinput
