\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{TP_Counting_Poker_Hands}
    \pcomment{Converted from prob1.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 04:04:28 PM EDT}
    \pcomment{revised by drewe 3 aug 2011}
\end{pcomments}

\begin{problem}

%% type: short-answer
%% title: Counting Poker Hands

Indicate how many 5-card hands there are of each of the following
kinds.  %We use notation $C(n,k)$ for $\binom{n}{k}$, and $P(n,k)$ for $\frac{n!}{(n-k)!}$.  

\bparts

\ppart 
A \textbf{Sequence} is a hand consisting of five consecutive cards of
any suit, such as
\begin{equation*}
    5\heartsuit -6\heartsuit -7\spadesuit -8\blacklozenge -9\clubsuit .
\end{equation*}
Note that an Ace may either be high (as in 10-J-Q-K-A), or low (as in
A-2-3-4-5), but can't go ``around the corner'' (that is, Q-K-A-2-3 is
\emph{not} a sequence).

How many different \textbf{Sequence} hands are possible?

\begin{solution}
$10*4^5=10240$

There are 10 possible choices for the lowest card, which determines
uniquely the sequence of ranks, and 4 choices for the suit of $each$
card.  So we have $10*4^{5}$ possible \textbf{Sequence}
hands.
\end{solution}

\ppart
A \textbf{Matching Suit} is a hand consisting of cards that are all of
the same suit in any order.

How many different \textbf{Matching
  Suit} hands are possible?

\begin{solution}
$4*C(13,5)$

There are 4 possible choices for the suit, and $C(13,5)$ for the set
of 5 ranks within the 13 in the suit, so that we have $4*C(13,5) =
5148$ possible \textbf{Matching Suit} hands.
\end{solution}

\ppart 
A \textbf{Straight Flush} is a hand that is both a \emph{Sequence} and
a \emph{Matching Suit}.

How many different \textbf{Straight Flush} hands are possible?

\begin{solution}
$4*10$

There are 4 possible choices for the suit, and 10 for the lowest card
which determines uniquely the sequence of ranks, so that we have $4*10
= 40$ possible \textbf{Straight Flush} hands.
\end{solution}

\ppart

A \textbf{Straight} is a hand that is a \emph{Sequence} but not a
\emph{Matching Suit}.

How many possible \textbf{Straights} are there?

%%  Enter your answer as a single integer without commas, such as `10001' 

\begin{solution}

10200


This is \emph{exactly} the number of \emph{Sequence} hands that are
not \emph{Straight Flush} hands, so that we have $10240 - 40 = 10200$
possible \textbf{Straights}.
\end{solution}

\ppart
A \textbf{Flush} is a hand that is a \emph{Matching Suit} but not a
\emph{Sequence}.

How many possible \textbf{Flushes} are there?
         

%% Enter your answer as a single integer without commas, such as
%% `10001'

\begin{solution}

5108


This is \emph{exactly} the number of \emph{Matching Suit} hands that
are not \emph{Straight Flush} hands, so that we have $5148 - 40 =
5108$ possible \textbf{Flushes}.
\end{solution}

\eparts

\end{problem}

\endinput
