\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{TP_Expection_of_xsquared}
    \pcomment{Converted from uniform-mean-var.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 05:11:14 PM EDT}
    \pcomment{edited by drewe 11 July 2011}
\end{pcomments}

\begin{problem}

%% type: short-answer
%% title: Expection of <em>X</em><sup>2</sup>

Let $X$ be a random variable with uniform distribution over the
integers from $-n$ to $n$.  Let $Y ::= X^{2}$. Determine whether each of the following statements is true or false.  Justify your answers briefly.

\begin{enumerate}
 
\item $\expect{X} = 0$
 
\item $\expect{Y} = 0$
 
\item $\expect{Y} > \expect{X}$
 
\item $\expect{X+Y} = \expect{X} + \expect{Y}$

\item $\expect{XY} = \expect{X} \expect{Y}$
 
\item $X$, $Y$ are independent variables.

\item $\variance{X} = 0$
 
\item $\variance{Y} = 0$

\end{enumerate}

% Give you answer as a sequence of numbers separated by spaces (e.g.,
% ``6 2 4 ``).

\begin{solution}

1. True. Since the \pdf\ of $X$ is symmetric around 0, we know the mean has
to be 0.

2. False. All values of $Y$ are nonnegative and some of them are actually
positive with non-zero probability. So, there is no way for the mean
to be 0. It has to be some positive value.

3. True. Obvious from our answers to 1 and 2. 

4. True.  By linearity of expectation, no matter what $X$ and $Y$  
we are actually talking about. 

5. True.  This is tricky: the equation does not hold in general for
non-independent $X$ and $Y$. However, in this particular case, it
happens to hold.  To see this, note that the right hand side is 0, by
our answer to $1$. At the same time, the variable $XY$ is simply
$X^{3}$. So, its \pdf\ is symmetric around $0$ (the same way the
\pdf\ of $X$ is).  Therefore, its mean has to be $0$, as well.

6. False. For example, without any knowledge about $X$, there
is some non-zero probability for $Y$ to be $n^{2}$.  However, if we
know that $X=0$, then the probability that $Y=n^{2}$ becomes $0$.

7.  False. Variance is zero only for for random variables that are constant.
$X$ is not constant.

8.  False. Variance is zero only for for random variables that are constant.
$Y$ is also not constant.
\end{solution}

\end{problem}

\endinput
