\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{TP_mating_ritual_invariant}
  \pcomment{from F07.miniquiz-oct12 edited by ARM 10/8/09}
\end{pcomments}

\pkeywords{
 stable_matching
 Mating_ritual
 invariant
 }


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem} Suppose that Harry is one of the boys and Alice is one of
  the girls in the \emph{Mating Ritual}.  Which of the properties below
  are preserved invariants?  Why?

\begin{itemize}

\item[a.] Alice is the only girl on Harry's list.

\item[b.] There is a girl who does not have any boys serenading her.

\item[c.] If Alice is not on Harry's list, then Alice has a suitor that
  she prefers to Harry.

%\item[d.] All the boys have the same number of girls left uncrossed in their list.

\item[d.] Alice is crossed off Harry's list and Harry prefers Alice to anyone
  he is serenading.

\item[e.]  If Alice is on Harry's list, then she prefers to Harry to
  any suitor she has.

\begin{staffnotes}
This part is obscure \& probably too tricky.  Tell students to come
back to it if there is time, and intervene with the solution after a
few minutes.
\end{staffnotes}

\end{itemize}

\begin{solution}

The 1st, 3rd, and 4th are preserved invariants.

\begin{enumerate}

\item[a.] A preserved invariant; no girl will be added to Harry's
  list.  If Alice got crossed off, there would be no one for Harry to
  marry.  So she must remain as the sole girl on his list.
  \textbf{Reminder}: A \emph{preserved invariant} need not be true all
  the time, as in this example.  It only needs to stay true once it
  first becomes true.

\item[b.] Not preserved; a girl may not have a suitor on the first day ---if,
for example, she's not at the top of any boy's list ---but every girl is
guaranteed to have one at the end, namely, her husband.

\item[c.] A preserved invariant; this is the basic invariant used to
  verify the Ritual.

\item[d.] A preserved invariant; Harry crosses off the girls in his
  order of preference, so if Alice is crossed off, Harry likes her
  better than anybody that's left.

\item[e.] Not preserved.  Suppose the preferences among two couples
  and a third boy are:

\begin{tabular}{llll}
Harry:    & Alice,   & Elvira,   & \dots \\
Billy:    & Elvira,  & Alice,    & \dots \\
Wilfred:  & Elvira,  & \dots             \\
Alice:    & Billy,   & Harry,    & \dots \\
Elvira:   & Wilfred, & Billy,    & \dots
\end{tabular}

The alleged invariant is true on the first day since Harry is Alice's
only suitor.  But Elvira rejects Billy in favor of Wilfred on the
first afternoon, so on the second day, Billy and Harry are serenading
Alice.  Since Alice prefers Billy to Harry, the alleged invariant is
no longer true, so it was not preserved.

\end{enumerate}

\end{solution}

\end{problem}

\endinput
