 \documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{TP_well_order_examples}
  \pcomment{ARM 2/9/12, revised 2/12/12, minor edit 3/28/13}
\end{pcomments}

\pkeywords{
  well_order
  minimum
  strictly_decreasing
  quotient
  remainder
  floor
}

\begin{problem}

\inhandout{A set of numbers is \emph{well ordered} when each of its nonempty
subsets has a minimum element.  The Well Ordering Principle says, of
course, that the set of nonnegative integers is well ordered, but so
are lots of other sets, for example, the set $r\naturals$ of numbers
of the form $rn$, where $r$ is a positive real number and $n \in
\naturals$.}

Indicate which of the following sets of numbers have a minimum element
and which are well ordered.  For those that are not well ordered, give
an example of a subset with no minimum element.

\bparts
    
\ppart\label{geqsqrt2} The integers $\geq - \sqrt{2}$.

\begin{solution}
These have a minimum element -1, and are well ordered by
Corollary~\bref{lower_bound_cor}.
\end{solution}

\iffalse

\ppart The integers $> \sqrt{2}$.

\begin{solution}
\dots same as part~\eqref{geqsqrt2}.
\end{solution}
\fi

\ppart The rational numbers $\geq \sqrt{2}$.

\begin{solution}
These have no minimum element and so are not well ordered.

They have no minimum element because $\sqrt{2}$ is irrational
(Theorem~\bref{thm:sqrt2irr_by_contra}).  So if $r$ is rational number
$\geq \sqrt{2}$, then in fact $r > \sqrt{2}$.  This means that $r$ is
not a minimum element $\geq \sqrt{2}$ because the open-ended real
interval $(\sqrt{2},r)$ has positive length, and therefore contains a
rational number, $q$, which is less than $r$ and greater than
$\sqrt{2}$.
\end{solution}

\ppart The set of rationals of the form $1/n$ where $n$ is a
positive integer.

\begin{solution}
No minimum element.
\end{solution}


\iffalse

\ppart The set of rationals of the form $1/n$ where $n$ is a positive
integer less than or equal to a $g \eqdef 10^{100}$ (a \emph{googol}).

\begin{solution}
This is a finite set of size a googol, and so is well ordered.  Its
minimum element $1/g$.
\end{solution}
\fi

\ppart The set $G$ of rationals of the form $m/n$ where $m,n >0$ and
$n \leq g$ where $g$ is a \emph{googol}, $10^{100}$.

\begin{staffnotes}
\hint Express the elements in $G$ in terms of a common denominator.
\end{staffnotes}

\begin{solution}
$G$ has minimum element $1/g$ and is well ordered.

To see why $G$ is well ordered, note that all the numbers in $G$ can
be expressed with a common denominator.  Let
\[
d \eqdef g! \eqdef 2\cdot 3\cdot 4 \cdots (g-1) \cdot g.
\]
Then each number in $G$ can be written in the form $k/d$ for some
\emph{unique} nonnegative integer numerator $k$.  So to find the
minimum element in any nonempty subset of $G$, express all the
elements with the common denominator $d$ and choose the one with
minimum numerator ---which exists by the WOP.
\end{solution}

\ppart The set, $\twdone$, of fractions of the form $n/(n+1)$:
\[
\frac01, \frac12, \frac23, \frac34, \dots.
\]

\begin{solution}
Zero is the minimum element.  The minimum element of a nonempty subset
of these numbers is simply the one with the minimum numerator when
expressed in the form $n/(n+1)$.
\end{solution}

\ppart\label{Weqnat2} Let $W \eqdef \naturals \union \twdone$ be the set consisting
of the nonnegative integers along with all the fractions of the form
$n/(n+1)$.  Describe a length 5 decreasing sequence of elements of
$W$ starting with 1,\dots length 50 decreasing sequence,\dots length 500.

\begin{staffnotes}
If students finish quickly, which they ofen do, ask them to prove that
the set $W$ in part~\eqref{Weqnat2} is well-ordered, that is any
nonempty subset of $W$ has a least element.  If a hint is needed,
suggest considering the two cases that the subset includes a fraction
from $\twdone$ or not.
\end{staffnotes}

\begin{solution}
$W$ has minimum element 0 and is well ordered
  (Lemma~\bref{to1_well-order}).

\iffalse
To show it is well ordered, we must show that every nonempty subset of
$W$ has a minimum element.  But if a subset includes a fraction from
$\twdone$, then, because all the elements in $\twdone$ are less than
1, the minimum element in the subset is the same as the minimum
element among those in $\twdone$, which exists since $\twdone$ is
well ordered.  Otherwise, the subset consists solely of nonnegative
integers and has a minimum element by the WOP.
\fi

$W$ has the unexpected property that some elements, for example the
integer one, can be the first element in strictly decreasing sequences
of elements of \emph{arbitrary} finite length.  In this sense, the
``height'' in $W$ of 1 above 0 is infinite, even though, since $W$ is
well ordered, every decreasing sequence in $W$ starting at 1 is
finite---since the elements in any infinite decreasing sequence would
be a subset with no minimum element.

For example, the following decreasing sequences of elements in $W$ all
start with 1:
\[\begin{array}{lr}
1, 0.\\
1, \frac12, 0.\\
1, \frac23, \frac12, 0.\\
1,  \frac34, \frac23, \frac12, 0.  & \text{(length 5)}\\
1,  \frac{48}{49}, \frac{47}{48}, \frac{46}{47}, \dots, \frac12, 0.  & \text{(length 50)}\\
1,  \frac{498}{499}, \frac{497}{498}, \frac{496}{497}, \dots, \frac12, 0.  & \text{(length 500)}
\end{array}\]
\end{solution}

\iffalse

\ppart\label{fin-below} A set, $R$, of real numbers with the property
that every element of $R$ has only finitely many elements of $R$ below
it.  All the sets $r\naturals$ for real $r>0$ have this property.

\hint You may assume that every finite set of real numbers has a
minimum element.  (If you feel this assumption requires a proof, see
Problem~\bref{TP_min_of_finite_WOP}).

\begin{staffnotes}
This is a point where students may get lost in unfamiliar reasoning
about abstract properties.  If they are stuck, give them $S_{\leq r}$
below and have them verify that $\min S_{\leq r} = \min S$.
\end{staffnotes}

\begin{solution}
To show that $R$ is well ordered, let $S$ be a nonempty subset of $R$,
and suppose $r \in S$.  We claim $S$ has a minimum element.  To see
why, let $S_{\leq r}$ be the set of elements in $S$ that are less than
or equal to $r$.  A minimum number in $S_{\leq r}$ is obviously also a
minimum number in $S$.\footnote{Some things really are too obvious to
  belabor with a proof---we're not trying to intimidate you into
  accepting something doubtful.}  But $S_{\leq r}$ is finite by
hypothesis, and therefore has a minimum element.
\end{solution}
\fi


\eparts

\end{problem}

\endinput
