\documentclass[problem]{mcs}

\begin{pcomments}
    \pcomment{Converted from matching.scm
              by scmtotex and dmj
              on Sun 13 Jun 2010 11:22:25 AM EDT}
\end{pcomments}

\begin{problem}

%% type: multi-part
%% title: Matching

%% type: short-answer
%% title: A Perfect Matching

\bparts

\ppart

Find a perfect matching the bipartite simple graph $G$ whose vertices and
edges are given by the following sets:

\begin{align*}
V & := \{a,b,c,d\} \union \{1,2,3,4\} \\
E & := \{\, \{a,1\}, \{a,3\}, \{b,2\}, \{c,3\}, \{c,4\}, \{d,1\}, \{d,2\} \,\}
\end{align*}

%%  Type in the box the edges of the matching. For each edge, you have to write the 
%%  letter in the pair first. For instance, you should write (a 1) and not (1 a).
%%  
%%  List the edges separated by spaces, for example, if you want to answer {1
%%  a} and {2 b}, type
%%  
%%  \begin{equation*}
%%  (a 1) (b 2)
%%  
%%  \end{equation*}
%%  
%%  The order in which you list the edges does not matter.

\begin{solution}

(a 3),
(b 2),
(c 4),
(d 1)

In this graph, the perfect matching happens to be unique.
\end{solution}

\ppart
%% type: short-answer
%% title: No Perfect Matching

The bipartite simple graph $G$ whose vertices and edges are given by
the folowing sets:

\begin{align*}
V & := \{a,b,c,d\} \union \{1,2,3,4\} \\
E & := \{\, \{a,3\}, \{a,4\}, \{b,1\}, \{b,2\},
       \{c,1\}, \{c,2\}, \{d,1\}, \{d,2\} \,\}
\end{align*}
does not have a perfect matching.

Which of the following properties of $G$ make a perfect matching
impossible?

\begin{enumerate}

\item
The set \{1,3,4\} has only 2 neighbors.

\item
The vertices $a$, $b$, $c$, $d$, on the ``left'' side, all have degree~2, but
none of the vertices 1, 2, 3, 4, on the ``right'' side, has degree~2.

\item
The set $\{b,c,d\}$ has only 2 neighbors.

\item
$G$ has 8 edges.

\item
The set $\{3,4\}$ has only 1 neighbor.  

\item
Vertex~1 has degree~3, but each of its neighbors only has degree~2.

\item
The set $\{a,b\}$ has 4 neighbors.

\end{enumerate}

%%  Give your answer as a sequence of numbers separated by some 
%%  spaces 
%%  
%%  (e.g.,  ``6 9 ``). Don't use commas or parentheses.


\begin{solution}
3,
5

If we let the letters be the ``girls'' and the numbers be the
``boys'', then property~3 is a bottleneck that prevents a perfect
match.  If we let the letters be the ``boys'' and the numbers be the
``girls'', then property~5. is another bottleneck in~$G$.

Property~1 also describes a bottleneck, but it's not true \texttt{:-)}
\end{solution}

\eparts

\end{problem}

\endinput
