\documentclass[handout]{mcs}

\begin{document}

\renewcommand{\reading}{
\begin{itemize}
\item Chapter~\bref{proofs_chap}{.\ \emph{What is a Proof?}}

\item  Chapter~\bref{well_ordering_chap}{.\ \emph{The Well Ordering
      Principle}} through~\bref{factor_sec}{;}
  (omit~\bref{well_ordering_sec}{.\ \emph{Well Ordered Sets}}).

  Chapter~\bref{logicform_chap}{.\ \emph{Logical Formulas}}
  through~\bref{equiv_valid_sec}, and~\bref{SAT_sec}{;}
  (optional:~\bref{prop_algebra_sec}{.})
\end{itemize}
These assigned readings do \textbf{not}
  include the Problem sections.  (Many of the problems in the text
  will appear as class or homework problems.)}

\problemset{1}

\medskip

\textbf{Reminder}:

\begin{itemize}

\item Pset performance does \textbf{not count} toward the final grade,
  but the exams will be based heavily on psets.  Devoting up to three
  hours working on the pset is important for learning the material.

  Problems should be submitted electronically by 11AM on the due date
  following the instructions on the
  \href{https://stellar.mit.edu/S/course/6/fa15/6.042}{class Stellar
    website}.  Graders will provide comments and grades (for feedback,
  not credit) on problems that are submitted on time.

\item The class has a
  \href{http://piazza.com/mit/fall2015/6042j18062j/home} {Piazza
    forum}.  With Piazza you may post questions---both administrative
  and content related---to the entire class or to just the staff.  You
  are likely to get faster response through Piazza than from direct
  email to staff.

\iffalse
Posting a question or comment to Piazza is required during both the
first and second weeks of the class; after that Piazza use is
optional.
\fi

\end{itemize}

\begin{staffnotes}
Lectures covered: Intro, Proof by Contradiction, WOP, Propositional Formulas
\end{staffnotes}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problems start here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pinput{MQ_log4_of_6_irrational}

\pinput{PS_3_exponent_inequality_wop}

\pinput{CP_valid_vs_satisfiable}

\pinput{PS_faster_adder_logic}


%\pinput{CP_AMM_root_2_proof}
%%\pinput{CP_roots_of_polynomials} used S14

%\pinput{CP_generalize_root_2_proof}                 on cp1f
%\pinput{CP_irrational_raised_to_an_irrational}      on cp1f
%\pinput{MQ_prove_by_wop_odds}
%\pinput{MQ_wop_proof_sumofodds}                     cp2m

%%\pinput{PS_6_10_15_stamps_by_WOP}   used S14

%\pinput{CP_binary_adder_logic}  %on cp2w
%\pinput{PS_faster_adder_logic.tex}

%% \pinput{MQ_truth_table_case_reasoning}  used S14

%\pinput{PS_printout_binary_strings}
%\pinput{TP_Quantifiers}

\end{document}
