\problemdata                                                             
        {Physics models} % latex-friendly label for the prob.
        {Statistical Mechanics} % The topic(s) of the problem content.  
        {?} % Source (if known)                                          
        {S01 Tutorial 11, 3} % Usage (list ones you are aware of). 
        {Tina Wang, S02} % Last revision info (author, date).   
                                                                         
                                                                         
\begin{problem}           
Statistical Mechanics studies the behavior of large systems of
particles (such as a kettle of boiling water).  One of the most basic
physical systems is a {\em gas:} a collection of particles (electrons,
atoms, photons) floating around in a large container. 

One way to describe this system is to divide the container into lots
of tiny boxes.  The ``state'' of the system says how the particles are
distributed among the tiny boxes.  The system is equally likely to be
in any one of its states (assuming they all have the same energy, but
discussing this would take us too far afield).  Many facts about the
aggregate behavior of the system arise from averaging over all its
possible states.

It turns out that different kinds of particles exhibit different kinds
of aggregate behavior.  These differences can be explained by
different assumptions about the system states.

The {\em Maxwell-Boltzmann\/} model assumes that the particles are
{\em distinguishable}, say, numbered $1,\dots$.  A state is described
by where particle $1$ is, where particle $2$ is, and so on.  In the
Maxwell-Boltzmann model, a single particle is equally likely to be in
any one of the tiny boxes.  Nature follows these rules when all the
particles are distinct, that is, different atoms.

\begin{problemparts}

\problempart Suppose that there are $r$ particles in
the container and $n$ tiny boxes.  How many possible states are there
under the Maxwell-Boltzmann model?

\begin{solution}
Number of states: $n^r$
\end{solution}

On the other hand, the {\em Bose-Einstein\/} model assumes that the
particles are {\em indistinguishable\/}.  A state is described by the
{\em number\/} of particles appearing in each tiny box.  In the
Bose-Einstein model, all the distributions of numbers of particles in
tiny boxes are equally likely.  Gases made up of photons obey this
model.
% Photons are a kind of {\em Boson}
% and so is Helium 2 (the superconducting stuff).
% [[[ What is this last sentence saying?]]]

\problempart How many possible states are there under the Bose-Einstein model?

\begin{solution}
Using the stars and bars model, we get that the number of possible
states is ${n+r-1 \choose r}$.
\end{solution}

The Maxwell-Boltzmann and Bose-Einstein models predict very different
aggregate system behavior.  For example, suppose $r/n = \lambda$ is
the average number of particles per cell.

\problempart Prove that in the Maxwell-Boltzmann model, the fraction of empty
cells is about
\[
e^{-\lambda}
\].

\begin{solution}
The probability that a given cell is empty is just $(1 -
\frac{1}{n})^r$.
Using the approximation $(1 - \frac{1}{n}) \approx e^{-\frac{1}{n}}$
(valid for large $n$), 
this yields approximately $e^{-\frac{r}{n}} = e^{-\lambda}$.
This probability should be close to the fraction of empty cells.
\end{solution}

\problempart Prove that under Bose--Einstein Statistics, the fraction of empty
cells is about
\[
\frac{1}{1+\lambda}
\]

\begin{solution}
The probability that a given cell is empty is just the ratio of the
number of states in which that cell is empty to the total number of
states.  The total number of states is ${n+r-1 \choose r}$.  The
number of states in which the given cell is empty is the same as the
number of states for a system with $n-1$ tiny boxes and $r$ particles,
namely, ${n+r-2 \choose r}$.  The ratio works out to
$\frac{n-1}{n+r-1} = \frac{1 - \frac{1}{n}}{1 +
\frac{r}{n} - \frac{1}{n}}$.
Since we assume $n$ is very large, this is approximately $\frac{1}{1
+ \frac{r}{n}}$, as needed.

When $\lambda$ is large, the first quantity is a lot less than the
second.  In other words, Maxwell-Boltzmann predicts a more spread-out
gas than you would actually find.
\end{solution}

\end{problemparts}
\end{problem}
                                 
\endinput
