\documentclass[12pt]{article}
\usepackage{light}

% add not exists = forall not

\hidesolutions
\showsolutions

\newcommand{\card}[1]{\left|#1\right|}

\newcommand{\C}{\textit{incabal}}

\begin{document}

\recitation{1}{September 7, 2012}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Team Problem:  A Mystery}

A certain cabal within the 6.042 course staff is plotting to make the
final exam \textit{ridiculously hard}.  (``Problem 1.  Prove that the
axioms of mathematics are complete and consistent.  Express your
answer in Mayan hieroglyphics.'')  The only way to stop their evil plan
is to determine exactly who is in the cabal.  The course staff
consists of nine people:
%
\[
%\set{\text{Spyros}, \text{Ling}, \text{Chieu}, \text{Nick},
%\text{Bill}, \text{Jay}, \text{Brooke}, \text{Marten}, \text{Tom}}
\set{\text{Keshav}, \text{Stephanie}, \text{Gabriel}, \text{Sharon}, \text{Giuliano}, \text{Maria}, \text{Ying}, \text{Neil}, \text{Tom}}
\]
% Spyros -> Keshav
%Ling -> Ying
% Chieu->Stephanie
%Marten->Neil
%Nick->Gabriel
%Bill->Giuliano
%Brooke->Maria
%Jay->Sharon
%
The cabal is a subset of these nine.  A membership roster has been
found and appears below, but it is deviously encrypted in logic
notation.  The predicate $\C$ indicates who is in the cabal; that is,
$\C(x)$ is true if and only if $x$ is a member.  Translate each
statement below into English and deduce who is in the cabal.

\begin{enumerate}[\upshape (i)]

\item\label{eee} $\exists x \ \exists y \ \exists z \
    (x \neq y \wedge
     x \neq z \wedge
     y \neq z \wedge
     \C(x) \wedge \C(y) \wedge \C(z))$

\solution[\vspace{0.3in}]{A direct English paraphrase would be ``There
exist people we'll call $x,y$, and $z$, who are all different, such that
$x,y$ and $z$ are each in the cabal.''  A better version would use the
fact that there's no need in this case to give names to the people.
Namely, a better paraphrase is, ``There are 3 different people in the
cabal.''  Perhaps a simpler way to say this is, ``The cabal is of size at
least 3.''}

\item\label{nYW} $\neg (\C(\text{Ying}) \wedge \C(\text{Giuliano}))$

\solution[\vspace{0.3in}]{Ying and Giuliano are not both in the cabal.
Equivalently: at least one of Ying and Giuliano is not in the cabal.}

\item\label{Fall} ($\C(\text{Maria}) \vee \C(\text{Gabriel})) \rightarrow \forall x \ \C(x)$

\solution[\vspace{0.3in}]{If either Maria or Gabriel is in the cabal, then everyone is.}

\item\label{YW} $\C(\text{Ying}) \rightarrow \C(\text{Giuliano})$

\solution[\vspace{0.3in}]{If Ying is in the cabal, then Giuliano is also.}

\item\label{CB} $\C(\text{Stephanie}) \rightarrow \C(\text{Maria})$

\solution[\vspace{0.3in}]{If Stephanie is in the cabal, then Maria is also.}

\item\label{SJnT}
$(\C(\text{Keshav}) \vee \C(\text{Sharon})) \rightarrow \neg \C(\text{Tom})$

\solution[\vspace{0.3in}]{If either of Keshav or Sharon is in the cabal,
then Tom is not.  Equivalently, if Tom \emph{is} in the cabal, then neither
Keshav nor Sharon is.}

\item\label{SWnM}
$(\C(\text{Keshav}) \vee \C(\text{Giuliano})) \rightarrow \neg \C(\text{Neil})$

\solution[\vspace{0.3in}]{If either of Keshav or Giuliano is in the cabal,
then Neil is not.  Equivalently, if Neil \emph{is} in the cabal, then
neither Keshav nor Giuliano is.  }
\end{enumerate}

\insolutions{So much for the translations.  We now argue that the only
cabal satisfying all seven propositions above is one whose members are
exactly Keshav, Giuliano, and Sharon.

We first observe that by~\eqref{nYW}, there must be someone --- either
Ying or Giuliano --- who is not in the cabal.  But if either Maria
or Gabriel were in the cabal, then by~\eqref{Fall}, everyone would be.
So we conclude by contradiction that

\begin{equation}\label{nF}
\text{Maria and Gabriel are not in the cabal.}
\end{equation}

Now consider that~\eqref{CB} implies its contrapositive: if Maria is
not in the cabal, then neither is Stephanie. Therefore, since Maria is
not in the cabal,

\begin{equation}
\text{Stephanie is not in the cabal.}
\end{equation}

Next observe that if Ying were in the cabal, then by~\eqref{YW},
Giuliano would be too, contradicting~\eqref{nYW}.  So by again
contradiction, we conclude that
\begin{equation}\label{nC}
\text{Ying is not in the cabal.}
\end{equation}

Now suppose Tom is in the cabal.  Then by~\eqref{SJnT}, Keshav and
Sharon are not. We already know Maria, Gabriel, Stephanie, and Ying are not
in the cabal, leaving only three who could be --- Tom, Neil, and
Giuliano.  But by~\eqref{eee} the cabal must have at least three
members, so it follows that the cabal must consist of exactly these
three.  This proves:
\begin{lemma}\label{TMW}
\text{If Tom is in the cabal, then Neil and Giuliano are in the cabal.}
\end{lemma}

But by~\eqref{SWnM}, if Giuliano is the cabal, then Neil is not.  That is, 
\begin{lemma}\label{WnM}
\text{Giuliano and Neil cannot both be in the cabal.}
\end{lemma}
Now from Lemma~\ref{WnM} we conclude that the conclusion of
Lemma~\ref{TMW} is false.  So by contrapositive, the hypothesis of
Lemma~\ref{TMW} must also be false, namely,
\begin{equation}\label{nT}
\text{Tom is not in the cabal.}
\end{equation}

Finally, suppose Neil is in the cabal.  Then by~\eqref{SWnM},
Keshav and Giuliano are not, and we already know Maria, Gabriel, Stephanie,
Ying, and Tom are not. So the cabal must consist of at most two
people (Neil and Sharon). This contradicts~\eqref{eee}, and we conclude
by contradiction that
\begin{equation}\label{nM}
\text{Neil is not in the cabal.}
\end{equation}
So the only remaining people who could be in the cabal are Keshav,
Giuliano, and Sharon.  Since the cabal must have at least three members,
we conclude that
\begin{lemma}
The only possible cabal consists of Keshav, Giuliano, and Sharon.
\end{lemma}

But we're not done yet: we haven't shown that a cabal consisting of
Keshav, Giuliano, and Sharon actually does satisfy all seven conditions.
So let $\mathcal{A} =\set{\text{Keshav}, \text{Giuliano},
  \text{Sharon}}$, and let's quickly check that $\mathcal{A}$
satisfies~\eqref{eee}--\eqref{SWnM}:

\begin{itemize}

\item $\card{A} = 3$, so $A$ satisfies~\eqref{eee}.
\item Ying is not in $A$, so $A$ satisfies~\eqref{nYW} and~\eqref{YW}.
\item Neither Maria nor Gabriel is in $A$, so the hypothesis
  of~\eqref{Fall} is false, which means that $A$
  satisfies~\eqref{Fall}.
\item Stephanie is not in $A$, so $A$ satisfies~\eqref{CB}.
\item Finally, Tom and Neil are not in $A$, so the conclusions of
both~\eqref{SJnT} and~\eqref{SWnM} are true, and so $A$
satisfies~\eqref{SJnT} and ~\eqref{SWnM}.

\end{itemize}

So now we have proved
\begin{proposition*}
$\set{\text{Keshav}, \text{Giuliano}, \text{Sharon}}$ is the \emph{unique} cabal
satisfying conditions~\eqref{eee}--\eqref{SWnM}.
\end{proposition*}}

\end{document}
