\documentclass[handout]{mcs}

\makeatletter

\def\ps@myheadings{%
    \let\@oddfoot\@empty\let\@evenfoot\@empty
    \def\@evenhead{\thepage\hfil\slshape\leftmark}%
    \def\@oddhead{\rlap{\slshape\rightmark}\hfil\thepage\hfil \llap{Name:\brule{2in}}}%
    \let\@mkboth\@gobbletwo
    \let\sectionmark\@gobble
    \let\subsectionmark\@gobble
    }

\makeatother

\begin{document}

\paper{exit}{}{End-of-term Survey}

This survey asks for your feedback on how well 6.042J/18.062J helped
you learn and appreciate the subject, and \iffalse .  Comparing
student self-assessments given in this survey to student grades helps
us determine how to improve the course.  We would also be grateful\fi
for any improvements you care to suggest.  We also encourage you to
submit an
\href{http://web.mit.edu/subjectevaluation/evaluate.html}{MIT subject
  evaluation} which asks a different set of questions.

We can make best use of your response if it is signed, but \textbf{you
  may submit this form anonymously}.  (It would still be helpful if
you would indicate your Table \& Session Time.)

\large{Your Name:} \brule{4in}

Circle your session time: 1PM\quad 2:30PM\qquad\  Table letter/number: \brule{0.5in}

\section*{Course Activities}

Please indicate with a digit how \textbf{helpful} the following
features of 6.042 were to you in learning the class material, where
digits from five (5) to one (1) mean
\begin{center}
\textbf{5}  \textbf{very} \qquad
\textbf{4}  \textbf{moderately}\qquad
\textbf{3} \textbf{somewhat}\qquad
\textbf{2} \textbf{barely}\qquad
\textbf{1} \textbf{not}
\end{center}

\iffalse
How helpful have the following aspects of the course been in
achieving the subject outcomes for you personally:
\fi

\begin{center}
\begin{tabular}{| l | c |}
%| c |}
%                & \textbf{understanding the big ideas}
%                & \textbf{preparing for exams}
\hline
\hspace{1in} feature &  value\\  %\hspace{0.3in}
\hline  \hline
\textbf{Before Class} \\ \hline \hline
   The Math for CS text          & \\  \hline
   online tutor problems & \\  \hline
   watching lecture videos before class    & \\  \hline
   reading lecture slides before class    & \\  \hline
%   NB annotation system  & \\  \hline
   Piazza class forum  & \\  \hline
\hline
\textbf{During Class}\\ \hline \hline
   team problem-solving in class  & \\   \hline
   TA/LA in class        & \\  \hline
%   lecture/presentations by Prof. Meyer  & \\  \hline
   Prof. Meyer during team-problem solving &\\ \hline
   reading lecture slides during class    & \\  \hline
   daily microquizzes &\\ \hline
   weekly miniquizzes                & \\  \hline
   midterm exams                   & \\  \hline
\hline
\textbf{After Class}\\ \hline \hline
   reviewing team problem solutions       & \\  \hline
   reading lecture slides after class     & \\  \hline
   watching lecture videos after class     & \\  \hline
%   staff email           &\\ \hline
   doing problem sets          & \\  \hline
   reviewing problem set solutions & \\  \hline
   collaborating on psets                 & \\  \hline
   staff outside class (office hours/email/\dots) & \\  \hline
   Piazza class forum  & \\  \hline
%   online grade reports                   & \\  \hline
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{| l | c |}
%| c |}
%                & \textbf{understanding the big ideas}
%                & \textbf{preparing for exams}
\hline
\hspace{1in} feature &  value\\  %\hspace{0.3in}
\hline  \hline
\textbf{Overall} \\ \hline \hline
learning the material offered by 6.042 & \\  \hline
appreciating the role of Math in CS  & \\  \hline
\end{tabular}
\end{center}


%\newpage
\section*{Learning Outcomes}

Please indicate with a digit how thoroughly the following outcomes
were \textbf{personally achieved} for yourself in taking 6.042, where
digits from five (5) to one (1) mean the outcome was
\begin{center}
\textbf{5}  \textbf{thoroughly} \qquad
\textbf{4}  \textbf{adequately}\qquad
\textbf{3} \textbf{somewhat}\qquad
\textbf{2} \textbf{barely}\qquad
\textbf{1} \textbf{not}
\end{center}

\iffalse

\subsection{Objectives}
On completion of 6.042, students will be able to
\begin{enumerate}
\item
\label{Basic Discrete Mathematics Concepts}
\textbf{reason mathematically about basic data types and structures} (such
as numbers, sets, graphs, and trees) used in computer algorithms and
systems; distinguish rigorous definitions and conclusions from merely
plausible ones; synthesize elementary proofs, especially proofs by
induction.\brule{0.5in}

\item
\label{Computational Processes} 
\textbf{model and analyze computational processes} using state machine
methods.\brule{0.5in}

\item \label{Discrete Probability} \textbf{apply principles of discrete
probability} to calculate probabilities and expectations of simple random
processes.\brule{0.5in}

\item 
\label{teams} 
\textbf{work in small teams} to accomplish all the objectives above.\brule{0.5in}
\end{enumerate}

\section*{Learning Outcomes}

In the indicated space next to each item, please enter a digit from five
(5) to one (1) where

\begin{center}
\begin{tabular}{rcll}
\hline
\textbf{5} & means &  ``this objective/outcome was & \textbf{thoroughly}
achieved for me personally.''\\
\textbf{4} & means &  ``\dots & \textbf{adequately} \dots''\\
\textbf{3} & means &  ``\dots & \textbf{somewhat} \dots''\\
\textbf{2} & means &  ``\dots & \textbf{barely} \dots''\\
\textbf{1} & means &  ``\dots & \textbf{not} \dots''\\
\hline
\end{tabular}
\end{center}
\fi

Upon completion of 6.042, students will be able to:
\begin{enumerate}

\item \label{basic} read and use \textbf{logical notation} in
  definitions and proofs. \hfill \brule{0.5in}

\item \label{logic} reason about logical
  concepts such as \textbf{validity and satisfiability} \hfill \brule{0.5in}

\item \label{sets} know the definitions and elementary properties of
  basic math concepts such as \textbf{sets, relations, and
    functions}. \hfill \brule{0.5in}

%(enter 5 ``thoroughly achieved, \dots, 1 ``not achieved'') \hfill \brule{0.5in}

\item \label{proofs} create elementary \textbf{mathematical
proofs} and identify fallacious \emph{reasoning} (not just fallacious
conclusions).  \hfill \brule{0.5in}

\item \label{well ordering principle} use the \textbf{well ordering
  principle} in simple proofs.  \hfill \brule{0.5in}

\item \label{induction} synthesize \textbf{induction hypotheses} and
  simple proofs using various forms of \textbf{induction}.  \hfill
  \brule{0.5in}

\item \label{invariants} apply the method of invariants
\iffalse and well-ordering\fi
to prove correctness and termination of \textbf{state
    machines} and processes.  \hfill \brule{0.5in}

\iffalse

\item \label{structural induction} find simple proofs using
  \textbf{structural induction}.  \hfill \brule{0.5in}
\fi

\item \label{arithmetic} do calculations and prove elementary
  properties for \textbf{modular arithmetic}.  Explain applications of
  modular arithmetic \iffalse in Computer Science, for example\fi in
  \textbf{cryptography}\iffalse or hashing\fi.\hfill \brule{0.5in}

\item \label{graphs} apply \textbf{graph theory} models of data
  structures to solve problems of matching, scheduling and
  connectivity.  \hfill \brule{0.5in}

\item \label{asymptotics} compare asymptotic growth rates of
  elementary functions and explain properties \textbf{asymptotic
    relations} such as O() and o().

\item \label{counting} find \textbf{counting formulas} for
  \textbf{permutations and combinations} and other simple
  combinatorial objects. \hfill \brule{0.5in}

\item\label{generating functions} use \textbf{generating functions} to solve
  linear recurrences and elementary counting problems.
  \hfill \brule{0.5in}

\item \label{probability} define and prove properties of elementary
  \textbf{discrete probability} models; calculate probabilities and
  conditional probabilities of events.  \hfill \brule{0.5in}

\item \label{random_variables} define \textbf{random variables},
  calculate \textbf{expectations}, and explain their use
  in \textbf{sampling}; explain \textbf{confidence levels} \hfill
  \brule{0.5in}

\item \label{student teams} problem-solve in a \textbf{small team} with
fellow students.  \hfill \brule{0.5in}

\item \label{writing} \textbf{write} brief, clear
  \textbf{explanations} and solutions for problems \hfill
  \brule{0.5in}
\end{enumerate}

\newpage
\section*{Further Comments}

How interested would you be in having another class in the
team-problem/recorded-lecture style of 6.042?
\begin{center}
\begin{tabular}{ccccc}
enthusiastic &  interested &  somewhat interested  &  uninterested &  unwilling
\end{tabular}
\end{center}

\noindent How interested would you be in serving as a 6.042 LA in some future
term?
\begin{center}
\begin{tabular}{ccccc}
enthusiastic &  interested &  somewhat interested  &  uninterested &  unwilling
\end{tabular}
\end{center}

\vspace{0.5in}
We would be pleased to hear any other comments or suggestions you may have
about the course:

\textbox{\hspace{7in}
\vspace{6in}}

\end{document}
