\hyperdef{while}{programs}{\chapter{Reasoning About \textbf{While} Programs}}\label{while_chap}

\def\movesto{\mathrel{\longrightarrow}}
\newcommand{\while}{\text{\textbf{while}}}
\newcommand{\docomm}{\text{\textbf{do}}}
\newcommand{\odcomm}{\text{\textbf{od}}}
\newcommand{\wloop}[2]{\while\ #1\ \docomm\ #2\ \odcomm}
%\newcommand{\wpr}{\while\text{ program}}
\newcommand{\assigned}{\mathbin{\mathtt{:=}}}
\newcommand{\condif}{\text{\textbf{if}}}
\newcommand{\condthen}{\text{\textbf{then}}}
\newcommand{\condelse}{\text{\textbf{else}}}
\newcommand{\condcomm}[3]{\condif\ #1\ \condthen\ #2\ \condelse\ #3}
\newcommand{\seqcomm}[2]{#1\mathbf{;}#2}
\newcommand{\Env}{\text{Env}}
\newcommand{\halt}{\text{\textbf{Done}}}
\newcommand{\state}[2]{\ang{#1,\, #2}}
\newcommand{\step}[4]{\state{#1}{#2} \movesto \state{#3}{#4}}
\newcommand{\haltswith}[3]{\state{#1}{#2}\movesto^*  \state{\halt}{#3}}
\newcommand{\after}[3]{#1 \: \set{#2}\: #3}

Real programs and programming languages are often huge and complicated,
making them hard to model and even harder to reason about.  Still, making
programs ``reasonable'' is a crucial aspect of software engineering.  In
this section we'll illustrate what it means to have a clean mathematical
model of a simple programming language and reasoning principles that go
with it ---If only real programming languages allowed for such simple,
accurate modeling.

\subsection{\textbf{While} Programs}

The programs we'll study are called ``\term{\while\ programs}.''  We
can define them as a recursive data type:
\begin{definition}\label{whiledef} \mbox{}

\textbf{base cases:}
\begin{itemize}

\item $x \assigned e$ is a \while\ program, called an \term{assignment
    statement}, where $x$ is a variable and $e$ is an expression.

\item \term{$\halt$} is a \while\ program.

\end{itemize}

\textbf{constructor cases:}
If $C$ and $D$ are \while\ programs, and $T$ is a test, then the following are also
\while\ programs:
\begin{itemize}

\item $\seqcomm{C}{D}$
---called the \term{sequencing} of $C$ and $D$,

\item $\condcomm{T}{C}{D}$ ---called a \term{conditional} with
  \term{test}, $T$, and \term{branches}, $C$ and $D$,

\item $\wloop{T}{C}$
---called a \term{while loop} with \term{test}, $T$, and \term{body}, $C$.

\end{itemize}
\end{definition}

For simplicity we'll stick to \while\ programs operating on integers.  So by
expressions we'll mean any of the familiar integer valued
expressions involving integer constants and operations such as addition,
multiplication, exponentiation, quotient or remainder.  As \term{tests},
we'll allow propositional formulas built from basic formulas of the form
$e \leq f$ where $e$ and $f$ are expressions.  For example, here is the
Euclidean algorithm for $\gcd(a,b)$ expressed as a \while\ program.
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\kill
$\mathtt{x} \assigned a$\textbf{;} \\
$\mathtt{y} \assigned b$\textbf{;} \\
\while\ $\mathtt{y} \neq 0$ \docomm\\
   \> $\mathtt{t} \assigned \mathtt{y}$\textbf{;}\\
   \> $\mathtt{y} \assigned \rem{\mathtt{x}}{\mathtt{y}}$\textbf{;}\\
   \> $\mathtt{x} \assigned \mathtt{t}$
\odcomm\\
\end{tabbing}

\subsection{The \textbf{While} Program State Machine}

A \while\ program acts as a pure command: it is run solely for its
side effects on stored data and it doesn't return a value.  The data
consists of integers stored as the values of variables, namely
environments:
\begin{definition}
An \term{environment} is a total function from variables to integers.
Let $\Env$ be set of all environments.
\end{definition}
So if $\rho$ is an environment and $x$ is a variable, then $\rho(x)$ is an
integer.  More generally, the environment determines the integer value of
each expression, $e$, and the truth value of each test, $T$.  We can think
of an expression, $e$ as defining a function $\mean{e}:\Env \to
\integers$, and refer to this function, $\mean{e}$ as the \term{meaning}
of $e$, and likewise for tests.

It's standard in programming language theory to write $\mean{e}\rho$
as shorthand for $\mean{e}(\rho)$, that is, applying
the \emph{meaning}, $\mean{e}$, of $e$ to $\rho$.  For example, if
$\rho(\mathtt{x}) =4$, and $\rho(\mathtt{y}) =-2$, then
\begin{equation}\label{x2y311}
\mean{\mathtt{x^2+y -3}}\rho = \rho(\mathtt{x})^2 + \rho(y) -3 = 11.
\end{equation}

\iffalse
(A variable is a special case of an expression, so $\mean{x}\rho \eqdef \rho(x)$.)
\fi

Executing a program causes a succession of changes to the
environment\footnote{More sophisticated programming models distinguish
the environment from a \term{store} which is affected by commands, but
this distinction is unnecessary for our purposes.}  which may continue
until the program halts.  Actually the only command which immediately
alters the environment is an assignment command.  Namely, the effect
of the command
\[
x \assigned e
\]
on an environment is that the value assigned to the variable $x$ is
changed to the value of $e$ in the original environment.  We can say
this precisely and concisely using the following notation:
$\patch{f}{a}{b}$ is a function that is the same as the function, $f$,
except that when applied to element $a$ its value is $b$.  Namely,
\begin{definition}
If $f:A \to B$ is a function and $a,b$ are arbitrary elements, define
  \[
\patch{f}{a}{b}
\]
to be the function $g$ such that
\begin{equation*}
g(u) = \begin{cases}
         b & \text{ if } u = a.\\
         f(u) & \text{otherwise.}
       \end{cases}
\end{equation*}
\end{definition}

Now we can specify the step-by-step execution of a \while\ program as a state
machine, where the states of the machine consist of a \while\ program paired with an
environment.  The transitions of this state machine are defined
recursively on the definition of \while\ programs.

\begin{definition}

The transitions $\step{C}{\rho}{D}{\rho'}$ of the \while\ program state machine are
defined as follows:

\textbf{base cases:}

\[
\step{x \assigned e}{\rho}{\halt}{\patch{\rho}{x}{\mean{e}\rho}}
\]

\textbf{constructor cases:}
If $C$ and $D$ are \while\ programs, and $T$ is a test, then:

\begin{itemize}

\item if $\step{C}{\rho}{C'}{\rho'}$, then
\[
\step{\seqcomm{C}{D}}{\rho}{C';D}{\rho'}.
\]
Also,
\[
\step{\seqcomm{\halt}{D}}{\rho}{D}{\rho}.
\]

\item if $\mean{T}\rho = \true$, then
\[
\step{\condcomm{T}{C}{D}}{\rho}{C}{\rho},
\]
or if $\mean{T}\rho = \false$, then
\[
\step{\condcomm{T}{C}{D}}{\rho}{D}{\rho}.
\]

\item if $\mean{T}\rho = \true$, then
\[
\step{\wloop{T}{C}}{\rho}{\seqcomm{C}{\wloop{T}{C}}}{\rho}
\]
or if $\mean{T}\rho = \false$, then
\[
\step{\wloop{T}{C}}{\rho}{\halt}{\rho}.
\]
\end{itemize}
\end{definition}

Now \while\ programs are probably going to be the simplest kind of programs you will
ever see, but being condescending about them would be a mistake.  It turns
that \emph{every function on nonnegative integers that can be computed by
  any program} on any machine whatsoever can also be computed by \while\ programs
(maybe more slowly).  We can't take the time to explain how such a
sweeping claim can be justified, but you can find out by taking a course
in Computability Theory. % such as 6.045 or 6.840.

\subsection{Denotational Semantics}

The net effect of starting a \while\ program in some environment is reflected in the
final environment when the program halts.  So we can think of a \while\ program, $C$,
aas defining a function, $\mean{C}:\Env \to \Env$, from initial
environments to environments at halting.  The function $\mean{C}$ is
called the \term{meaning} of $C$.

$\mean{C}$ of a \while\ program, $C$ to be a partial function from $\Env$ to $\Env$
mapping an initial environment to the final halting environment.

We'll need one bit of notation first.  For any function $f:S \to S$, let
$f^{(n)}$ be the composition of $f$ with itself $n$ times where $n \in
\naturals$.   Namely,
\begin{align*}
f^{(0)} & \eqdef \ident{S}\\
f^{(n+1)} & \eqdef f \compose f^{(n)},
\end{align*}
where ``$\compose$'' denotes functional composition.

The recursive definition of the meaning of a program follows the
definition of the \while\ program recursive data type.
\begin{definition}

\textbf{base cases:}
\begin{itemize}

\item $\mean{x \assigned e}$ is the function from $\Env$ to $\Env$ defined
  by the rule:
\[
\mean{x \assigned e}\rho \eqdef \patch{\rho}{x}{\mean{e}\rho}.
\]

\item
\[
\mean{\halt} \eqdef \ident{\Env}
\]
where $\ident{\Env}$ is the identity function on $\Env$.  In other words,
$\mean{\halt}\rho \eqdef \rho$.

\end{itemize}

\textbf{constructor cases:}
If $C$ and $D$ are \while\ programs, and $T$ is a test, then:
\begin{itemize}

\item
\[
\mean{\seqcomm{C}{D}} \eqdef \mean{D} \compose \mean{C}
\]
That is,
\[
\mean{\seqcomm{C}{D}}\rho \eqdef \mean{D}(\mean{C}\rho).
\]

\item
\[
\mean{\condcomm{T}{C}{D}}\rho
\eqdef
\begin{cases}
\mean{C}\rho & \text{if } \mean{T}\rho = \true\\
\mean{D}\rho & \text{if } \mean{T}\rho = \false.
\end{cases}
\]

\item
\[
\mean{\wloop{T}{C}}\rho \eqdef \mean{C}^{(n)}\rho
\]
where $n$ is the least nonnegative integer such that
$\mean{T}(\mean{C}^{(n)}\rho) = \false$.  (If there is no such $n$, then
$\mean{\wloop{T}{C}}\rho$ is undefined.)
\end{itemize}

\end{definition}

We can use the denotational semantics of \while\ programs to reason about
\while\ programs using structural induction on programs, and this is often
much simpler than reasoning about them using induction on the number of
steps in an execution.  This is OK as long as the denotational semantics
accurately captures the state machine behavior.  In particular, using the
notation $\movesto^*$ for the transitive closure of the transition
relation:
\begin{theorem}\label{sound-semantics}
\[
\haltswith{C}{\rho}{\rho'} \qiff \mean{C}\rho = \rho'
\]
\end{theorem}
Theorem~\ref{sound-semantics} can be proved easily by induction; it
appear in~Problem\ref{PS_while_program_semantic_soundness_proof}.

\begin{problems}
\homeworkproblems
\pinput{PS_while_program_semantic_soundness_proof}
\end{problems}

\subsection{Logic of Programs}

A typical program specification describes the kind of inputs and
environments the program should handle, and then describes what should
result from an execution.  The specification of the inputs or initial
environment is called the \term{precondition} for program execution, and
the prescription of what the result of execution should be is called the
\emph{postcondition}.  So if $P$ is a logical formula expressing the precondition
for a program, $C$, and likewise $Q$ expresses the postcondition, the
specification requires that
\begin{quote}
If $P$ holds when $C$ is started, then $Q$ will hold if and when $C$ halts.
\end{quote}
We'll express this requirement as a formula
\[
\after{P}{C}{Q}
\]
called a \term{partial correctness assertion}.

For example, if $E$ is \while\ program above for the Euclidean
algorithm, then the partial correctness of $E$ can be expressed as
\begin{equation}\label{eucpc}
\after{\paren{a, b \in \naturals\ \QAND\  \mathtt{x} \neq 0}}{C}
                   {\paren{\mathtt{x} = \gcd(a,b)}}.
\end{equation}

More precisely, notice that just as the value of an expression in an
environment is an integer, the value of a logical formula in an
environment is a truth value.  For example, if $\rho(\mathtt{x}) =4$,
and $\rho(\mathtt{y}) =-2$, then by~\eqref{x2y311},
$\mean{\mathtt{x^2+y -3}}\rho = 11$, so
\begin{align*}
\mean{\exists \mathtt{z}.\, \mathtt{z} > 4 \QAND \mathtt{x}^2+\mathtt{y} -3=\mathtt{z}}\rho & = \true,\\
\mean{\exists \mathtt{z}.\, \mathtt{z} > 13 \QAND \mathtt{x}^2+\mathtt{y} -3=\mathtt{z}}\rho & = \false.
\end{align*}

\begin{definition}\label{def_afterPCQ}
For logical formulas $P$ and $Q$, and \while\ program, $C$, the
partial correctness assertion
\[
\after{P}{C}{Q}
\]
is true proving that for all environments, $rho$, if $\mean{P}\rho$ is
true, and $\haltswith{C}{\rho}{\rho'}$ for some $\rho'$, then
$\mean{Q}\rho'$ is true.
\end{definition}


In the 1970's, 
\begin{editingnotes}
check ref
\end{editingnotes}
Prof.\ Tony Hoare (now Sir Anthony) at Univ. Dublin formulated a set
of inference rules for proving partial correctness formulas.  These
rules are known as \term{Hoare Logic}.

The first rule captures the fact that strengthening the preconditions
and weakening the postconditions makes a partial correctness
specification easier to satisfy:

\Rule{P \ \QIMPLIES\  R,
     \quad \after{R}{C}{S},
     \quad S \ \QIMPLIES\  Q}{\after{P}{C}{Q}}

The rest of the logical rules follow the recursive definition of \while\ programs.
There are axioms for the base case commands:
\[\begin{array}{c}
  \after{P(x)}{x \assigned e}{P(e)}\\
  \after{P}{\halt}{P},
\end{array}\]
and proof rules for the constructor cases:

\begin{itemize}

\item
\Rule{\after{P}{C}{Q} \ \QAND\  \after{Q}{D}{R}}{\after{P}{\seqcomm{C}{D}}{R}}

\item
\Rule{\after{P\ \QAND\ T}{C}{Q}}
        {\after{P\ \QAND\ T}{\condcomm{T}{C}{D}}{Q\ \QAND\ T}}

\item
\Rule{\after{P\ \QAND\ T}{C}{P}}
     {\after{P}{\wloop{T}{C}}{P\ \QAND\ \QNOT(T)}}
\end{itemize}

\begin{example}
\textbf{Proof of partial correctness~\eqref{eucpc} for the Euclidean algorithm.}
\end{example}

\TBA{Brief discussion of ``relative completeness''.}

\endinput
