\documentclass[handout]{mcs}

\begin{document}

\inclassproblems{13, Mon.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problems start here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pinput{CP_a_baseball_series}
\pinput{CP_coin_flips}
\pinput{CP_system_component_failure}
\pinput{CP_proving_probability_rules}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problems end here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \instatements{\newpage}

\section{Appendix}

\subsection{The Four Step Method}

This is a good approach to questions of the form, ``What is the
probability that -------?''  Intuition can be misleading, but
this formal approach gives the right answer every time.

\begin{enumerate}
\item Find the sample space.  (Use a tree diagram.)

\item Define events of interest.  (Mark leaves corresponding to these
events.)

\item Determine outcome probabilities:

\begin{enumerate}

\item Assign edge probabilities.

\item Compute outcome probabilities.  (Multiply along root-to-leaf
paths.)

\end{enumerate}

\item Compute event probabilities.  (Sum the probabilities of all
outcomes in the event.)

\end{enumerate}

\subsection{Probability Spaces}

  A countable \term{sample space}, $\sspace$, is a nonempty countable set.
  An element $w \in \sspace$ is called an \term{outcome}.  A subset of
  $\sspace$ is called an \term{event}.

  A \term{probability space} consists of a sample
  space, $\sspace$, and a function $\pr{}: \sspace\to [0,1]$, called
  the \term{probability function}, such that
\[
\sum_{w \in \sspace} \pr{w} = 1.
\]

For any event, $E \subseteq \sspace$, the \term{probability of $E$} is
defined to be the sum of the probabilities of the outcomes in $E$:
\[
\pr{E} \eqdef \sum_{w \in E} \pr{w}.
\]

\subsection{Sum Rule \& Union Bound}

Let $E_0,E_1,\dots$ be a (possibly infinite) sequence of events.  These
events are said to be \emph{pairwise disjoint} if $E_i \intersect E_j =
\emptyset$ whenever $i \neq j$.

If these events are pairwise disjoint, then
\begin{equation}
\pr{\lgunion_{n \geq 0} E_n} = \sum_{n \geq 0} \pr{E_n}.\tag{Disjoint Sum Rule}
\end{equation}

Even if they are not pairwise disjoint,
\begin{equation}
\pr{\lgunion_{n \geq 0} E_n} \leq \sum_{i \geq n} \pr{E_n}.\tag{Union Bound}
\end{equation}

\end{document}
